---
title: "EVT-cohesive Type Clustering"
subtitle: "z_type clusters that map each EVT to a single cluster (one-to-many EVT->cluster)"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  evt_table: "~/Dropbox/git/vq-vae/data/LF2024_EVT.csv"
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  K: 20            # number of EVT-cohesive type clusters
  min_evt_n: 20    # min plots for an EVT to be a clustering atom; smaller -> nearest cluster
  phase_subdivide_frac: 0.25  # fraction of type clusters (highest z_phase spread) to subdivide
  k_phase: 3       # phase sub-groups per subdivided cluster
  K_strata: 5      # final strata after supervised aggregation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
```

## Overview

Cluster the **type** embedding `z_type` so the clustering is (as much as possible)
a **one-to-many** map from EVT to cluster: several EVTs may share a cluster, but a
single EVT should not be split across clusters.

The clean way to guarantee that is to cluster at the **EVT level**:

1. Represent each EVT by its **centroid** in standardized `z_type` space.
2. **Ward-cluster the EVT centroids** into `K` groups (reliable EVTs only, i.e.
   `n >= min_evt_n`; rare EVTs are assigned to the nearest cluster afterward).
3. Assign every plot to **its EVT's** cluster.

Because clusters are unions of whole EVTs, every EVT maps to exactly one cluster
**by construction** (clustering *completeness* = 1). EVT is an external vegetation
label, independent of the FIA response variables, so this is supervision that
cannot leak into BALIVE/removals. The interesting question it answers: **does
`z_type` naturally respect EVT boundaries?** — measured by how much the EVT
constraint costs vs a free clustering.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Load and per-EVT centroids

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(!is.na(evt), evt != -9999)                     # drop nodata EVT

Z    <- scale(as.matrix(emb[, emb_type_vars]))          # standardized z_type
evt  <- as.character(emb$evt)
evt_n <- table(evt)

# per-EVT centroid in standardized z_type
cent  <- rowsum(Z, evt) / as.numeric(evt_n[rownames(rowsum(Z, evt))])
big   <- names(evt_n)[evt_n >= params$min_evt_n]        # reliable centroids
cat(sprintf("plots: %d   EVTs: %d (%d with n >= %d used as clustering atoms)\n",
            nrow(emb), length(evt_n), length(big), params$min_evt_n))
```

## Ward-cluster the EVT centroids

```{r cluster, fig.width=11, fig.height=6}
cent_big <- cent[big, , drop = FALSE]
hc <- hclust(dist(cent_big), method = "ward.D2")
K  <- min(params$K, nrow(cent_big))
cl_big <- cutree(hc, k = K)                              # named by big EVT

# cluster centroids (rows ordered 1..K)
clcent <- t(sapply(sort(unique(cl_big)),
                   function(k) colMeans(cent_big[cl_big == k, , drop = FALSE])))

# assign every EVT: big keep their cut; rare -> nearest cluster centroid
nearest <- function(v) which.min(colSums((t(clcent) - v)^2))
evt_cluster <- vapply(rownames(cent), function(e)
  if (e %in% big) cl_big[[e]] else nearest(cent[e, ]), integer(1))
names(evt_cluster) <- rownames(cent)

emb$type_cluster <- factor(evt_cluster[evt])

# dendrogram of the big EVTs, boxed by cluster
op <- par(mar = c(6, 4, 3, 1))
plot(hc, labels = big, xlab = "EVT", sub = "", cex = 0.6,
     main = sprintf("EVT centroids in z_type -> %d clusters (Ward)", K))
rect.hclust(hc, k = K, border = 2:(K + 1))
par(op)
```

## Verify the one-to-many property

```{r verify}
tab <- table(evt = evt, cluster = emb$type_cluster)
per_evt_clusters <- rowSums(tab > 0)                    # clusters each EVT appears in
cat(sprintf("EVTs mapping to exactly one cluster: %d / %d (%.1f%%)\n",
            sum(per_evt_clusters == 1), length(per_evt_clusters),
            100 * mean(per_evt_clusters == 1)))
stopifnot(all(per_evt_clusters == 1))                   # true by construction
```

## Cluster composition

Each cluster aggregates several EVTs (expected). `n_evts` = distinct EVTs pooled
into the cluster; `top_evts` = the largest few by plot count.

```{r composition}
emb %>%
  group_by(type_cluster) %>%
  summarize(n_plots = n(),
            n_evts  = dplyr::n_distinct(evt),
            top_evts = paste(names(sort(table(evt), decreasing = TRUE))[1:3], collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_plots))
```

## Does z_type respect EVT? (cost of the constraint)

Compare within-cluster `z_type` dispersion of the EVT-cohesive clustering to a
**free** k-means at the same `K` (which minimizes it). Ratio near 1 means the
embedding already separates EVTs cleanly, so forcing EVT-cohesion costs almost
nothing; a large ratio means the constraint fights the embedding.

```{r compactness}
tot_within <- function(cl) {
  ss <- sum(vapply(split(seq_len(nrow(Z)), cl), function(ix) {
    m <- Z[ix, , drop = FALSE]; sum(sweep(m, 2, colMeans(m))^2)
  }, numeric(1)))
  ss / nrow(Z)
}
set.seed(1)
km <- kmeans(Z, centers = K, nstart = 10)

w_evt <- tot_within(emb$type_cluster)
w_km  <- tot_within(km$cluster)
data.frame(
  clustering = c("EVT-cohesive (this)", "free k-means (same K)"),
  within_var = round(c(w_evt, w_km), 3),
  ratio_to_kmeans = round(c(w_evt / w_km, 1), 3)
)

# homogeneity: how EVT-mixed are the clusters (low is expected -- many EVTs/cluster)
homogeneity <- 1 - sum(apply(tab, 2, function(col) {
  p <- col[col > 0] / sum(col); -sum(p * log(p))
}) * colSums(tab)) / sum(tab) /
  ( -sum((rowSums(tab)/sum(tab)) * log(rowSums(tab)/sum(tab))) )
cat(sprintf("completeness (EVT -> one cluster): 1.000 (by construction)\n"))
cat(sprintf("homogeneity  (cluster -> one EVT): %.3f (higher = clusters are EVT-coherent)\n",
            homogeneity))
```

## EVT -> cluster map

The full assignment (join `evt` -> `type_cluster` onto any plot table to use these
as type strata).

```{r map}
# LANDFIRE EVT lookup: VALUE (the EVT code) -> EVT_NAME
evt_lookup <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  select(VALUE, EVT_NAME)

evt_map <- tibble(VALUE = as.integer(names(evt_cluster)),
                  n_plots = as.integer(evt_n[names(evt_cluster)]),
                  type_cluster = as.integer(evt_cluster),
                  atom = names(evt_cluster) %in% big) %>%
  left_join(evt_lookup, by = "VALUE") %>%
  select(VALUE, EVT_NAME, n_plots, type_cluster, atom) %>%
  arrange(type_cluster, desc(n_plots))
evt_map
```

## Stratification: BALIVE, Volume, and Removals vs TCC

Build the fine bins as **EVT type clusters, then phase-subdivided**: take the
`r params$K` EVT-cohesive clusters, and split the top
`r round(100*params$phase_subdivide_frac)`% (by within-cluster `z_phase` spread)
into `r params$k_phase` phase sub-groups each (Ward on `z_phase`) — the same
type->phase refinement as the geometry hierarchical script, but with EVT-based
type bins. Then aggregate the fine bins into `K_strata` strata by Ward on the
cluster-mean response (supervised), and compare within-stratum SD against TCC and
a flat type+phase probe, by survey unit. Responses: **BALIVE**, **standing live
volume**, and **removals**. Stocks (BALIVE/volume) run on all forested plots;
removals additionally require remeasurement (with forested zero-removal plots
kept as 0).

```{r strat-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

cond <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, CONDPROP_UNADJ, BALIVE, FORTYPCD FROM COND WHERE STATECD='%s'",
  params$statecd)) %>% mutate(PLT_CN = as.character(PLT_CN))
balive <- cond %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")
fortyp <- cond %>%
  group_by(PLT_CN) %>% slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup() %>% select(PLT_CN, FORTYPCD)

dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# standing LIVE net cu ft volume per acre (STATUSCD=1)
vol <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
     FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

# annual per-acre removed GS volume; COALESCE inside SUM so no-removal plots are 0
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN))
```

```{r strat-tcc}
locs <- fia_locs %>% filter(CN %in% emb$CN) %>% distinct(CN, .keep_all = TRUE)
tccr <- rast(params$tcc)
pts  <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
tv   <- extract(tccr, project(pts, crs(tccr)), ID = FALSE)[[1]]
tv   <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
tcc_by_cn <- tibble(CN = locs$CN, tcc = tv)
```

```{r strat-assemble}
analysis <- emb %>%
  select(CN, type_cluster, all_of(emb_type_vars), all_of(emb_phase_vars)) %>%
  inner_join(balive, by = c("CN" = "PLT_CN")) %>%
  left_join(vol,     by = c("CN" = "PLT_CN")) %>%
  inner_join(fortyp, by = c("CN" = "PLT_CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD), by = "CN") %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  mutate(volume = coalesce(volume, 0)) %>%
  filter(!is.na(FORTYPCD), !is.na(UNITCD), !is.na(tcc), tcc >= 0, tcc <= 100)
cat(sprintf("forested plots with covariates: %d across %d units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
```

```{r strat-helpers}
pooled_within <- function(df, groups, y) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
dh_strata <- function(x, K, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(factor(rep(1L, length(x))))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  bnd <- sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
  factor(findInterval(x, c(-Inf, bnd, Inf)))
}
ward <- function(X, k) {
  d  <- dist(X)
  hc <- if (requireNamespace("fastcluster", quietly = TRUE))
          fastcluster::hclust(d, method = "ward.D2") else hclust(d, method = "ward.D2")
  cutree(hc, k = k)
}
# aggregate fine bins -> K strata by Ward on the fine-bin-mean response (supervised)
agg_by <- function(bin, y, K) {
  d  <- tibble(b = bin, my = y) %>% group_by(b) %>% summarize(my = mean(my), .groups = "drop")
  Kk <- min(K, nrow(d))
  d$s <- cutree(hclust(dist(scale(d$my)), method = "ward.D2"), Kk)
  factor(d$s[match(bin, d$b)])
}
# flat OOF supervised probe on `vars` -> DH bins (the direct-supervision ceiling)
oof_dh <- function(df, y, vars, K) {
  set.seed(1); folds <- sample(rep(1:5, length.out = nrow(df)))
  d <- df %>% mutate(y_resp = .data[[y]])
  form <- reformulate(vars, "y_resp"); s <- numeric(nrow(df))
  for (f in 1:5) { fit <- lm(form, d[folds != f, ]); s[folds == f] <- predict(fit, d[folds == f, ]) }
  dh_strata(s, K)
}
```

### Fine bins: EVT type clusters -> phase subdivision

```{r fine-bins}
# within-cluster z_phase spread; subdivide the top fraction into k_phase groups
Zp   <- scale(as.matrix(analysis[, emb_phase_vars]))
pvar <- tapply(seq_len(nrow(analysis)), analysis$type_cluster,
               function(ix) sum(apply(Zp[ix, , drop = FALSE], 2, var)))
n_cl   <- length(pvar)
n_sub  <- max(1, round(params$phase_subdivide_frac * n_cl))
sub_cl <- names(sort(pvar, decreasing = TRUE))[seq_len(n_sub)]

analysis$phase_sub <- 1L
for (b in sub_cl) {
  ix <- which(as.character(analysis$type_cluster) == b)
  if (length(ix) >= 10 * params$k_phase)
    analysis$phase_sub[ix] <- ward(Zp[ix, , drop = FALSE], params$k_phase)
}
analysis$fine_bin <- interaction(analysis$type_cluster, analysis$phase_sub, drop = TRUE)
cat(sprintf("fine bins: %d (from %d type clusters, %d phase-subdivided)\n",
            nlevels(analysis$fine_bin), n_cl, length(sub_cl)))
```

### Removals population (forested + remeasured)

```{r removals-pop}
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # remeasured; zeros kept
cat(sprintf("forested remeasured plots: %d (%.1f%% with any removal)\n",
            nrow(ar), 100 * mean(ar$remv_cfgs > 0)))
```

### Evaluation

```{r strat-eval}
resp_table <- function(df, y, vars = emb_vars) {
  s_tcc  <- dh_strata(df$tcc, params$K_strata)
  s_evt  <- agg_by(df$fine_bin, df[[y]], params$K_strata)
  s_flat <- oof_dh(df, y, vars, params$K_strata)
  vv <- c(pooled_within(df, "UNITCD", y),
          pooled_within(mutate(df, .s = s_tcc),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_evt),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_flat), c("UNITCD", ".s"), y))
  data.frame(
    stratification  = c("UNITCD only (baseline)",
                        sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                        sprintf("UNITCD x EVT type/phase hierarchical (%d)", params$K_strata),
                        sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var      = round(vv, 2),
    within_sd       = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1))
}
```

#### BALIVE

```{r strat-balive}
resp_table(analysis, "BALIVE")
```

#### Standing live volume

```{r strat-volume}
resp_table(analysis, "volume")
```

#### Removals

```{r strat-removals}
resp_table(ar, "remv_cfgs")
```

```{r strat-cleanup}
dbDisconnect(con)
```

## Notes

- **Guaranteed one-to-many.** Clustering at the EVT level makes completeness = 1
  exactly; `homogeneity` is informational (clusters intentionally pool EVTs).
- **The compactness ratio is the headline diagnostic.** Near 1.0 = `z_type`
  already encodes EVT structure (the model learned vegetation type), so
  EVT-cohesive strata are also embedding-compact. Far above 1 = EVTs are smeared
  across the embedding and the constraint is expensive.
- **Rare EVTs** (`n < min_evt_n`) are assigned to the nearest cluster centroid, so
  they still map one-to-one to a cluster; raise `min_evt_n` to treat more of them
  as noise-assigned rather than clustering atoms.
- **Choosing K.** Read it off the dendrogram; `K` is a param. Unweighted centroids
  give rare and common EVTs equal say in the grouping -- weight `cent_big` rows by
  `sqrt(n)` if you want common EVTs to dominate the merges.
- **Using it downstream.** `emb$type_cluster` (or `evt_map`) is a drop-in
  replacement for the k-means/Ward `type_bin` step in the hierarchical
  stratification -- an EVT-supervised type binning that never sees the FIA target.
- **Fine bins = EVT type + phase.** The EVT clusters are phase-subdivided (top
  `phase_subdivide_frac` by `z_phase` spread, `k_phase` groups each via Ward) to
  match the geometry hierarchical script; set `phase_subdivide_frac = 0` to test
  EVT-type-only fine bins.
- **Responses.** BALIVE and standing live volume (`TREE.VOLCFNET`, `STATUSCD=1`)
  are stocks on all forested plots; removals (`TREE_GRM_COMPONENT`, annual per-acre
  removed GS volume) additionally require remeasurement, with forested no-removal
  plots kept as 0. The `EVT type/phase hierarchical` row aggregates the fine bins
  to `K_strata` by Ward on the fine-bin-mean response; `flat embedding supervised`
  is the direct OOF (type+phase) probe ceiling for the same information.
