---
title: "FIA Post-Stratification Efficiency: TCC vs Embedding Strata"
subtitle: "Within-stratum BALIVE variance by survey unit, Virginia FIA plots"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  reticulate_python: "/home/nnagle/miniconda3/envs/r-tf/bin/python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi = 110
)
```

## Overview

This report asks a **post-stratification efficiency** question: to estimate
`BALIVE` (live basal area) within each FIA survey unit (`UNITCD`), is it better
to post-stratify plots by **tree canopy cover (TCC)** or by **learned embedding
space**?

The relevant quantity is the **within-stratum variance of `BALIVE`**. Under an
approximately equal-probability sample (the FIA base grid, so stratum weight
`W_h ≈ n_h / n`) with proportional allocation, the variance of the
post-stratified mean is approximately

$$\operatorname{Var}(\bar y_{\text{strat}}) \;\approx\; \frac{1}{n}\sum_h W_h\,S_h^2
  \;=\; \frac{S^2_{\text{within}}}{n},$$

so the **pooled within-stratum variance** `S²_within` is exactly the number to
minimize. Both stratifications get **4 strata** and are crossed with `UNITCD`, so
the comparison is like-for-like.

- **TCC strata:** canopy-cover bins `0–10, 11–48, 49–85, 86–100`.
- **Embedding strata:** two variants — unsupervised k-means on `z_type`, and a
  supervised (out-of-fold) BALIVE read-out of the embedding binned into 4. The
  supervised one is the fair analog to TCC (a single BALIVE-correlated covariate
  binned into 4); k-means tests whether raw embedding *geometry* stratifies BALIVE.

The baseline is `UNITCD`-only (no sub-stratification); TCC and embedding are
judged by how far below that baseline they push the residual `BALIVE` variance.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## FIA population tables (reference)

Pulled for context / later design-based weighting; not used in the variance
comparison below.

```{r pop-tables}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

stratum_data <- dbGetQuery(
  con,
  sprintf("SELECT * FROM POP_STRATUM WHERE STATECD='%s' AND EVALID=512201", params$statecd)
)
eval_data <- dbGetQuery(
  con,
  sprintf("SELECT * FROM POP_EVAL WHERE STATECD='%s'", params$statecd)
)
```

## Data assembly

### Plot locations

`CN`/`ID` are FIA control numbers that exceed IEEE-754 double precision, and the
embedding CSV / `COND` table carry them as **text**. They must be read as
character here too, or the downstream joins silently drop or mismatch rows.

```{r fia-locs}
fia_locs <- read_csv(
  params$fia_locs,
  col_types = cols(CN = "c", ID = "c")   # keep control numbers exact
) %>%
  mutate(CN = sub("\\.0$", "", CN),
         ID = sub("\\.0$", "", ID))
```

### Extract TCC at each plot

Points are built in NAD83 (FIA datum), reprojected into the raster's AEA CRS, and
the extracted value column is renamed to `tcc`.

```{r tcc-extract}
tcc <- rast(params$tcc)

pts     <- vect(fia_locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
pts_aea <- project(pts, crs(tcc))

vals <- extract(tcc, pts_aea, ID = FALSE)   # nearest cell
names(vals) <- "tcc"                        # drop the long raster layer name

# The science-TCC raster carries a categorical attribute table, so `extract`
# returns a FACTOR with levels "0".."100","254","255" -- not numbers. Recover the
# numeric percent via as.character() first (as.numeric() on a factor would give
# level indices, not values), then blank out the nodata codes: 254 = non-
# processing area, 255 = fill, and anything > 100 is not a canopy-cover percent.
vals$tcc <- as.numeric(as.character(vals$tcc))
vals$tcc[vals$tcc > 100] <- NA_real_

fia_merged <- bind_cols(fia_locs, vals)
```

### BALIVE per plot (from COND)

Area-weighted over conditions (`CONDPROP_UNADJ`), with nonforest `BALIVE` (`NA`)
treated as 0 — the same convention as the embedding-evaluation report.

```{r balive}
cond_data <- dbGetQuery(
  con,
  sprintf("SELECT PLT_CN, CONDPROP_UNADJ, BALIVE FROM COND WHERE STATECD='%s'",
          params$statecd)
)

balive_plot <- cond_data %>%
  mutate(PLT_CN = as.character(PLT_CN),
         BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ),
            .groups = "drop")
```

### Embeddings (z_type block)

```{r embeddings}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)   # full embedding for the supervised score

emb_data <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  distinct(CN, .keep_all = TRUE)             # one embedding row per plot
```

### One analysis frame

Plot → `UNITCD`, `tcc`, `BALIVE`, `z_type`. Restricted to complete cases with a
valid canopy-cover value (0–100; the science-TCC raster uses >100 for nodata) so
that the **TCC and embedding stratifications are computed on the identical set of
plots**.

```{r assemble}
analysis <- fia_merged %>%
  select(CN, UNITCD, tcc) %>%
  inner_join(balive_plot, by = c("CN" = "PLT_CN")) %>%
  inner_join(select(emb_data, CN, all_of(emb_vars)), by = "CN") %>%
  filter(!is.na(UNITCD), !is.na(BALIVE), !is.na(tcc), tcc >= 0, tcc <= 100)

cat(sprintf("plots in analysis frame: %d across %d survey units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
```

## Define the 4-way stratifications

TCC is a single **BALIVE-correlated** axis (canopy cover ≈ a structural proxy for
basal area) cut into 4 ordered bins, so its strata are naturally homogeneous in
BALIVE. An unsupervised k-means on the embedding is *not* comparable: it splits
the 48-D space along its highest-variance directions — forest type/species — which
need not align with BALIVE magnitude. We therefore report **two** embedding
stratifications:

- **k-means (unsupervised):** 4 regions of `z_type` geometry, agnostic to BALIVE.
  A reference for "does raw embedding geometry stratify BALIVE?"
- **supervised (the fair analog to TCC):** a single BALIVE-predictive score read
  out of the embedding, cut into 4 quantile bins. The score is an **out-of-fold**
  (5-fold) linear fit of BALIVE on the full embedding, so every plot is scored by
  a model that never saw it — no in-sample optimism that TCC also never gets.

```{r strata}
# --- TCC strata: canopy-cover percent bins ----------------------------------
analysis <- analysis %>%
  mutate(tcc_stratum = cut(tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                           labels = c("0-10", "11-48", "49-85", "86-100")))

# --- Embedding strata, UNSUPERVISED: 4 k-means regions of z_type space -------
set.seed(1)
Z  <- scale(as.matrix(analysis[, emb_type_vars]))
km <- kmeans(Z, centers = 4, nstart = 25)
analysis$emb_km <- factor(km$cluster, labels = paste0("K", 1:4))

# --- Embedding strata, SUPERVISED: quantile bins of a cross-fit BALIVE read-out
# Out-of-fold OLS of BALIVE on the full embedding (type + phase), then equal-count
# quantile bins -- the learned analog of "a BALIVE-correlated covariate binned
# into 4," matched to how TCC is used.
set.seed(1)
folds    <- sample(rep(1:5, length.out = nrow(analysis)))
form_bal <- reformulate(emb_vars, response = "BALIVE")
oof      <- numeric(nrow(analysis))
for (f in 1:5) {
  fit <- lm(form_bal, data = analysis[folds != f, ])
  oof[folds == f] <- predict(fit, newdata = analysis[folds == f, ])
}
analysis$emb_score <- oof
analysis$emb_sup   <- factor(dplyr::ntile(oof, 4), labels = paste0("S", 1:4))

cat("TCC stratum counts:\n");                 print(table(analysis$tcc_stratum))
cat("\nEmbedding k-means (unsup) counts:\n");  print(table(analysis$emb_km))
cat("\nEmbedding supervised (OOF) counts:\n"); print(table(analysis$emb_sup))
```

## Within-stratum variance of BALIVE

`pooled_within()` returns the residual `BALIVE` variance after removing cell
means — `SS_within / (N − H)` for cells defined by the grouping columns. Strata
are always crossed with `UNITCD`, so a "cell" is one `(unit, stratum)`
combination. Empty and singleton cells are handled correctly (they contribute 0
to `SS` and reduce the degrees of freedom).

```{r within-var-fn}
pooled_within <- function(df, groups) {
  g <- df %>%
    group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((BALIVE - mean(BALIVE))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
```

### Headline: overall pooled within-stratum variance

Baseline is `UNITCD`-only. Lower `within_var` (and `pct_of_baseline < 100`) means
the sub-stratification made `BALIVE` more homogeneous within cells — i.e. a more
efficient post-stratified estimator.

```{r headline}
v_unit <- pooled_within(analysis, "UNITCD")
v_tcc  <- pooled_within(analysis, c("UNITCD", "tcc_stratum"))
v_km   <- pooled_within(analysis, c("UNITCD", "emb_km"))
v_sup  <- pooled_within(analysis, c("UNITCD", "emb_sup"))

data.frame(
  stratification  = c("UNITCD only (baseline)", "UNITCD x TCC (4)",
                      "UNITCD x embedding k-means (4, unsup)",
                      "UNITCD x embedding supervised (4, OOF)"),
  within_var      = round(c(v_unit, v_tcc, v_km, v_sup), 1),
  within_sd       = round(sqrt(c(v_unit, v_tcc, v_km, v_sup)), 2),
  pct_of_baseline = round(100 * c(v_unit, v_tcc, v_km, v_sup) / v_unit, 1)
)
```

### Per survey unit: TCC vs embedding

For each `UNITCD`, the unstratified variance and the within-stratum variance
under each 4-way scheme (`*_ratio` = fraction of the unit's variance that
remains; smaller is better).

```{r per-unit}
per_unit <- analysis %>%
  group_by(UNITCD) %>%
  group_modify(~ tibble(
    n         = nrow(.x),
    var_total = var(.x$BALIVE),
    var_tcc   = pooled_within(.x, "tcc_stratum"),
    var_km    = pooled_within(.x, "emb_km"),
    var_emb   = pooled_within(.x, "emb_sup")   # supervised = the fair contender
  )) %>%
  ungroup() %>%
  mutate(tcc_ratio = round(var_tcc / var_total, 3),
         emb_ratio = round(var_emb / var_total, 3),
         winner    = if_else(var_emb < var_tcc, "embedding", "tcc"),
         across(c(var_total, var_tcc, var_km, var_emb), ~ round(.x, 1)))
per_unit
```

### Per-cell detail (the within-strata variances)

Mean, variance, and n of `BALIVE` in every `(unit, stratum)` cell — the raw
within-strata variances, for both stratifications.

```{r per-cell}
cell_tcc <- analysis %>%
  group_by(UNITCD, tcc_stratum) %>%
  summarize(n = n(), mean_balive = round(mean(BALIVE), 1),
            var_balive = round(var(BALIVE), 1), .groups = "drop")

cell_emb <- analysis %>%
  group_by(UNITCD, emb_sup) %>%
  summarize(n = n(), mean_balive = round(mean(BALIVE), 1),
            var_balive = round(var(BALIVE), 1), .groups = "drop")

cat("=== BALIVE within (UNITCD x TCC stratum) ===\n"); cell_tcc
cat("\n=== BALIVE within (UNITCD x embedding supervised stratum) ===\n"); cell_emb
```

### Visual: within-stratum SD by unit

```{r plot-compare, fig.width=9, fig.height=5}
op <- par(mar = c(4.5, 4.5, 3, 1))
ymax <- sqrt(max(per_unit$var_total, na.rm = TRUE)) * 1.05
plot(seq_len(nrow(per_unit)), sqrt(per_unit$var_total), type = "n",
     ylim = c(0, ymax), xaxt = "n",
     xlab = "survey unit (UNITCD)", ylab = "within-stratum SD of BALIVE (sq ft/acre)",
     main = "Post-stratification: TCC vs embedding, by survey unit")
axis(1, at = seq_len(nrow(per_unit)), labels = per_unit$UNITCD)
points(seq_len(nrow(per_unit)), sqrt(per_unit$var_total), pch = 1,  col = "grey50", cex = 1.4)
points(seq_len(nrow(per_unit)), sqrt(per_unit$var_tcc),   pch = 16, col = "firebrick", cex = 1.4)
points(seq_len(nrow(per_unit)), sqrt(per_unit$var_km),    pch = 4,  col = "grey40",    cex = 1.2)
points(seq_len(nrow(per_unit)), sqrt(per_unit$var_emb),   pch = 16, col = "steelblue", cex = 1.4)
legend("topright", c("unstratified", "TCC (4)", "embedding k-means (4)", "embedding supervised (4)"),
       pch = c(1, 16, 4, 16), col = c("grey50", "firebrick", "grey40", "steelblue"), bty = "n")
par(op)
```

## Cleanup

```{r cleanup}
dbDisconnect(con)
```

## Notes and design choices

- **Why pooled within-stratum variance.** Under proportional allocation and an
  equal-probability sample, it is (up to the `1/n` factor) the variance of the
  post-stratified estimator, so it is the direct efficiency criterion. The
  `pct_of_baseline` / `*_ratio` columns are the variance-reduction achieved over
  `UNITCD`-only estimation.
- **Like-for-like.** Both schemes use exactly 4 strata on the same plots, crossed
  with `UNITCD`. Giving the embedding more clusters would trivially lower variance
  (more, smaller cells), so 4-vs-4 is the fair comparison.
- **Why two embedding stratifications.** Unsupervised k-means splits the embedding
  along its highest-variance directions (forest type/species), which need not align
  with BALIVE — so it is expected to stratify BALIVE weakly, and is kept only as a
  geometry reference. The supervised quantile stratifier is the fair comparison to
  TCC. It is *not* circular: the score is fit **out-of-fold**, and post-stratification
  by definition uses a target-correlated covariate — TCC is simply one whose
  correlation is built in rather than learned. If TCC still wins, that is a real
  result: canopy cover is a near-direct structural measurement of basal area, while
  the embedding was trained for forest-type similarity, not BALIVE regression.
- **Combining them** (e.g. TCC × embedding, or embedding strata *within* TCC bins)
  is the natural next step if you want the embedding to add resolution on top of the
  structural signal TCC already captures.
- **Not design-based (yet).** This uses the proportional-allocation approximation
  `W_h ≈ n_h/n`. `POP_STRATUM` (pulled above) carries FIA's official
  area-expansion factors; a fully design-based variance would weight strata by
  those areas rather than plot counts.
- **k-means caveat.** Clusters depend on the seed and on standardization; with
  `nstart = 25` and `set.seed(1)` the solution is stable, but report the seed.
