---
title: "FIA Hierarchical Type/Phase Strata for BALIVE"
subtitle: "20 type bins -> phase subdivision -> 5 BALIVE-homogeneous strata, Virginia FIA plots"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  k_type: 20                 # number of z_type bins
  phase_subdivide_frac: 0.5  # fraction of type bins (highest phase variability) to split
  k_phase: 3                 # phase sub-groups per subdivided bin
  K_strata: 5                # final number of strata
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
```

## Overview

A **hierarchical, structure-aware** stratification of the embedding space for
estimating BALIVE:

1. **~`r params$k_type` type bins** — k-means on `z_type`.
2. **Phase variability per bin** — within each type bin, the spread of `z_phase`;
   rank bins by it.
3. **Subdivide the most-variable bins** — the top `r round(100*params$phase_subdivide_frac)`%
   each split into `r params$k_phase` phase sub-groups (k-means on `z_phase`);
   the rest stay whole. -> "fine bins".
4. **Aggregate to `r params$K_strata` strata** — Ward hierarchical clustering of
   the fine bins on their **mean BALIVE**, cut to `r params$K_strata`.

The idea: respect the type structure, refine only where the temporal (phase) axis
actually varies, and organize the result into a few BALIVE-homogeneous strata.
Evaluated by pooled within-stratum BALIVE variance (by `UNITCD`), against
baseline, TCC, and a *flat* supervised stratifier.

> **Optimism note.** Plot -> fine-bin assignment uses embeddings only (no leak).
> The 40 -> `r params$K_strata` aggregation uses bin-mean BALIVE — a coarse
> supervised step, so the optimism vs an out-of-fold build is small; treat the
> absolute numbers as slightly optimistic but the *comparison* as fair.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Data assembly

```{r assemble}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)

con <- dbConnect(RSQLite::SQLite(), params$fiadb)
cond_data <- dbGetQuery(
  con, sprintf("SELECT PLT_CN, CONDPROP_UNADJ, BALIVE FROM COND WHERE STATECD='%s'",
               params$statecd))
balive <- cond_data %>%
  mutate(PLT_CN = as.character(PLT_CN), BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")

emb_data <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  distinct(CN, .keep_all = TRUE)

fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID))

base <- balive %>%
  inner_join(select(emb_data, CN, all_of(c(emb_type_vars, emb_phase_vars))),
             by = c("PLT_CN" = "CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD, LON_ACTUAL, LAT_ACTUAL),
             by = c("PLT_CN" = "CN")) %>%
  filter(!is.na(UNITCD))

# TCC at the plot locations (categorical raster -> factor -> numeric; >100 = nodata)
tcc     <- rast(params$tcc)
pts     <- vect(base, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
pts_aea <- project(pts, crs(tcc))
tv <- extract(tcc, pts_aea, ID = FALSE)[[1]]
tv <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
base$tcc <- tv

analysis <- base %>% filter(!is.na(tcc), tcc >= 0, tcc <= 100)
cat(sprintf("plots: %d across %d survey units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
```

## Step 1 - Type bins

```{r type-bins}
set.seed(1)
Zt <- scale(as.matrix(analysis[, emb_type_vars]))
analysis$type_bin <- factor(kmeans(Zt, centers = params$k_type, nstart = 20)$cluster)
table(analysis$type_bin)
```

## Step 2 - Phase variability per type bin

Total within-bin spread of (standardized) `z_phase` = sum of per-dimension
variances. Higher = this forest type spans diverse temporal states.

```{r phase-var}
Zp <- scale(as.matrix(analysis[, emb_phase_vars]))

phase_var <- tapply(seq_len(nrow(analysis)), analysis$type_bin, function(idx)
  sum(apply(Zp[idx, , drop = FALSE], 2, var)))

n_sub    <- round(params$phase_subdivide_frac * params$k_type)
sub_bins <- names(sort(phase_var, decreasing = TRUE))[seq_len(n_sub)]

pv <- tibble(type_bin = names(phase_var),
             n = as.integer(table(analysis$type_bin)[names(phase_var)]),
             phase_var = round(as.numeric(phase_var), 2)) %>%
  arrange(desc(phase_var)) %>%
  mutate(subdivided = type_bin %in% sub_bins)
pv
```

## Step 3 - Subdivide high-phase-variability bins -> fine bins

Each selected bin (with enough plots) splits into `r params$k_phase` phase
sub-groups; others keep a single sub-group.

```{r subdivide}
analysis$phase_sub <- 1L
set.seed(1)
for (b in sub_bins) {
  idx <- which(analysis$type_bin == b)
  if (length(idx) >= 10 * params$k_phase) {         # enough points to split
    analysis$phase_sub[idx] <- kmeans(Zp[idx, , drop = FALSE],
                                      centers = params$k_phase, nstart = 10)$cluster
  }
}
analysis$fine_bin <- interaction(analysis$type_bin, analysis$phase_sub, drop = TRUE)
cat(sprintf("fine bins: %d (from %d type bins, %d subdivided)\n",
            nlevels(analysis$fine_bin), params$k_type, length(sub_bins)))
```

## Step 4 - Aggregate fine bins into strata (Ward on mean BALIVE)

```{r aggregate, fig.width=10, fig.height=5}
bin_stats <- analysis %>%
  group_by(fine_bin) %>%
  summarize(n = n(), mean_bal = mean(BALIVE), sd_bal = sd(BALIVE), .groups = "drop")

K  <- min(params$K_strata, nrow(bin_stats))
hc <- hclust(dist(scale(bin_stats$mean_bal)), method = "ward.D2")
bin_stats$stratum <- factor(cutree(hc, k = K))
analysis$stratum  <- bin_stats$stratum[match(analysis$fine_bin, bin_stats$fine_bin)]

# dendrogram of fine bins, leaves ordered by mean BALIVE
op <- par(mar = c(6, 4, 3, 1))
plot(hc, labels = round(bin_stats$mean_bal), xlab = "fine bin (mean BALIVE)",
     main = sprintf("Fine bins -> %d strata (Ward on mean BALIVE)", K), sub = "")
rect.hclust(hc, k = K, border = 2:(K + 1))
par(op)

# stratum summary
analysis %>%
  group_by(stratum) %>%
  summarize(n_plots = n(), n_fine_bins = dplyr::n_distinct(fine_bin),
            mean_bal = round(mean(BALIVE), 1), sd_bal = round(sd(BALIVE), 1),
            .groups = "drop") %>%
  arrange(mean_bal)
```

## Evaluation - within-stratum BALIVE variance

```{r eval}
resp <- "BALIVE"
pooled_within <- function(df, groups, y = resp) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
dh_strata <- function(x, K = 5, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(factor(rep(1L, length(x))))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  bnd <- sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
  factor(findInterval(x, c(-Inf, bnd, Inf)))
}

# reference stratifications at the same K
analysis$tcc_dh <- dh_strata(analysis$tcc, params$K_strata)

# flat supervised: out-of-fold BALIVE probe on the full embedding, DH-binned
set.seed(1)
folds <- sample(rep(1:5, length.out = nrow(analysis)))
emb_vars <- c(emb_type_vars, emb_phase_vars)
oof <- numeric(nrow(analysis))
form <- reformulate(emb_vars, response = "BALIVE")
for (f in 1:5) {
  fit <- lm(form, data = analysis[folds != f, ])
  oof[folds == f] <- predict(fit, newdata = analysis[folds == f, ])
}
analysis$emb_flat <- dh_strata(oof, params$K_strata)

v_unit <- pooled_within(analysis, "UNITCD")
v_tcc  <- pooled_within(analysis, c("UNITCD", "tcc_dh"))
v_flat <- pooled_within(analysis, c("UNITCD", "emb_flat"))
v_hier <- pooled_within(analysis, c("UNITCD", "stratum"))

vv <- c(v_unit, v_tcc, v_flat, v_hier)
data.frame(
  stratification  = c("UNITCD only (baseline)",
                      sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                      sprintf("UNITCD x embedding flat supervised (%d)", params$K_strata),
                      sprintf("UNITCD x hierarchical type/phase (%d)", params$K_strata)),
  within_var      = round(vv, 1),
  within_sd       = round(sqrt(vv), 2),
  pct_of_baseline = round(100 * vv / v_unit, 1)
)
```

## Removals evaluation (multi-target test)

Now score the **same fine-bin structure** on **removals** (annual per-acre removed
growing-stock volume), aggregated two ways:

1. **BALIVE-aggregated strata applied to removals** — the transfer / "one
   stratification for many targets" case (this is the `stratum` from Step 4).
2. **Removals-aggregated strata** — re-run the fine-bin -> `K` merge with **mean
   removals** as the clustering feature.

Removals come from `TREE_GRM_COMPONENT` and are only defined on remeasured plots,
so this runs on that subset (`ar`). Reference rows: TCC and a flat OOF removals
probe.

```{r removals-pull}
dbWriteTable(con, "tmp_plots", tibble(PLT_CN = analysis$PLT_CN),
             temporary = TRUE, overwrite = TRUE)
remv <- dbGetQuery(
  con,
  "SELECT g.PLT_CN,
          SUM(g.REMVCFGS_FOREST * g.SUBP_TPAREMV_UNADJ_GS_FOREST) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_plots t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))
# (if TREE_GRM_COMPONENT lacks PLT_CN, join TREE on g.TRE_CN = TREE.CN)

analysis <- analysis %>% left_join(remv, by = "PLT_CN")
ar <- analysis %>% filter(!is.na(remv_cfgs))       # remeasured plots (removals defined)
cat(sprintf("remeasured plots with removals: %d (%.1f%% with any removal)\n",
            nrow(ar), 100 * mean(ar$remv_cfgs > 0)))
```

```{r removals-strata}
K <- min(params$K_strata, dplyr::n_distinct(ar$fine_bin))

# (2) hierarchical strata re-aggregated by REMOVALS (same fine bins, new merge)
bin_remv <- ar %>% group_by(fine_bin) %>%
  summarize(mean_remv = mean(remv_cfgs), .groups = "drop")
hc_r <- hclust(dist(scale(bin_remv$mean_remv)), method = "ward.D2")
bin_remv$stratum_remv <- factor(cutree(hc_r, k = K))
ar$stratum_remv <- bin_remv$stratum_remv[match(ar$fine_bin, bin_remv$fine_bin)]

# reference: flat OOF removals probe + TCC, both DH-binned
set.seed(2)
fr <- sample(rep(1:5, length.out = nrow(ar)))
oofr <- numeric(nrow(ar)); formr <- reformulate(emb_vars, "remv_cfgs")
for (f in 1:5) { fit <- lm(formr, ar[fr != f, ]); oofr[fr == f] <- predict(fit, ar[fr == f, ]) }
ar$remv_flat <- dh_strata(oofr,   params$K_strata)
ar$tcc_dh    <- dh_strata(ar$tcc, params$K_strata)
```

```{r removals-eval}
vr_unit <- pooled_within(ar, "UNITCD",                    y = "remv_cfgs")
vr_tcc  <- pooled_within(ar, c("UNITCD", "tcc_dh"),       y = "remv_cfgs")
vr_bal  <- pooled_within(ar, c("UNITCD", "stratum"),      y = "remv_cfgs")  # BALIVE-agg (transfer)
vr_rem  <- pooled_within(ar, c("UNITCD", "stratum_remv"), y = "remv_cfgs")  # removals-agg
vr_flat <- pooled_within(ar, c("UNITCD", "remv_flat"),    y = "remv_cfgs")

vr <- c(vr_unit, vr_tcc, vr_bal, vr_rem, vr_flat)
data.frame(
  stratification  = c("UNITCD only (baseline)",
                      sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                      sprintf("UNITCD x hierarchical, BALIVE-aggregated (%d)", params$K_strata),
                      sprintf("UNITCD x hierarchical, removals-aggregated (%d)", params$K_strata),
                      sprintf("UNITCD x flat removals supervised (%d)", params$K_strata)),
  within_var      = round(vr, 2),
  within_sd       = round(sqrt(vr), 3),
  pct_of_baseline = round(100 * vr / vr_unit, 1)
)
```

## Cleanup

```{r cleanup}
dbDisconnect(con)
```

## Notes

- **Phase-variability metric.** Total within-bin `z_phase` variance (trace of the
  covariance). Swap for mean pairwise distance or a per-dimension max if you want
  a different notion of "temporal diversity".
- **Aggregation feature.** Ward clusters fine bins on **mean BALIVE** only. To make
  strata differ in spread as well as level, cluster on `c(mean_bal, sd_bal)` (scaled)
  or a vector of within-bin BALIVE quantiles.
- **Weighting.** Fine bins enter the clustering unweighted (each bin one point).
  For size-aware strata, replicate/weight bins by `n`, or post-balance strata sizes.
- **Fairness of the comparison.** All rows use `K_strata` strata crossed with
  `UNITCD`; TCC and the flat probe use DH bins. The hierarchical method is the only
  one using the type/phase structure, so a win over `emb_flat` means the structure
  (not just supervision) is buying you something.
- **Knobs.** `k_type`, `phase_subdivide_frac`, `k_phase`, `K_strata` are params;
  e.g. raise `k_type` for finer type resolution or `phase_subdivide_frac` to split
  more aggressively on the temporal axis.
