---
title: "FIA Embedding Evaluation: Learned Embeddings vs. Raw Inputs"
subtitle: "BALIVE regression and FORTYPCD classification on Virginia FIA plots"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  reticulate_python: "/home/nnagle/miniconda3/envs/r-tf/bin/python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi = 110
)

```

## Overview

This report is a **downstream evaluation** of the FoR-EST representation model. It
is *not* part of model training. It answers the core scientific question the
project exists to answer:

> Do the learned per-pixel embeddings (`z_type`, `z_phase`) beat the **raw
> satellite/CCDC inputs** (`x_type`, `x_phase`) for predicting real forest
> attributes measured on the ground by the USFS Forest Inventory and Analysis
> (FIA) program?

Two targets are evaluated:

- **`BALIVE`** — live tree basal area (sq ft/acre), a continuous target for
  regression / k-nearest-neighbor imputation.
- **`FORTYPCD`** — forest type code, a categorical target for classification.

The comparison recurs throughout — **`emb`** (learned embeddings) vs. **`inp`**
(raw inputs) — using linear probes, k-nearest-neighbor imputation (with metric
tuning), and forest-type classifiers. Because the model produces two separate
embeddings (`z_type` = *what kind of forest*, `z_phase` = *temporal recovery
dynamics*), the two blocks are handled and weighted separately throughout,
mirroring the dual-pathway architecture.

> **Note on the embedding source.** These are `frl_v0_exp033` embeddings. The
> project's `CLAUDE.md` declares the model "complete" as of `exp017`, so this is
> post-completion validation work on a later checkpoint — evaluation, not
> development.

Key paths and the target state are exposed as **knit parameters** (see the YAML
header): `params$emb_csv`, `params$fiadb`, `params$statecd`.

## Environment

```{r reticulate}
# Point reticulate at the conda env that carries TensorFlow/Keras.
Sys.setenv(RETICULATE_PYTHON = params$reticulate_python)

library(reticulate)
py_config()
```

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(FNN)      # get.knnx() for k-nearest-neighbor imputation
library(nnet)     # multinom() for the multinomial forest-type baseline

```

## Data assembly

### Load embeddings

The embeddings CSV is written by `frl/training/embed_locations.py`. `CN`/`ID` are
huge FIA control numbers; they are read as **character** and stripped of any
trailing `.0` so they are not mangled by float parsing.

```{r load-embeddings}
emb_data <- read_csv(
  params$emb_csv,
  col_types = cols(CN = "c", ID = "c")   # read as text, don't parse
) %>%
  mutate(
    CN = sub("\\.0$", "", CN),           # "1279297123456789.0" -> "1279297123456789"
    ID = sub("\\.0$", "", ID)
  )
```

Column families in the embeddings file, and the preprocessing each one already
carries (all applied inside `frl/training/embed_locations.py`, sampled at the
plot's center pixel):

The inputs are **not raw** despite the `x_` prefix: they carry whatever
`FeatureBuilder.build_feature` applied in the DataLoader worker — the *same*
transform the encoder was trained on. Crucially, the two input blocks are
normalized **differently**, controlled by whether the binding config gives the
feature a `covariance` block (`frl_binding_v1.yaml`):

| Prefix          | Block | Preprocessing baked into the value                                                                                                                                                                                                                     |
|-----------------|-------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `x_type_*`      | input | **Mahalanobis-whitened** (`type_encoder_input` has `covariance: calculate: true`): log pre-transforms on skewed channels → center → multiply by the whitening matrix (Cholesky of inverse covariance). Euclidean distance here **equals Mahalanobis distance** in the original space, so the block is centered *and decorrelated across channels*. Whitening subsumes the per-channel z-score/robust-IQR labels — they are not separately applied. Atemporal, center pixel. |
| `x_phase_*`     | input | Per-channel path (`phase_ccdc` has **no** covariance): mostly **z-score** (mean/std, clamp ±6), `temporal_position` left identity, `spectral_velocity` log-then-z-score. **Not whitened / not decorrelated.** Taken at the measurement year. *Not used below.* |
| `x_phase_dm_*`  | input | Same per-channel z-score as `x_phase`, then **temporally demeaned** (per-channel mean over all years subtracted before extracting the measurement year). This is what the report uses as `inp_phase_vars`.                                                |
| `z_type_*` (48) | emb   | **Raw model output**, center pixel — *no* L2 normalization (the type pathway leaves magnitude unconstrained by design).                                                                                                                                |
| `z_phase_*` (8) | emb   | **Post-FiLM** phase embedding at the measurement year (conditioned on `z_type` via FiLM). Written raw.                                                                                                                                                 |
| `g_type_*`      | emb   | SimCLR projection head applied to `z_type` (`g(h)`). Present in the file but not used in this report.                                                                                                                                                  |

**Where the normalization lives, and why the kNN preprocessing below is
asymmetric.** The DataLoader worker normalizes only the encoder **inputs**; the
`z_*` embeddings are the model's **outputs**, computed downstream of the
DataLoader and never passed through it (neither the type forward pass nor the raw
post-FiLM `gamma·h + beta` phase pass applies any output normalization). So the
`x_*` blocks already sit on a comparable, model-ready scale while the `z_*`
blocks — `z_type` especially — do not. That is why Section 2 applies a block-wise
population L2 normalization to the `z` blocks only and feeds the `x` blocks to
kNN as-is. Note the consequence for the distance metric: `x_type` is *already
whitened*, so unmodified Euclidean kNN on it is exactly Mahalanobis kNN, whereas
`x_phase` is only per-channel scaled (not decorrelated). The one further
transform this report adds on top is the scalar phase-weight `w` (Section 3),
which retunes the type-vs-phase balance.

### Pull FIA ground truth (COND table)

```{r fia-cond}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

cond_data <- dbGetQuery(
  con,
  sprintf(
    "SELECT PLT_CN, STATECD, FORTYPCD, FLDSZCD, CONDPROP_UNADJ,
            BALIVE, LIVE_CANOPY_CVR_PCT, NBR_LIVE_STEMS
       FROM COND
      WHERE STATECD='%s'",
    params$statecd
  )
)

```

### Reduce multi-condition plots to one row per plot

FIA plots can span multiple conditions. We collapse two ways:

- **area-weighted average** of `BALIVE` and canopy cover (weighted by
  `CONDPROP_UNADJ`, the condition's area proportion), and
- the **dominant condition's** `FORTYPCD` (largest area proportion).

```{r plot-reduce}
plot_cond_data_avg <- cond_data %>%
  semi_join(emb_data, by = c("PLT_CN" = "CN")) %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(
    BALIVE              = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ),
    LIVE_CANOPY_CVR_PCT = sum(LIVE_CANOPY_CVR_PCT * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ)
  ) %>%
  ungroup()

plot_cond_data_max <- cond_data %>%
  semi_join(emb_data, by = c("PLT_CN" = "CN")) %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  slice_max(order_by = CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(PLT_CN, FORTYPCD)
```

### Pull removals (GRM), a second continuous target

`BALIVE` (from `COND`) is a *stock*. Removals are a *flux* — annual per-acre
removed growing-stock volume — and live in `TREE_GRM_COMPONENT`, defined only on
**remeasured** plots. A remeasured plot with no cut trees is a genuine `0`;
plots not in the GRM table get `NA` and are excluded from the removals kNN in
Section 2b (they have no removals to predict). This mirrors the `remv_cfgs`
target and population used in `fia_knn_models.Rmd`.

```{r fia-removals}
dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb_data$CN)),
             temporary = TRUE, overwrite = TRUE)
remv_data <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))
```

### Join to a single regression frame

Every downstream analysis (linear probes, kNN imputation, forest-type
classification) is restricted to plots with a **known forest type**: rows with a
missing `FORTYPCD` are dropped here at the source, so the whole report runs on the
same forest-type population. `remv_cfgs` is joined on as well (left join, so
non-remeasured plots carry `NA` and are filtered only where removals are used).

```{r reg-data}
reg_data <- plot_cond_data_avg %>%
  left_join(plot_cond_data_max, by = "PLT_CN") %>%
  inner_join(emb_data, by = c("PLT_CN" = "CN")) %>%
  left_join(remv_data, by = "PLT_CN") %>%          # removals: NA off the GRM/remeasured panel
  filter(!is.na(FORTYPCD)) %>%                     # only plots with a forest type
  mutate(FORTYPGRP = as.factor(substr(FORTYPCD, 1, 2)))

```

### Feature-block name vectors

```{r feature-vars}
inp_type_vars  <- paste0("x_type_",     seq(0, 32))
inp_phase_vars <- paste0("x_phase_dm_", seq(0, 12))
emb_type_vars  <- paste0("z_type_",     seq(0, 47))
emb_phase_vars <- paste0("z_phase_",    seq(0, 7))
```

## 1. Linear probes (OLS)

How much of `BALIVE` is *linearly* recoverable from each representation? Adjusted
R² is the headline. This is the classic linear-probe comparison: raw inputs vs.
embeddings, type-only vs. phase-only vs. combined.

```{r lm-formulas}
form_inp_type <- as.formula(paste("BALIVE ~", paste(inp_type_vars, collapse = " + ")))
form_inp_phase <- as.formula(paste("BALIVE ~", paste(inp_phase_vars, collapse = " + ")))
form_inp_all  <- as.formula(paste("BALIVE ~", paste(c(inp_type_vars, inp_phase_vars), collapse = " + ")))

form_emb_type <- as.formula(paste("BALIVE ~", paste(emb_type_vars, collapse = " + ")))
form_emb_phase <- as.formula(paste("BALIVE ~", paste(emb_phase_vars, collapse = " + ")))
form_emb_all  <- as.formula(paste("BALIVE ~", paste(c(emb_type_vars, emb_phase_vars), collapse = " + ")))

# rows with a known forest type
rd <- reg_data[!is.na(reg_data$FORTYPCD), ]

```

```{r lm-fits}
summary(lm(form_inp_type,  data = rd))
summary(lm(form_inp_phase, data = rd))
summary(lm(form_inp_all,   data = rd))

summary(lm(form_emb_phase, data = rd))
summary(lm(form_emb_all,   data = rd))
summary(lm(form_emb_type,  data = rd))
```

### Adjusted-R² summary table

```{r lm-r2}
r2 <- function(form, data = rd) summary(lm(form, data = data))$adj.r.squared

data.frame(
  block = c("type", "phase", "all"),
  emb   = c(r2(form_emb_type), r2(form_emb_phase), r2(form_emb_all)),
  inp   = c(r2(form_inp_type), r2(form_inp_phase), r2(form_inp_all))
)
```

## 2. k-Nearest-Neighbor imputation of BALIVE

This is the most scientifically honest test for the project's stated goal: the
embedding is explicitly designed as a **metric space** for kNN-style
post-stratification and small-area estimation, not for linear models.

### Split by the model's train/val/test assignment

```{r knn-splits}
AnnDat_train <- reg_data[which(reg_data$split == "train" & !is.na(reg_data$FORTYPCD)), ]
AnnDat_val   <- reg_data[which(reg_data$split == "val"   & !is.na(reg_data$FORTYPCD)), ]
AnnDat_test  <- reg_data[which(reg_data$split == "test"  & !is.na(reg_data$FORTYPCD)), ]
```

### Block-wise population L2 normalization

`z_type` is deliberately *unconstrained in magnitude* (per the architecture),
while `z_phase` lives on a smaller scale. To put them on comparable footing
before concatenation, each block is divided by its **mean training L2 norm** and
then row-normalized — separately.

```{r pop-l2}
l2_norm      <- function(x) sqrt(rowSums(x^2))
l2_normalize <- function(x) x / l2_norm(x)

pop_l2_normalize <- function(train_block, val_block, test_block) {
  mean_train_norm <- mean(l2_norm(train_block), na.rm = TRUE)
  list(
    train = l2_normalize(train_block / mean_train_norm),
    val   = l2_normalize(val_block   / mean_train_norm),
    test  = l2_normalize(test_block  / mean_train_norm)
  )
}

type_emb  <- pop_l2_normalize(AnnDat_train[, emb_type_vars],
                              AnnDat_val[,   emb_type_vars],
                              AnnDat_test[,  emb_type_vars])
phase_emb <- pop_l2_normalize(AnnDat_train[, emb_phase_vars],
                              AnnDat_val[,   emb_phase_vars],
                              AnnDat_test[,  emb_phase_vars])

```

### kNN prediction helper

`BALIVE` of a plot is imputed as the **mean BALIVE of its k train neighbors**.
Train-on-train drops the self-match.

```{r knn-helper}
knn_balive <- function(nn, balive_source, drop_self = FALSE) {
  idx <- if (drop_self) nn$nn.index[, -1] else nn$nn.index
  rowMeans(matrix(balive_source[idx], nrow = nrow(nn$nn.index)), na.rm = TRUE)
}

mse <- function(actual, predicted) mean((actual - predicted)^2, na.rm = TRUE)
```

### Baseline kNN (k = 10), embeddings vs. inputs

```{r knn-baseline}
emb_train <- cbind(type_emb$train, phase_emb$train)
emb_val   <- cbind(type_emb$val,   phase_emb$val)
emb_test  <- cbind(type_emb$test,  phase_emb$test)

nn_emb_train <- get.knnx(data = emb_train, query = emb_train, k = 11)
nn_emb_val   <- get.knnx(data = emb_train, query = emb_val,   k = 10)
nn_emb_test  <- get.knnx(data = emb_train, query = emb_test,  k = 10)

nn_inp_train <- get.knnx(data = AnnDat_train[, c(inp_type_vars, inp_phase_vars)],
                         query = AnnDat_train[, c(inp_type_vars, inp_phase_vars)], k = 11)
nn_inp_val   <- get.knnx(data = AnnDat_train[, c(inp_type_vars, inp_phase_vars)],
                         query = AnnDat_val[,   c(inp_type_vars, inp_phase_vars)], k = 10)
nn_inp_test  <- get.knnx(data = AnnDat_train[, c(inp_type_vars, inp_phase_vars)],
                         query = AnnDat_test[,  c(inp_type_vars, inp_phase_vars)], k = 10)

pred_emb_train <- knn_balive(nn_emb_train, AnnDat_train$BALIVE, drop_self = TRUE)
pred_emb_val   <- knn_balive(nn_emb_val,   AnnDat_train$BALIVE, drop_self = FALSE)
pred_emb_test  <- knn_balive(nn_emb_test,  AnnDat_train$BALIVE, drop_self = FALSE)

pred_inp_train <- knn_balive(nn_inp_train, AnnDat_train$BALIVE, drop_self = TRUE)
pred_inp_val   <- knn_balive(nn_inp_val,   AnnDat_train$BALIVE, drop_self = FALSE)
pred_inp_test  <- knn_balive(nn_inp_test,  AnnDat_train$BALIVE, drop_self = FALSE)

data.frame(
  split   = c("train", "val", "test"),
  mse_emb = c(mse(AnnDat_train$BALIVE, pred_emb_train),
              mse(AnnDat_val$BALIVE,   pred_emb_val),
              mse(AnnDat_test$BALIVE,  pred_emb_test)),
  mse_inp = c(mse(AnnDat_train$BALIVE, pred_inp_train),
              mse(AnnDat_val$BALIVE,   pred_inp_val),
              mse(AnnDat_test$BALIVE,  pred_inp_test))
)

```

## 2b. k-Nearest-Neighbor imputation of removals

`BALIVE` above is a *stock*; this repeats the kNN test for **removals**
(`remv_cfgs`), the *flux* target, on the remeasured/GRM subset. Removals are
strongly **zero-inflated and right-skewed** — most plots have no cut — so MSE is
dominated by the zero mass and a handful of large cuts, and R² runs low. Read the
**embedding-vs-input gap**, not the absolute level. The block-L2 normalization and
`k` are refit on this population's train rows; `k` is tuned on validation MSE.

```{r knn-removals-prep}
# remeasured subset of each split (removals defined; NA = off the GRM panel)
rd_r_train <- AnnDat_train[!is.na(AnnDat_train$remv_cfgs), ]
rd_r_val   <- AnnDat_val[  !is.na(AnnDat_val$remv_cfgs),   ]
rd_r_test  <- AnnDat_test[ !is.na(AnnDat_test$remv_cfgs),  ]

# block-L2-normalized embedding, refit on THIS population's train rows
type_emb_r  <- pop_l2_normalize(rd_r_train[, emb_type_vars],  rd_r_val[, emb_type_vars],  rd_r_test[, emb_type_vars])
phase_emb_r <- pop_l2_normalize(rd_r_train[, emb_phase_vars], rd_r_val[, emb_phase_vars], rd_r_test[, emb_phase_vars])
emb_r_train <- cbind(type_emb_r$train, phase_emb_r$train)
emb_r_val   <- cbind(type_emb_r$val,   phase_emb_r$val)
emb_r_test  <- cbind(type_emb_r$test,  phase_emb_r$test)

# raw inputs (already model-ready scale) as-is
inp_r_train <- rd_r_train[, c(inp_type_vars, inp_phase_vars)]
inp_r_val   <- rd_r_val[,   c(inp_type_vars, inp_phase_vars)]
inp_r_test  <- rd_r_test[,  c(inp_type_vars, inp_phase_vars)]

cat(sprintf("removals kNN population: train=%d val=%d test=%d | mean(train)=%.3f  frac_zero(train)=%.2f\n",
            nrow(rd_r_train), nrow(rd_r_val), nrow(rd_r_test),
            mean(rd_r_train$remv_cfgs), mean(rd_r_train$remv_cfgs == 0)))
```

```{r knn-removals-eval}
r2v <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
kgrid_r <- c(5, 10, 15, 20, 30, 50, 100)
kgrid_r <- kgrid_r[kgrid_r < nrow(rd_r_train)]     # k must not exceed the reference size

# impute removals as the mean of the k train neighbors; tune k on val, score test
knn_rmv <- function(ref, qval, qtest) {
  y_tr <- rd_r_train$remv_cfgs
  val_mse <- sapply(kgrid_r, function(k)
    mse(rd_r_val$remv_cfgs, knn_balive(get.knnx(ref, qval, k = k), y_tr)))
  kbest  <- kgrid_r[which.min(val_mse)]
  p_test <- knn_balive(get.knnx(ref, qtest, k = kbest), y_tr)
  data.frame(k_best = kbest,
             val_MSE  = round(min(val_mse), 3),
             test_MSE = round(mse(rd_r_test$remv_cfgs, p_test), 3),
             test_R2  = round(r2v(rd_r_test$remv_cfgs, p_test), 3))
}

rbind(
  cbind(representation = "embedding", knn_rmv(emb_r_train, emb_r_val, emb_r_test)),
  cbind(representation = "raw input", knn_rmv(inp_r_train, inp_r_val, inp_r_test))
)
```

## 3. Tuning the type-vs-phase scale, then sweeping k

Because type and phase are separate blocks, a scalar weight `w` on the phase
block controls how much the recovery-dynamics signal contributes to the distance
metric. `w` is chosen to minimize **validation** MSE — separately for embeddings
and inputs (inputs use a 2× grid because they sit on a different scale).

```{r w-sweep-emb}
k <- 150
w_grid <- seq(0.5, 6, by = 0.25)

mse_w <- sapply(w_grid, function(w) {
  emb_train_w <- cbind(type_emb$train, w * phase_emb$train)
  emb_val_w   <- cbind(type_emb$val,   w * phase_emb$val)
  nn_w <- get.knnx(data = emb_train_w, query = emb_val_w, k = k)
  mse(AnnDat_val$BALIVE, knn_balive(nn_w, AnnDat_train$BALIVE, drop_self = FALSE))
})
w_opt <- w_grid[which.min(mse_w)]
data.frame(w = w_grid, mse_val = mse_w)
cat("Optimal w (embedding):", w_opt, "\n")
```

```{r w-sweep-inp}
w_grid_inp <- w_grid * 2
mse_w_inp <- sapply(w_grid_inp, function(w) {
  inp_train_w <- cbind(AnnDat_train[, inp_type_vars], w * AnnDat_train[, inp_phase_vars])
  inp_val_w   <- cbind(AnnDat_val[,   inp_type_vars], w * AnnDat_val[,   inp_phase_vars])
  nn_w <- get.knnx(data = inp_train_w, query = inp_val_w, k = k)
  mse(AnnDat_val$BALIVE, knn_balive(nn_w, AnnDat_train$BALIVE, drop_self = FALSE))
})
w_opt_inp <- w_grid_inp[which.min(mse_w_inp)]
data.frame(w = w_grid_inp, mse_val = mse_w_inp)
cat("Optimal w (input):", w_opt_inp, "\n")

```

### k-sweep with per-k phase-weight re-optimization

For each neighborhood size k, the phase weight is re-optimized on val, then every
split is predicted from the **train** reference set. MSE is decomposed into
bias² + variance. The headline question: *at each k, does the embedding metric
give lower test error than raw inputs?*

```{r k-sweep}
w_grid     <- seq(0.5, 1.3, by = 0.1)
w_grid_inp <- seq(8,14,by=.5)
k_grid     <- c(10, 15, 20, 30, 50, 75, 100, 150, 200, 250)

metrics <- function(actual, predicted) {
  resid <- predicted - actual
  resid <- resid[!is.na(resid)]
  bias  <- mean(resid)
  m     <- mean(resid^2)
  mae   <- mean(abs(resid))
  c(mse = m, mae = mae, bias = bias, variance = m - bias^2)
}

optimize_w <- function(k, type_train, phase_train, type_val, phase_val,
                       bal_train, bal_val, w_candidates) {
  mse_w <- sapply(w_candidates, function(w) {
    tr <- cbind(type_train, w * phase_train)
    va <- cbind(type_val,   w * phase_val)
    nn <- get.knnx(data = tr, query = va, k = k)
    mse(bal_val, knn_balive(nn, bal_train, drop_self = FALSE))
  })
  w_candidates[which.min(mse_w)]
}

results_k <- list()
for (k in k_grid) {
  w_opt_emb <- optimize_w(k, type_emb$train, phase_emb$train, type_emb$val, phase_emb$val,
                          AnnDat_train$BALIVE, AnnDat_val$BALIVE, w_grid)
  w_opt_inp <- optimize_w(k, AnnDat_train[, inp_type_vars], AnnDat_train[, inp_phase_vars],
                          AnnDat_val[, inp_type_vars], AnnDat_val[, inp_phase_vars],
                          AnnDat_train$BALIVE, AnnDat_val$BALIVE, w_grid_inp)
  cat(sprintf("k = %-3d  optimal w:  emb = %.2f   inp = %.2f\n", k, w_opt_emb, w_opt_inp))

  emb_train <- cbind(type_emb$train, w_opt_emb * phase_emb$train)
  emb_val   <- cbind(type_emb$val,   w_opt_emb * phase_emb$val)
  emb_test  <- cbind(type_emb$test,  w_opt_emb * phase_emb$test)
  inp_train <- cbind(AnnDat_train[, inp_type_vars], w_opt_inp * AnnDat_train[, inp_phase_vars])
  inp_val   <- cbind(AnnDat_val[,   inp_type_vars], w_opt_inp * AnnDat_val[,   inp_phase_vars])
  inp_test  <- cbind(AnnDat_test[,  inp_type_vars], w_opt_inp * AnnDat_test[,  inp_phase_vars])

  pred_emb_train <- knn_balive(get.knnx(emb_train, emb_train, k = k + 1), AnnDat_train$BALIVE, drop_self = TRUE)
  pred_emb_val   <- knn_balive(get.knnx(emb_train, emb_val,   k = k),     AnnDat_train$BALIVE, drop_self = FALSE)
  pred_emb_test  <- knn_balive(get.knnx(emb_train, emb_test,  k = k),     AnnDat_train$BALIVE, drop_self = FALSE)
  pred_inp_train <- knn_balive(get.knnx(inp_train, inp_train, k = k + 1), AnnDat_train$BALIVE, drop_self = TRUE)
  pred_inp_val   <- knn_balive(get.knnx(inp_train, inp_val,   k = k),     AnnDat_train$BALIVE, drop_self = FALSE)
  pred_inp_test  <- knn_balive(get.knnx(inp_train, inp_test,  k = k),     AnnDat_train$BALIVE, drop_self = FALSE)

  results_k[[length(results_k) + 1]] <- do.call(rbind, list(
    data.frame(k = k, rep = "emb", w = w_opt_emb, split = "train", t(metrics(AnnDat_train$BALIVE, pred_emb_train))),
    data.frame(k = k, rep = "emb", w = w_opt_emb, split = "val",   t(metrics(AnnDat_val$BALIVE,   pred_emb_val))),
    data.frame(k = k, rep = "emb", w = w_opt_emb, split = "test",  t(metrics(AnnDat_test$BALIVE,  pred_emb_test))),
    data.frame(k = k, rep = "inp", w = w_opt_inp, split = "train", t(metrics(AnnDat_train$BALIVE, pred_inp_train))),
    data.frame(k = k, rep = "inp", w = w_opt_inp, split = "val",   t(metrics(AnnDat_val$BALIVE,   pred_inp_val))),
    data.frame(k = k, rep = "inp", w = w_opt_inp, split = "test",  t(metrics(AnnDat_test$BALIVE,  pred_inp_test)))
  ))
}

results_final <- do.call(rbind, results_k)
results_final[, c("mse", "mae", "bias", "variance")] <-
  round(results_final[, c("mse", "mae", "bias", "variance")], 1)
rownames(results_final) <- NULL
results_final <- results_final[order(results_final$rep, results_final$split, results_final$k), ]
results_final
```

### Headline test-error comparisons by k

`delta_pct < 0` means the embedding wins.

```{r headline-tables}
headline <- function(metric) {
  tw <- reshape(subset(results_final, split == "test", c("k", "rep", metric)),
                idvar = "k", timevar = "rep", direction = "wide")
  names(tw) <- sub(paste0("^", metric, "\\."), paste0(metric, "_"), names(tw))
  tw$delta_pct <- round(100 * (tw[[paste0(metric, "_emb")]] - tw[[paste0(metric, "_inp")]]) /
                          tw[[paste0(metric, "_inp")]], 1)
  tw
}

cat("Test MSE by k (delta_pct < 0 = embedding wins):\n"); headline("mse")
cat("Test MAE by k (delta_pct < 0 = embedding wins):\n"); headline("mae")
cat("Test Var by k (delta_pct < 0 = embedding wins):\n"); headline("variance")

```

## 4. Where in neighbor space does the embedding win?

A direct probe of the distance metric **by scale**, independent of the kNN
averaging step. Query = test, reference = train. For the r-th nearest neighbor,
how close is its BALIVE to the query's? The `gap = input − embedding` curve shows
at what neighbor rank the embedding's structure is most advantageous.

```{r perrank-setup}
R_max        <- 150
w_emb        <- 1.00
w_inp        <- 2.00      # k=50 optimum region; inp metric is w-sensitive
ranks_report <- c(1, 5, 10, 25, 50, 75, 100, 150)

bal_train <- AnnDat_train$BALIVE
bal_test  <- AnnDat_test$BALIVE

emb_ref <- cbind(type_emb$train, w_emb * phase_emb$train)
emb_qry <- cbind(type_emb$test,  w_emb * phase_emb$test)
inp_ref <- cbind(AnnDat_train[, inp_type_vars], w_inp * AnnDat_train[, inp_phase_vars])
inp_qry <- cbind(AnnDat_test[,  inp_type_vars], w_inp * AnnDat_test[,  inp_phase_vars])

nn_emb <- get.knnx(data = emb_ref, query = emb_qry, k = R_max)$nn.index
nn_inp <- get.knnx(data = inp_ref, query = inp_qry, k = R_max)$nn.index

absdelta_by_rank <- function(nn_index, bal_ref, bal_query) {
  neigh_bal <- matrix(bal_ref[nn_index], nrow = nrow(nn_index))
  abs(neigh_bal - bal_query)
}
ad_emb <- absdelta_by_rank(nn_emb, bal_train, bal_test)
ad_inp <- absdelta_by_rank(nn_inp, bal_train, bal_test)
```

```{r perrank-tables}
perrank <- data.frame(rank = 1:R_max,
                      emb = colMeans(ad_emb, na.rm = TRUE),
                      inp = colMeans(ad_inp, na.rm = TRUE))
perrank$gap       <- perrank$inp - perrank$emb
perrank$delta_pct <- round(100 * (perrank$emb - perrank$inp) / perrank$inp, 1)

M       <- matrix(0, R_max, R_max); M[upper.tri(M, diag = TRUE)] <- 1
cummean <- function(ad) colMeans(sweep(ad %*% M, 2, 1:R_max, "/"), na.rm = TRUE)
cumulative <- data.frame(rank = 1:R_max, emb = cummean(ad_emb), inp = cummean(ad_inp))
cumulative$delta_pct <- round(100 * (cumulative$emb - cumulative$inp) / cumulative$inp, 1)

cat("Per-rank mean |dBALIVE| to the r-th neighbor:\n")
round(perrank[perrank$rank %in% ranks_report, c("rank", "emb", "inp", "gap", "delta_pct")], 2)

cat("Cumulative mean |dBALIVE| within the first-r neighborhood:\n")
round(cumulative[cumulative$rank %in% ranks_report, ], 2)

peak <- perrank$rank[which.max(perrank$gap)]
cat(sprintf("Embedding advantage (per-rank gap) is largest at rank %d (%.2f sq ft/acre).\n",
            peak, max(perrank$gap)))

```

```{r perrank-plot, fig.width=11, fig.height=5}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 3, 1))
plot(perrank$rank, perrank$emb, type = "l", col = "steelblue", lwd = 2,
     ylim = range(perrank$emb, perrank$inp),
     xlab = "neighbor rank r", ylab = "mean |dBALIVE| to query (sq ft/acre)",
     main = "r-th neighbor homogeneity")
lines(perrank$rank, perrank$inp, col = "firebrick", lwd = 2)
legend("topleft", c("embedding", "input"), col = c("steelblue", "firebrick"), lwd = 2, bty = "n")

plot(perrank$rank, perrank$gap, type = "l", col = "darkgreen", lwd = 2,
     xlab = "neighbor rank r", ylab = "input - embedding |dBALIVE| (>0 = emb better)",
     main = "advantage by scale")
abline(h = 0, lty = 3); abline(v = peak, lty = 2, col = "grey50")
par(op)
```

### Banded (5-rank) paired advantage ± 2 SE

```{r banded, fig.width=11, fig.height=5}
band_w <- 5
band   <- (seq_len(R_max) - 1L) %/% band_w + 1L
band_emb <- tapply(colMeans(ad_emb, na.rm = TRUE), band, mean)
band_inp <- tapply(colMeans(ad_inp, na.rm = TRUE), band, mean)

D <- ad_inp - ad_emb   # >0 = embedding better, paired per query
band_stats <- as.data.frame(t(sapply(sort(unique(band)), function(b) {
  cols      <- which(band == b)
  per_query <- rowMeans(D[, cols, drop = FALSE], na.rm = TRUE)
  n  <- sum(is.finite(per_query))
  c(mid = mean(cols), gap = mean(per_query, na.rm = TRUE),
    se = sd(per_query, na.rm = TRUE) / sqrt(n))
})))
band_stats$delta_pct <- round(100 * (band_emb - band_inp) / band_inp, 2)
band_stats$sig       <- ifelse(abs(band_stats$gap) > 2 * band_stats$se, "*", "")

round(band_stats[, c("mid", "gap", "se", "delta_pct")], 3)
cat("bands where embedding is >2 SE better:",
    paste0(round(band_stats$mid[band_stats$sig == "*" & band_stats$gap > 0]), collapse = ", "), "\n")

op <- par(mfrow = c(1, 2), mar = c(4.2, 4.6, 3, 1))
plot(band_stats$mid, band_emb, type = "o", pch = 16, col = "steelblue", lwd = 2,
     ylim = range(band_emb, band_inp),
     xlab = "neighbor rank (5-rank bands)", ylab = "mean |dBALIVE| (sq ft/acre)",
     main = "banded homogeneity")
lines(band_stats$mid, band_inp, type = "o", pch = 16, col = "firebrick", lwd = 2)
legend("topleft", c("embedding", "input"), col = c("steelblue", "firebrick"),
       lwd = 2, pch = 16, bty = "n")
with(band_stats, {
  plot(mid, gap, type = "o", pch = 16, col = "darkgreen", lwd = 2,
       ylim = range(gap - 2 * se, gap + 2 * se, 0),
       xlab = "neighbor rank (5-rank bands)",
       ylab = "input - embedding |dBALIVE| (>0 = emb better)",
       main = "paired advantage +/- 2 SE")
  arrows(mid, gap - 2 * se, mid, gap + 2 * se, angle = 90, code = 3, length = 0.03, col = "darkgreen")
  abline(h = 0, lty = 3)
})
par(op)

```

## 5. Per-stratum error (zero class + positive deciles)

Stratifying into a zero class plus deciles of the positive values shows *where in
the distribution* the embedding wins, and whether that advantage grows with k
(the "regression to the mean" bias plot).

> **Caveat on this evaluation set.** In this VA / `exp033` FIA sample `BALIVE` is
> only **~1% zeros** (see the distribution plot in Section 6), so the zero
> stratum holds barely a dozen test plots — any per-stratum result *at the zero
> class* is noise-dominated and should not be read as a headline. (An earlier
> comment here claimed ~31% zeros; that figure came from a different pixel
> population and does not hold for these plot locations.) A second caveat: the
> `w_inp_*` phase weights below are **hardcoded**, not the val-tuned optima from
> Section 3 (which land near `w_inp ≈ 9–11`). The input baseline here therefore
> runs at a sub-optimal phase weight, so the per-stratum emb-vs-inp gap is
> indicative of *where* the two differ, not a fully fair magnitude — re-run with
> the tuned weights before quoting the stratum percentages.

```{r per-stratum, fig.width=11, fig.height=5}
k_lo <- 50 ; w_emb_lo <- 1.00; w_inp_lo <- 3.50
k_hi <- 250; w_emb_hi <- 0.75; w_inp_hi <- 2.00

bal_train <- AnnDat_train$BALIVE
bal_test  <- AnnDat_test$BALIVE
fin <- is.finite(bal_test)

q_full <- quantile(bal_test[fin], probs = seq(0, 1, 0.1), na.rm = TRUE)
edges  <- c(0, unique(q_full[q_full > 0]))
nbin   <- length(edges) - 1
strat  <- rep(NA_integer_, length(bal_test))
strat[fin & bal_test == 0] <- 0L
strat[fin & bal_test  > 0] <- findInterval(bal_test[fin & bal_test > 0], edges,
                                           rightmost.closed = TRUE, all.inside = TRUE)
S_levels <- 0:nbin
S_label  <- c("zero", paste0("d", (10 - nbin + 1):10))

predict_knn <- function(type_tr, phase_tr, type_te, phase_te, w, k) {
  ref <- cbind(type_tr, w * phase_tr)
  qry <- cbind(type_te, w * phase_te)
  knn_balive(get.knnx(data = ref, query = qry, k = k), bal_train, drop_self = FALSE)
}
P <- function(w, k, emb) if (emb)
  predict_knn(type_emb$train, phase_emb$train, type_emb$test, phase_emb$test, w, k) else
    predict_knn(AnnDat_train[, inp_type_vars], AnnDat_train[, inp_phase_vars],
                AnnDat_test[,  inp_type_vars], AnnDat_test[,  inp_phase_vars], w, k)

pred_emb_lo <- P(w_emb_lo, k_lo, TRUE); pred_inp_lo <- P(w_inp_lo, k_lo, FALSE)
pred_emb_hi <- P(w_emb_hi, k_hi, TRUE); pred_inp_hi <- P(w_inp_hi, k_hi, FALSE)

stratum_stats <- function(p) {
  r <- p - bal_test; g <- factor(strat, levels = S_levels)
  data.frame(bal_mean = as.numeric(tapply(bal_test, g, mean, na.rm = TRUE)),
             rmse     = sqrt(as.numeric(tapply(r, g, function(x) mean(x^2, na.rm = TRUE)))),
             bias     = as.numeric(tapply(r, g, mean, na.rm = TRUE)))
}
compare_k <- function(pe, pi, klab) {
  se <- stratum_stats(pe); si <- stratum_stats(pi)
  data.frame(k = klab, stratum = S_label, bal_mean = round(se$bal_mean, 1),
             n = as.integer(table(factor(strat, levels = S_levels))),
             rmse_emb = round(se$rmse, 1), rmse_inp = round(si$rmse, 1),
             rmse_dpct = round(100 * (se$rmse - si$rmse) / si$rmse, 1),
             bias_emb = round(se$bias, 1), bias_inp = round(si$bias, 1))
}
tab_lo <- compare_k(pred_emb_lo, pred_inp_lo, k_lo)
tab_hi <- compare_k(pred_emb_hi, pred_inp_hi, k_hi)

cat(sprintf("=== per-stratum, k = %d (w_emb=%.2f, w_inp=%.2f) ===\n", k_lo, w_emb_lo, w_inp_lo)); tab_lo
cat(sprintf("=== per-stratum, k = %d (w_emb=%.2f, w_inp=%.2f) ===\n", k_hi, w_emb_hi, w_inp_hi)); tab_hi

grow <- data.frame(stratum = S_label, bal_mean = tab_lo$bal_mean,
                   dpct_k50 = tab_lo$rmse_dpct, dpct_k250 = tab_hi$rmse_dpct,
                   change = round(tab_hi$rmse_dpct - tab_lo$rmse_dpct, 1))
cat("=== emb RMSE advantage by stratum: does it grow from k=50 to k=250? ===\n"); grow

op <- par(mfrow = c(1, 2), mar = c(4.4, 4.4, 3, 1))
plot(tab_lo$bal_mean, tab_lo$rmse_dpct, type = "o", pch = 16, col = "steelblue", lwd = 2,
     ylim = range(tab_lo$rmse_dpct, tab_hi$rmse_dpct, 0),
     xlab = "stratum mean BALIVE (x=0 = zero class)",
     ylab = "emb RMSE advantage %  (<0 = emb better)", main = "Where the embedding wins")
lines(tab_hi$bal_mean, tab_hi$rmse_dpct, type = "o", pch = 16, col = "darkorange", lwd = 2)
abline(h = 0, lty = 3)
legend("bottomleft", c(paste0("k=", k_lo), paste0("k=", k_hi)),
       col = c("steelblue", "darkorange"), lwd = 2, pch = 16, bty = "n")
plot(tab_hi$bal_mean, tab_hi$bias_inp, type = "o", pch = 16, col = "firebrick", lwd = 2,
     ylim = range(tab_hi$bias_emb, tab_hi$bias_inp, 0),
     xlab = "stratum mean BALIVE (x=0 = zero class)",
     ylab = sprintf("mean residual (pred-actual), k=%d", k_hi), main = "Regression to the mean")
lines(tab_hi$bal_mean, tab_hi$bias_emb, type = "o", pch = 16, col = "steelblue", lwd = 2)
abline(h = 0, lty = 3)
legend("topright", c("embedding", "input"), col = c("steelblue", "firebrick"),
       lwd = 2, pch = 16, bty = "n")
par(op)
```

## 6. Recovery-stage (ysfc) structure

`z_phase` was designed to encode recovery stage (years-since-fast-change, `ysfc`).
These diagnostics stress-test that pathway.

### BALIVE distribution by ysfc recovery class

```{r ysfc-hist, fig.width=11, fig.height=7}
bal  <- c(AnnDat_train$BALIVE, AnnDat_test$BALIVE)
ysfc <- c(AnnDat_train$ysfc,   AnnDat_test$ysfc)
if (exists("AnnDat_val")) { bal <- c(bal, AnnDat_val$BALIVE); ysfc <- c(ysfc, AnnDat_val$ysfc) }

brks <- c(-Inf, 1, 5, 15, 32, Inf)
labs <- c("recent (0-1)", "young (2-5)", "mid (6-15)", "mature (16-32)", "stable (33+)")
cls  <- cut(ysfc, brks, labels = labs)
ok   <- is.finite(bal) & !is.na(cls); bal <- bal[ok]; cls <- cls[ok]

xmax   <- ceiling(quantile(bal, 0.995, na.rm = TRUE) / 10) * 10
brks_h <- seq(0, xmax, by = 10)
bal_cl <- pmin(bal, xmax)
pal    <- c("#d73027", "#fc8d59", "#fee08b", "#91cf60", "#1a9850")

op <- par(mfrow = c(2, 3), mar = c(4, 4, 2.5, 1), oma = c(0, 0, 2, 0))
for (i in seq_along(labs)) {
  b <- bal_cl[cls == labs[i]]
  hist(b, breaks = brks_h, freq = FALSE, col = pal[i], border = "grey40",
       xlab = "BALIVE (sq ft/acre)", main = sprintf("%s  (n=%d)", labs[i], length(b)))
  abline(v = median(b), col = "black", lwd = 2, lty = 2)
  legend("topright", bty = "n", cex = 0.85, legend = c(
    sprintf("%%zero = %.0f%%", 100 * mean(b == 0)),
    sprintf("median = %.0f", median(b)),
    sprintf("mean = %.0f",   mean(b))))
}
dens <- lapply(labs, function(L) density(bal_cl[cls == L], from = 0))
plot(NA, xlim = c(0, xmax), ylim = c(0, max(sapply(dens, function(d) max(d$y)))),
     xlab = "BALIVE (sq ft/acre)", ylab = "density", main = "all classes")
for (i in seq_along(labs)) lines(dens[[i]], col = pal[i], lwd = 2)
legend("topright", labs, col = pal, lwd = 2, bty = "n", cex = 0.8)
mtext("BALIVE distribution by years-since-fast-change class", outer = TRUE, cex = 1.1, font = 2)
par(op)

```

### ysfc neighborhood coherence

Do an embedding's kNN neighbors share the query's recovery class? Purity, lift
over base rate (controls for the dominant stable class), and within-neighborhood
ysfc spread — emb vs. inp.

```{r ysfc-coherence, fig.width=13, fig.height=5}
k <- 25; w_emb <- 2.00; w_inp <- 3.50

brks <- c(-Inf, 2, 5, 10, 20, Inf)
labs <- c("recent(0-2)", "young(3-5)", "mid(6-10)", "mature(11-19)", "stable(20+)")
ysfc_train <- AnnDat_train$ysfc; ysfc_test <- AnnDat_test$ysfc
cls_train <- cut(ysfc_train, brks, labels = labs)
cls_test  <- cut(ysfc_test,  brks, labels = labs)

nn_emb <- get.knnx(cbind(type_emb$train, w_emb * phase_emb$train),
                   cbind(type_emb$test,  w_emb * phase_emb$test), k = k)$nn.index
nn_inp <- get.knnx(cbind(AnnDat_train[, inp_type_vars], w_inp * AnnDat_train[, inp_phase_vars]),
                   cbind(AnnDat_test[,  inp_type_vars], w_inp * AnnDat_test[,  inp_phase_vars]), k = k)$nn.index

nb_table <- function(nn_index) {
  qrep <- rep(cls_test, times = ncol(nn_index))
  ncls <- cls_train[as.vector(nn_index)]
  table(query = qrep, neighbor = ncls)
}
tab_emb <- nb_table(nn_emb); tab_inp <- nb_table(nn_inp)

report <- function(tab) {
  p_same <- diag(prop.table(tab, 1))
  base   <- colSums(tab) / sum(tab)
  rbind(purity = round(p_same, 3), lift = round(p_same / base, 2))
}
cat("== EMBEDDING ==\n"); report(tab_emb)
cat("== INPUT ==\n");     report(tab_inp)

nb_sd <- function(nn_index) apply(matrix(ysfc_train[nn_index], nrow = nrow(nn_index)), 1, sd, na.rm = TRUE)
sd_emb <- tapply(nb_sd(nn_emb), cls_test, mean, na.rm = TRUE)
sd_inp <- tapply(nb_sd(nn_inp), cls_test, mean, na.rm = TRUE)

pal <- c("#d73027", "#fc8d59", "#fee08b", "#91cf60", "#1a9850")
op <- par(mfrow = c(1, 3), mar = c(6, 4, 3, 1))
mosaicplot(tab_emb, color = pal, las = 2, cex.axis = 0.7,
           main = sprintf("Embedding (k=%d)", k), xlab = "query class", ylab = "neighbor class")
mosaicplot(tab_inp, color = pal, las = 2, cex.axis = 0.7,
           main = sprintf("Input (k=%d)", k), xlab = "query class", ylab = "neighbor class")
barplot(rbind(embedding = sd_emb, input = sd_inp), beside = TRUE, las = 2,
        col = c("steelblue", "firebrick"), ylab = "mean within-neighborhood ysfc SD (yrs)",
        main = "Neighbor ysfc spread")
legend("topright", c("embedding", "input"), fill = c("steelblue", "firebrick"), bty = "n")
par(op)
```

### BALIVE distribution + zero-inflated strata

```{r balive-strata, fig.width=9, fig.height=5}
bal_train <- AnnDat_train$BALIVE
bal_test  <- AnnDat_test$BALIVE
fin <- is.finite(bal_test)

cat(sprintf("BALIVE == 0:  test %.1f%%   train %.1f%%\n",
            100 * mean(bal_test[fin] == 0), 100 * mean(bal_train[is.finite(bal_train)] == 0)))

edges <- unique(quantile(bal_test[fin & bal_test > 0], probs = seq(0, 1, 0.1), na.rm = TRUE))
nbin  <- length(edges) - 1
assign_strata <- function(x) {
  f <- is.finite(x); s <- rep(NA_integer_, length(x))
  s[f & x == 0] <- 0L
  s[f & x  > 0] <- findInterval(x[f & x > 0], edges, rightmost.closed = TRUE, all.inside = TRUE)
  s
}
strat       <- assign_strata(bal_test)
strat_train <- assign_strata(bal_train)

data.frame(
  stratum = c("zero", paste0("pos_d", 1:nbin)),
  lower   = round(c(0, edges[1:nbin]), 1),
  upper   = round(c(0, edges[2:(nbin + 1)]), 1),
  mean    = round(as.numeric(tapply(bal_test, strat, mean, na.rm = TRUE)), 1),
  n_test  = as.integer(table(factor(strat,       levels = 0:nbin))),
  n_train = as.integer(table(factor(strat_train, levels = 0:nbin))))

op <- par(mar = c(4.4, 4.4, 3, 1))
h <- hist(bal_test[fin & bal_test > 0], breaks = 50, plot = FALSE)
ymax <- max(h$density) * 1.15; p0 <- mean(bal_test[fin] == 0)
plot(h, freq = FALSE, col = "grey88", border = "grey70",
     xlim = c(0, max(bal_test[fin])), ylim = c(0, ymax), xlab = "BALIVE (sq ft/acre)",
     main = sprintf("BALIVE: %.0f%% zeros + positive-value deciles", 100 * p0))
abline(v = edges[2:nbin], col = "steelblue", lwd = 1.3, lty = 2)
arrows(0, 0, 0, ymax * 0.9, col = "firebrick", lwd = 3, length = 0)
text(0, ymax * 0.95, sprintf("zeros %.0f%%", 100 * p0), col = "firebrick", pos = 4, cex = 0.9)
lines(density(bal_train[is.finite(bal_train) & bal_train > 0]), col = "darkorange", lwd = 2)
legend("topright", c("test positive (hist)", "train positive (density)", "decile cut", "zeros"),
       col = c("grey70", "darkorange", "steelblue", "firebrick"),
       lwd = c(6, 2, 1.3, 3), lty = c(1, 1, 2, 1), bty = "n")
par(op)

```

## 7. Forest-type (FORTYPCD) recovery

A good metric for basal area should also keep neighbors within-forest-type
without being told to — the "soft stratification" the embedding is meant to
provide. These reuse the **final BALIVE kNN neighborhoods** (`nn_emb_*`,
`nn_inp_*`) from the k=150 baseline.

### kNN majority-vote classification

```{r fortyp-knn}
knn_mode <- function(nn, label_source, drop_self = FALSE) {
  idx <- if (drop_self) nn$nn.index[, -1, drop = FALSE] else nn$nn.index
  nbr <- matrix(label_source[idx], nrow = nrow(idx))
  apply(nbr, 1, function(row) {
    tab <- table(row)
    if (length(tab) == 0) return(NA_real_)
    as.numeric(names(tab)[which.max(tab)])
  })
}
acc <- function(actual, predicted) mean(actual == predicted, na.rm = TRUE)

ftype_train <- AnnDat_train$FORTYPGRP

fortypcd_results <- data.frame(
  split   = c("train", "val", "test"),
  acc_emb = c(acc(AnnDat_train$FORTYPGRP, knn_mode(nn_emb_train, ftype_train, drop_self = TRUE)),
              acc(AnnDat_val$FORTYPGRP,   knn_mode(nn_emb_val,   ftype_train)),
              acc(AnnDat_test$FORTYPGRP,  knn_mode(nn_emb_test,  ftype_train))),
  acc_inp = c(acc(AnnDat_train$FORTYPGRP, knn_mode(nn_inp_train, ftype_train, drop_self = TRUE)),
              acc(AnnDat_val$FORTYPGRP,   knn_mode(nn_inp_val,   ftype_train)),
              acc(AnnDat_test$FORTYPGRP,  knn_mode(nn_inp_test,  ftype_train)))
)
base_rate <- max(prop.table(table(AnnDat_train$FORTYPGRP)))
cat(sprintf("Majority-class baseline (train): %.3f\n", base_rate))
fortypcd_results
```

### Neighbor forest-type purity

```{r fortyp-purity}
knn_purity <- function(nn, query_labels, ref_labels, drop_self = FALSE) {
  idx <- if (drop_self) nn$nn.index[, -1, drop = FALSE] else nn$nn.index
  nbr <- matrix(ref_labels[idx], nrow = nrow(idx))
  rowMeans(nbr == query_labels, na.rm = TRUE)
}
purity_results <- data.frame(
  split      = c("train", "val", "test"),
  purity_emb = c(mean(knn_purity(nn_emb_train, AnnDat_train$FORTYPGRP, ftype_train, drop_self = TRUE), na.rm = TRUE),
                 mean(knn_purity(nn_emb_val,   AnnDat_val$FORTYPGRP,   ftype_train), na.rm = TRUE),
                 mean(knn_purity(nn_emb_test,  AnnDat_test$FORTYPGRP,  ftype_train), na.rm = TRUE)),
  purity_inp = c(mean(knn_purity(nn_inp_train, AnnDat_train$FORTYPGRP, ftype_train, drop_self = TRUE), na.rm = TRUE),
                 mean(knn_purity(nn_inp_val,   AnnDat_val$FORTYPGRP,   ftype_train), na.rm = TRUE),
                 mean(knn_purity(nn_inp_test,  AnnDat_test$FORTYPGRP,  ftype_train), na.rm = TRUE))
)
rand_purity <- sum(prop.table(table(AnnDat_train$FORTYPGRP))^2)
cat(sprintf("Random-neighbor purity floor: %.3f\n", rand_purity))
purity_results

```

### Multinomial logit on 2-digit forest-type groups

Restricted to plots where the embedding pixel sits exactly at the plot location
(`PLT_CN == ID`), and to type groups with more than 10 plots.

```{r multinom}
form_fortyp_inp_typ <- as.formula(paste("FORTYPGRP ~", paste(inp_type_vars, collapse = " + ")))
form_fortyp_emb_typ <- as.formula(paste("FORTYPGRP ~", paste(emb_type_vars, collapse = " + ")))

reg_data_typ <- reg_data %>% filter(PLT_CN == ID)
forest_types <- table(reg_data_typ$FORTYPGRP)
reg_data_typ <- reg_data_typ %>%
  mutate(FORTYPGRP = if_else(FORTYPGRP %in% names(forest_types[forest_types > 10]), FORTYPGRP, NA),
         FORTYPGRP = as.factor(as.character(FORTYPGRP)))

mod_fortypcd_inp <- multinom(form_fortyp_inp_typ, data = reg_data_typ, maxit = 1000)
pred_inp <- predict(mod_fortypcd_inp, newdata = reg_data_typ)
ok <- !is.na(reg_data_typ$FORTYPGRP) & !is.na(pred_inp)
cat(sprintf("Input multinom accuracy: %.3f\n", mean(pred_inp[ok] == reg_data_typ$FORTYPGRP[ok])))

mod_fortypcd_emb <- multinom(form_fortyp_emb_typ, data = reg_data_typ, maxit = 1000)
pred_emb <- predict(mod_fortypcd_emb, newdata = reg_data_typ)
# NOTE: original script reused the `pred_inp` mask here; corrected to `pred_emb`.
ok <- !is.na(reg_data_typ$FORTYPGRP) & !is.na(pred_emb)
cat(sprintf("Embedding multinom accuracy: %.3f\n", mean(pred_emb[ok] == reg_data_typ$FORTYPGRP[ok])))
```


## Cleanup

```{r cleanup}
dbDisconnect(con)
```

## Interpretation

Across all three probe families — linear, kNN imputation, and forest-type
classification — this report is one sustained head-to-head benchmark of **learned
embeddings vs. raw satellite inputs** on held-out Virginia FIA plots, testing the
project's premise that the embedding is a better *distance metric* for FIA
post-stratification and small-area estimation than the raw features. The recurring
`type`/`phase` block split, separate normalization, and phase-weight tuning mirror
the dual-pathway `z_type`/`z_phase` architecture, and the ysfc-coherence
diagnostics stress-test the phase pathway.

The honest read of the results is more textured than "the embedding wins," and the
figures carry most of the nuance. What follows is grounded in the run shown above;
exact percentages shift with the phase-weight grid and the forest-type granularity
(full `FORTYPCD` vs. 2-digit group), but the qualitative findings are stable.

### What the figures actually show

- **The kNN win is real but modest and scale-broad, not scale-peaked.** The raw
  per-rank homogeneity curve (Section 4, left figure) is too jagged to trust on
  its own — the "peak at rank ~33" rides on that noise. The **banded** version is
  the trustworthy one: after 5-rank smoothing the embedding's neighbors are
  persistently ~1–2 sq ft/acre more BALIVE-homogeneous than the input's across the
  *entire* rank range, most bands clearing ±2 SE. Notably this per-neighbor
  advantage **does not decay at large neighborhoods**, even though the kNN
  *prediction* MSE advantage does shrink with `k` (Section 3). Those are different
  quantities: averaging `k` neighbors converges the prediction to a local mean
  regardless of how tight the individual neighbors are, so homogeneity and
  prediction-MSE need not track.

- **The advantage is localized, and it is a variance win, not a bias win**
  (Section 5). The "Where the embedding wins" panel is *not* a uniform win: the
  embedding is better at the low-BALIVE end, crosses over around mean-BALIVE ≈ 65,
  and is **distinctly worse (≈ +15–25% RMSE) in the mid-productive range**
  (~110–135 sq ft), before recovering at the top. The companion "regression to the
  mean" panel shows the emb and inp residual curves almost superimposed — both
  over-predict low BALIVE and under-predict high BALIVE by nearly the same amount.
  So wherever the embedding wins at the low end, it wins by **lower variance**, not
  by resisting shrinkage. (Read this together with the Section-5 caveats: the zero
  stratum is ~a dozen plots, and the input baseline runs at a hardcoded,
  sub-optimal phase weight.)

- **The recovery signal is real and BALIVE-relevant — this is the strongest
  positive result** (Section 6, distribution-by-ysfc figure). BALIVE rises
  smoothly and monotonically with recovery stage (recent → young → mid → mature →
  stable), and the all-classes overlay is a clean left-to-right march. This is the
  mechanism behind the low-BALIVE win: low-basal-area plots are the
  recently-disturbed ones, exactly what the phase pathway is built to encode.

- **But the phase pathway does *not* yield tighter recovery-stage neighborhoods
  here** (Section 6, ysfc-coherence figure). The two mosaics are visually
  near-identical (both dominated by the ~78% stable block), matching the tied
  purity/lift numbers. The tell is the right-hand barplot: **within-neighborhood
  ysfc spread is *higher* (worse) for the embedding on every disturbed class** —
  recent, young, mid, mature. The embedding gets the coarse recovery class right a
  hair more often, yet its disturbed-plot neighborhoods span a *wider* range of
  recovery stage than the raw inputs. The "z_phase gives coherent recovery
  neighborhoods" story does not hold in this sample.

- **Forest type is a near-tie** (Section 7). The BALIVE-optimal neighborhoods
  respect forest type about as well as the raw inputs do — the embedding is
  marginally ahead on kNN accuracy and purity, marginally behind on the
  multinomial probe. The "soft stratification" is present but not a differentiator.

- **No distribution-shift artifact** (Section 6, distribution + deciles figure):
  the test-positive histogram tracks the train-positive density, so the kNN
  reference set is representative of what it predicts.

### Bottom line

For the stated goal — kNN-style post-stratification and identifying
poorly-sampled / disturbed forest conditions — the embedding delivers a **small,
real, and well-placed improvement**: a few-percent reduction in kNN imputation
error concentrated in the **low / recently-disturbed** end of the BALIVE
distribution, mechanistically tied to the recovery signal the phase pathway
encodes. It is **not** a uniform improvement: it is roughly neutral-to-worse in
the mid-productive range, a variance win rather than a bias win, a near-tie on
forest type, and — despite exp017's recovery-discrimination loss — it does not
produce metrically tighter ysfc neighborhoods than the raw inputs in this
evaluation. Those last two points are the natural targets for the next iteration.
