---
title: "kNN embedding models for FIA attributes"
subtitle: "TPA, hardwood/softwood TPA, basal area, live volume, removals"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  split_col: "split"          # train/val/test column in the embeddings CSV
  k_grid: "5,10,20,40"        # neighbour counts tuned on the validation split
  K_strata: 5                 # strata formed from the kNN prediction (DH bins)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 110)
```

## Overview

kNN regression from the **entire embedding** (`z_type` + `z_phase`, all
dimensions) to six FIA attributes:

| target | source |
|--------|--------|
| **TPA** (all live) | `TREE.TPA_UNADJ`, `STATUSCD=1` |
| **TPA hardwood** | as above, `SPCD >= 300` |
| **TPA softwood** | as above, `SPCD < 300` |
| **basal area** | `COND.BALIVE` |
| **standing live volume** | `TREE.VOLCFNET * TPA_UNADJ`, `STATUSCD=1` |
| **removals** | `TREE_GRM_COMPONENT` annual removed GS volume |

Uses the existing **train / val / test** split: the model (train neighbours) is
fit on train, `k` is tuned on val, and test is scored once. The embedding is
standardized with **train-only** statistics (no leakage). Distances are Euclidean
over the full embedding — note kNN in this many dimensions is subject to distance
concentration, so treat these as a baseline "is the signal linearly *and* locally
present" check against the linear probes elsewhere.

Populations: TPA / basal area / volume are over the full plot population
(non-forest plots contribute 0); removals additionally require GRM/remeasured
data.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(FNN)
library(terra)
```

## Load embeddings and split

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(if_all(all_of(emb_vars), ~ !is.na(.)))

# train/val/test come from the `split` column in the embeddings CSV, the same
# source fia_embedding_evaluation.Rmd uses (reg_data$split)
if (!params$split_col %in% names(emb))
  stop(sprintf("split column '%s' not in embeddings CSV; columns include: %s",
               params$split_col, paste(head(names(emb), 15), collapse = ", ")))
split_df <- emb %>% transmute(CN, split = as.character(.data[[params$split_col]])) %>%
  filter(split %in% c("train", "val", "test")) %>%
  distinct(CN, .keep_all = TRUE)
cat("split counts:\n"); print(table(split_df$split))
```

## FIA targets

```{r fia-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)
dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# basal area (live) per acre + condition status, from COND
cond <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, INVYR, CONDPROP_UNADJ, BALIVE, COND_STATUS_CD FROM COND WHERE STATECD='%s'",
  params$statecd)) %>%
  mutate(PLT_CN = as.character(PLT_CN), BALIVE = ifelse(is.na(BALIVE), 0, BALIVE))
balive <- cond %>% group_by(PLT_CN, INVYR) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")
# dominant-condition status; COND_STATUS_CD == 4 is census water
cstat <- cond %>% group_by(PLT_CN) %>%
  slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup() %>% transmute(PLT_CN, cond_status = COND_STATUS_CD)

# TPA (total / hardwood SPCD>=300 / softwood SPCD<300) and live volume, from TREE
tree_stats <- dbGetQuery(con,
  "SELECT g.PLT_CN,
     SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_total,
     SUM(CASE WHEN g.STATUSCD=1 AND g.SPCD>=300 THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_hard,
     SUM(CASE WHEN g.STATUSCD=1 AND g.SPCD<300  THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_soft,
     SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
   FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

# removals: annual per-acre removed GS volume (GRM; 0 for no-removal plots)
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))
dbDisconnect(con)
```

```{r locs-tcc}
# survey unit + TCC (for the post-stratification baseline)
fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN))
locs <- fia_locs %>% filter(CN %in% emb$CN) %>% distinct(CN, .keep_all = TRUE)
tccr <- rast(params$tcc)
tv   <- extract(tccr, project(vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"),
                                   crs = "EPSG:4269"), crs(tccr)), ID = FALSE)[[1]]
tv   <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
tcc_by_cn <- tibble(CN = locs$CN, tcc = tv)
```

```{r assemble}
analysis <- emb %>% select(CN, all_of(emb_vars)) %>%
  inner_join(split_df, by = "CN") %>%
  left_join(tree_stats, by = c("CN" = "PLT_CN")) %>%
  left_join(balive,     by = c("CN" = "PLT_CN")) %>%
  left_join(cstat,      by = c("CN" = "PLT_CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD), by = "CN") %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  mutate(across(c(tpa_total, tpa_hard, tpa_soft, volume, BALIVE), ~ coalesce(., 0))) %>%
  filter(is.na(cond_status) | cond_status != 4,          # drop census water
         !is.na(UNITCD), !is.na(tcc), tcc >= 0, tcc <= 100,
         INVYR >=2019, INVYR <= 2024)
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # removals: GRM/remeasured
cat(sprintf("plots (stocks): %d   plots (removals): %d   (census water removed)\n",
            nrow(analysis), nrow(ar)))

# train-only standardization of the embedding (fit scaler on train rows)
tr_all <- analysis$split == "train"
mu     <- colMeans(as.matrix(analysis[tr_all, emb_vars]))
sdv    <- apply(as.matrix(analysis[tr_all, emb_vars]), 2, sd); sdv[sdv == 0] <- 1
```

## kNN evaluation

```{r knn-helpers}
kgrid <- as.integer(strsplit(params$k_grid, ",")[[1]])
r2   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)

run_knn <- function(df, yname, label) {
  X <- scale(as.matrix(df[, emb_vars]), center = mu, scale = sdv)
  y <- df[[yname]]
  tr <- which(df$split == "train"); va <- which(df$split == "val"); te <- which(df$split == "test")
  # tune k on validation
  val_r2 <- sapply(kgrid, function(k)
    r2(y[va], FNN::knn.reg(X[tr, , drop = FALSE], X[va, , drop = FALSE], y[tr], k = k)$pred))
  kbest <- kgrid[which.max(val_r2)]
  pte <- FNN::knn.reg(X[tr, , drop = FALSE], X[te, , drop = FALSE], y[tr], k = kbest)$pred
  # linear probe (same train -> test), to separate "embedding decodable" from "kNN-local"
  dtr <- as.data.frame(X[tr, , drop = FALSE]); dtr$.y <- y[tr]
  lm_pred <- as.numeric(predict(lm(.y ~ ., data = dtr), as.data.frame(X[te, , drop = FALSE])))
  list(
    metric = data.frame(target = label, k_best = kbest,
                        n_train = length(tr), n_val = length(va), n_test = length(te),
                        val_R2 = round(max(val_r2), 3), test_R2 = round(r2(y[te], pte), 3),
                        test_R2_lm = round(r2(y[te], lm_pred), 3),
                        test_RMSE = round(sqrt(mean((y[te] - pte)^2)), 2),
                        test_MAE = round(mean(abs(y[te] - pte)), 2)),
    pred = tibble(target = label, obs = y[te], pred = pte, lm_pred = lm_pred,
                  UNITCD = df$UNITCD[te], tcc = df$tcc[te]))
}

targets <- list(
  list(analysis, "tpa_total", "TPA (all live)"),
  list(analysis, "tpa_hard",  "TPA hardwood"),
  list(analysis, "tpa_soft",  "TPA softwood"),
  list(analysis, "BALIVE",    "basal area"),
  list(analysis, "volume",    "standing live volume"),
  list(ar,       "remv_cfgs", "removals"))
res <- lapply(targets, function(t) run_knn(t[[1]], t[[2]], t[[3]]))
```

### Test-set performance

`k_best` is chosen on validation; `test_R2/RMSE/MAE` are the held-out kNN scores.
`test_R2_lm` is a linear probe (trained on train, scored on test) on the same
embedding — the linear reference for kNN.

```{r knn-metrics}
bind_rows(lapply(res, `[[`, "metric"))
```

### Predicted vs observed (test)

```{r knn-scatter, fig.width=11, fig.height=7}
preds <- bind_rows(lapply(res, `[[`, "pred")) %>%
  mutate(target = factor(target, levels = sapply(targets, `[[`, 3)))
ggplot(preds, aes(obs, pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.4) +
  geom_point(alpha = 0.15, size = 0.5) +
  facet_wrap(~ target, scales = "free", ncol = 3) +
  labs(x = "observed", y = "kNN predicted") +
  theme_bw(base_size = 10)
```

## Post-stratification from kNN predictions

Use each kNN prediction to build strata and measure how much it reduces
within-stratum variance of the *actual* target, exactly as in the earlier
post-stratification analyses — but here the strata come from the held-out **test**
predictions (leak-free), and variance is measured on the **test** set. The
**baseline stratification is the TCC split** (FIA operational breaks); the kNN
strata are Dalenius-Hodges bins of the predicted value. Everything is pooled
within survey unit (`UNITCD`), and reported relative to the UNITCD-only variance.

```{r strat-helpers}
pooled_within <- function(df, groups, y) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
dh_breaks <- function(x, K, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(numeric(0))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
}
dh_strata <- function(x, K) {
  bnd <- dh_breaks(x, K); if (length(bnd) == 0) return(factor(rep(1L, length(x))))
  factor(findInterval(as.numeric(x), c(-Inf, bnd, Inf)))
}
# per target: within-var under UNITCD-only, UNITCD x TCC, UNITCD x kNN-prediction DH
strat_summary <- function(pd) {
  base  <- pooled_within(pd, "UNITCD", "obs")
  s_tcc <- cut(pd$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
               labels = c("0-10", "11-48", "49-85", "86-100"))
  s_tccd <- dh_strata(pd$tcc, params$K_strata)   # optimal (DH) 5-strata TCC
  s_lm  <- dh_strata(pd$lm_pred, params$K_strata)
  s_knn <- dh_strata(pd$pred, params$K_strata)
  v_tcc  <- pooled_within(mutate(pd, .s = s_tcc),  c("UNITCD", ".s"), "obs")
  v_tccd <- pooled_within(mutate(pd, .s = s_tccd), c("UNITCD", ".s"), "obs")
  v_lm  <- pooled_within(mutate(pd, .s = s_lm),  c("UNITCD", ".s"), "obs")
  v_knn <- pooled_within(mutate(pd, .s = s_knn), c("UNITCD", ".s"), "obs")
  # RE vs the (fixed-bin) TCC baseline and SRS-equivalent effective sample size.
  # Under a self-weighting sample the estimator variance is within_var / n, so
  # RE = V_TCC / V_method (>1 beats TCC) and ess = n_test * S2 / V_method, with
  # S2 the unstratified (SRS) per-plot variance of the target on the test set.
  S2 <- var(pd$obs); n <- nrow(pd)
  data.frame(target = pd$target[1], n_test = n,
             baseline_sd = round(sqrt(base), 3),
             tcc_pct   = round(100 * v_tcc  / base, 1),
             tccdh_pct = round(100 * v_tccd / base, 1),
             lm_pct  = round(100 * v_lm  / base, 1),
             knn_pct = round(100 * v_knn / base, 1),
             RE_knn_vs_tcc = round(v_tcc / v_knn, 2),
             ess_tcc = round(n * S2 / v_tcc),
             ess_knn = round(n * S2 / v_knn))
}
```

`tcc_pct` / `tccdh_pct` / `lm_pct` / `knn_pct` are within-stratum variance as a
percent of the UNITCD-only baseline (lower = better). `tcc_pct` uses the FIA
operational canopy breaks; `tccdh_pct` uses an optimal (Dalenius-Hodges)
5-strata split of the same TCC covariate — the strongest a single TCC axis can
do at matched strata count. `knn_pct < tccdh_pct` means the
kNN-prediction strata beat even the optimally-binned TCC split; comparing `knn_pct` to `lm_pct` (strata
from a linear probe trained the same way) separates *"the embedding is
decodable"* (both low) from *"kNN's local geometry helps"* (`knn_pct` << `lm_pct`).
`RE_knn_vs_tcc` is the kNN strata's relative efficiency over the fixed-bin TCC
baseline (`V_TCC / V_knn`; > 1 beats TCC). `ess_tcc` / `ess_knn` are
SRS-equivalent **effective sample sizes** (`n_test · S² / within_var`, `S²` the
unstratified target variance) — the SRS plot count matching each estimator's
precision. All three assume a self-weighting sample (`W_h = n_h/N`).

```{r strat-summary}
bind_rows(lapply(res, function(r) strat_summary(r$pred)))
```

Per-stratum detail (kNN-prediction strata) for each target — n, mean, and SD of
the actual target within each stratum:

```{r strat-detail}
bind_rows(lapply(res, function(r) {
  pd <- r$pred
  pd %>% mutate(stratum = as.character(dh_strata(pred, params$K_strata))) %>%
    group_by(target, stratum) %>%
    summarize(n = n(), mean = round(mean(obs), 2), sd = round(sd(obs), 2), .groups = "drop")
})) %>% arrange(target, stratum)
```

### Per-stratum distribution, by method

The tables above give the within-stratum variance (as a percent) and per-stratum
means; this shows the full distribution behind them. For every target (rows) and
every stratification method (columns), a boxplot of the **actual** target within
each stratum. A method separates the target well when its boxes sit at clearly
different levels with little overlap — the visual analog of a low `*_pct`. The
grid shares the y-axis **within each target row** (so the four methods are
directly comparable for a given target) and frees it across rows (targets are on
very different scales); the x-axis is free because strata labels differ across
methods.

```{r strat-boxplot, fig.width=11, fig.height=13}
strata_long <- function(pd) {
  meths <- list(
    "TCC fixed (4)" = cut(pd$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                          labels = c("0-10", "11-48", "49-85", "86-100")),
    "TCC DH (5)"    = dh_strata(pd$tcc,     params$K_strata),
    "LM DH (5)"     = dh_strata(pd$lm_pred, params$K_strata),
    "kNN DH (5)"    = dh_strata(pd$pred,    params$K_strata))
  bind_rows(lapply(names(meths), function(nm)
    tibble(target = pd$target[1], method = nm,
           stratum = as.character(meths[[nm]]), obs = pd$obs)))
}
box_long <- bind_rows(lapply(res, function(r) strata_long(r$pred))) %>%
  mutate(target = factor(target, levels = sapply(targets, `[[`, 3)),
         method = factor(method, levels = c("TCC fixed (4)", "TCC DH (5)",
                                            "LM DH (5)", "kNN DH (5)")))
ggplot(box_long, aes(x = stratum, y = obs)) +
  geom_boxplot(outlier.size = 0.4, fill = "grey85", linewidth = 0.3) +
  facet_grid(target ~ method, scales = "free") +
  labs(x = "stratum", y = "observed target value",
       title = "Per-stratum distribution of the actual target, by method") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Notes

- **Post-stratification is on held-out test data.** Strata are formed from the
  kNN **test** predictions and variance is measured on the test set — no plot is
  binned by its own in-sample prediction. The baseline stratification is the TCC
  split; kNN strata are `K_strata` DH bins of the prediction. Thin `UNITCD x
  stratum` cells on the test set (especially removals) make those numbers noisier
  than the full-population figures elsewhere.
- **Census water removed.** Plots whose dominant condition is census water
  (`COND_STATUS_CD == 4`) are dropped from the whole sample (kNN fit and
  stratification), along with plots missing `UNITCD` or TCC.
- **Split source.** The `train`/`val`/`test` label is the `split` column in the
  embeddings CSV — the same source `fia_embedding_evaluation.Rmd` uses
  (`reg_data$split`). Override with `split_col` if needed; the chunk errors and
  lists available columns if the column is missing.
- **No leakage.** The embedding is standardized with train-only mean/sd, `k` is
  tuned only on validation, and test is scored once with the selected `k`.
- **Hardwood/softwood split** uses the FIA convention `SPCD < 300` = softwood
  (conifer), `SPCD >= 300` = hardwood.
- **Populations.** TPA/basal area/volume use all plots (non-forest = 0); removals
  use the GRM/remeasured subset (`ar`), keeping forested zero-removal plots as 0.
- **Dimensionality caveat.** kNN uses Euclidean distance over the full
  `r length(emb_vars)`-dim embedding; distance concentration can blunt it. Compare
  `test_R2` here against the linear-probe R² in the other scripts — a large gap in
  either direction is informative about whether the signal is linear vs local.
- **Tuning.** `k` is tuned over `k_grid`; widen it (e.g. add 80, 160) if the best
  `k` sits at the edge of the grid.
