---
title: "EVT-cohesive Type Clustering"
subtitle: "z_type clusters that map each EVT to a single cluster (one-to-many EVT->cluster)"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  K: 20            # number of type clusters
  min_evt_n: 20    # min plots for an EVT to be a clustering atom; smaller -> nearest cluster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
```

## Overview

Cluster the **type** embedding `z_type` so the clustering is (as much as possible)
a **one-to-many** map from EVT to cluster: several EVTs may share a cluster, but a
single EVT should not be split across clusters.

The clean way to guarantee that is to cluster at the **EVT level**:

1. Represent each EVT by its **centroid** in standardized `z_type` space.
2. **Ward-cluster the EVT centroids** into `K` groups (reliable EVTs only, i.e.
   `n >= min_evt_n`; rare EVTs are assigned to the nearest cluster afterward).
3. Assign every plot to **its EVT's** cluster.

Because clusters are unions of whole EVTs, every EVT maps to exactly one cluster
**by construction** (clustering *completeness* = 1). EVT is an external vegetation
label, independent of the FIA response variables, so this is supervision that
cannot leak into BALIVE/removals. The interesting question it answers: **does
`z_type` naturally respect EVT boundaries?** — measured by how much the EVT
constraint costs vs a free clustering.

```{r libraries}
library(tidyverse)
```

## Load and per-EVT centroids

```{r load}
emb_type_vars <- paste0("z_type_", 0:47)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(!is.na(evt), evt != -9999)                     # drop nodata EVT

Z    <- scale(as.matrix(emb[, emb_type_vars]))          # standardized z_type
evt  <- as.character(emb$evt)
evt_n <- table(evt)

# per-EVT centroid in standardized z_type
cent  <- rowsum(Z, evt) / as.numeric(evt_n[rownames(rowsum(Z, evt))])
big   <- names(evt_n)[evt_n >= params$min_evt_n]        # reliable centroids
cat(sprintf("plots: %d   EVTs: %d (%d with n >= %d used as clustering atoms)\n",
            nrow(emb), length(evt_n), length(big), params$min_evt_n))
```

## Ward-cluster the EVT centroids

```{r cluster, fig.width=11, fig.height=6}
cent_big <- cent[big, , drop = FALSE]
hc <- hclust(dist(cent_big), method = "ward.D2")
K  <- min(params$K, nrow(cent_big))
cl_big <- cutree(hc, k = K)                              # named by big EVT

# cluster centroids (rows ordered 1..K)
clcent <- t(sapply(sort(unique(cl_big)),
                   function(k) colMeans(cent_big[cl_big == k, , drop = FALSE])))

# assign every EVT: big keep their cut; rare -> nearest cluster centroid
nearest <- function(v) which.min(colSums((t(clcent) - v)^2))
evt_cluster <- vapply(rownames(cent), function(e)
  if (e %in% big) cl_big[[e]] else nearest(cent[e, ]), integer(1))
names(evt_cluster) <- rownames(cent)

emb$type_cluster <- factor(evt_cluster[evt])

# dendrogram of the big EVTs, boxed by cluster
op <- par(mar = c(6, 4, 3, 1))
plot(hc, labels = big, xlab = "EVT", sub = "", cex = 0.6,
     main = sprintf("EVT centroids in z_type -> %d clusters (Ward)", K))
rect.hclust(hc, k = K, border = 2:(K + 1))
par(op)
```

## Verify the one-to-many property

```{r verify}
tab <- table(evt = evt, cluster = emb$type_cluster)
per_evt_clusters <- rowSums(tab > 0)                    # clusters each EVT appears in
cat(sprintf("EVTs mapping to exactly one cluster: %d / %d (%.1f%%)\n",
            sum(per_evt_clusters == 1), length(per_evt_clusters),
            100 * mean(per_evt_clusters == 1)))
stopifnot(all(per_evt_clusters == 1))                   # true by construction
```

## Cluster composition

Each cluster aggregates several EVTs (expected). `n_evts` = distinct EVTs pooled
into the cluster; `top_evts` = the largest few by plot count.

```{r composition}
emb %>%
  group_by(type_cluster) %>%
  summarize(n_plots = n(),
            n_evts  = dplyr::n_distinct(evt),
            top_evts = paste(names(sort(table(evt), decreasing = TRUE))[1:3], collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_plots))
```

## Does z_type respect EVT? (cost of the constraint)

Compare within-cluster `z_type` dispersion of the EVT-cohesive clustering to a
**free** k-means at the same `K` (which minimizes it). Ratio near 1 means the
embedding already separates EVTs cleanly, so forcing EVT-cohesion costs almost
nothing; a large ratio means the constraint fights the embedding.

```{r compactness}
tot_within <- function(cl) {
  ss <- sum(vapply(split(seq_len(nrow(Z)), cl), function(ix) {
    m <- Z[ix, , drop = FALSE]; sum(sweep(m, 2, colMeans(m))^2)
  }, numeric(1)))
  ss / nrow(Z)
}
set.seed(1)
km <- kmeans(Z, centers = K, nstart = 10)

w_evt <- tot_within(emb$type_cluster)
w_km  <- tot_within(km$cluster)
data.frame(
  clustering = c("EVT-cohesive (this)", "free k-means (same K)"),
  within_var = round(c(w_evt, w_km), 3),
  ratio_to_kmeans = round(c(w_evt / w_km, 1), 3)
)

# homogeneity: how EVT-mixed are the clusters (low is expected -- many EVTs/cluster)
homogeneity <- 1 - sum(apply(tab, 2, function(col) {
  p <- col[col > 0] / sum(col); -sum(p * log(p))
}) * colSums(tab)) / sum(tab) /
  ( -sum((rowSums(tab)/sum(tab)) * log(rowSums(tab)/sum(tab))) )
cat(sprintf("completeness (EVT -> one cluster): 1.000 (by construction)\n"))
cat(sprintf("homogeneity  (cluster -> one EVT): %.3f (low expected; clusters pool EVTs)\n",
            homogeneity))
```

## EVT -> cluster map

The full assignment (join `evt` -> `type_cluster` onto any plot table to use these
as type strata).

```{r map}
evt_map <- tibble(evt = as.integer(names(evt_cluster)),
                  n_plots = as.integer(evt_n[names(evt_cluster)]),
                  type_cluster = as.integer(evt_cluster),
                  atom = names(evt_cluster) %in% big) %>%
  arrange(type_cluster, desc(n_plots))
evt_map
```

## Notes

- **Guaranteed one-to-many.** Clustering at the EVT level makes completeness = 1
  exactly; `homogeneity` is informational (clusters intentionally pool EVTs).
- **The compactness ratio is the headline diagnostic.** Near 1.0 = `z_type`
  already encodes EVT structure (the model learned vegetation type), so
  EVT-cohesive strata are also embedding-compact. Far above 1 = EVTs are smeared
  across the embedding and the constraint is expensive.
- **Rare EVTs** (`n < min_evt_n`) are assigned to the nearest cluster centroid, so
  they still map one-to-one to a cluster; raise `min_evt_n` to treat more of them
  as noise-assigned rather than clustering atoms.
- **Choosing K.** Read it off the dendrogram; `K` is a param. Unweighted centroids
  give rare and common EVTs equal say in the grouping -- weight `cent_big` rows by
  `sqrt(n)` if you want common EVTs to dominate the merges.
- **Using it downstream.** `emb$type_cluster` (or `evt_map`) is a drop-in
  replacement for the k-means/Ward `type_bin` step in the hierarchical
  stratification -- an EVT-supervised type binning that never sees the FIA target.
