---
title: "EVT-trained Type Segmentation"
subtitle: "z_type segments trained on EVT, plots assigned by embedding location"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  evt_table: "~/Dropbox/git/vq-vae/data/LF2024_EVT.csv"
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  K: 20            # number of type clusters / segments (both methods)
  min_evt_n: 20    # min plots for an EVT centroid to seed clustering (both methods)
  phase_subdivide_frac: 0.25  # [original] fraction of type clusters to phase-subdivide
  k_phase: 3       # phase sub-groups per subdivided cluster (both methods)
  K_strata: 5      # [original] final strata; also the TCC/flat competitors
  tree_phase_frac: 0.20       # [constrained] top fraction of TREE segments to subdivide
  K_strata_tree: 3            # [constrained] Tree strata
  K_strata_nontree: 2         # [constrained] non-Tree strata (total = tree + nontree)
  K_full: 50                  # [full-embedding] k-means clusters over z_type + z_phase
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
```

## Overview

**Segment** the type embedding `z_type` using EVT as supervision, then assign each
plot to a segment by **its own location in `z_type`** — not by its EVT label.

1. Represent each EVT by its **centroid** in standardized `z_type` (reliable EVTs,
   `n >= min_evt_n`; rare EVTs attach to the nearest).
2. **Ward-cluster the EVT centroids** into `K` groups — EVT trains *which regions
   of the embedding form*.
3. Compute each group's centroid from its plots and **assign every plot to the
   nearest group centroid** — a Voronoi segmentation of `z_type` space.

So the segment boundaries are learned from EVT, but plots land in bins by *where
they sit* in the embedding; an EVT may split across bins if its plots are
embedding-diverse. EVT is external to the FIA responses, so this supervision
cannot leak into BALIVE/volume/removals. Diagnostics report how cohesive EVTs
remain and how compact the segmentation is vs a free k-means.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Load and per-EVT centroids

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(!is.na(evt), evt != -9999)                     # drop nodata EVT

Z    <- scale(as.matrix(emb[, emb_type_vars]))          # standardized z_type
evt  <- as.character(emb$evt)
evt_n <- table(evt)

# per-EVT centroid in standardized z_type
cent  <- rowsum(Z, evt) / as.numeric(evt_n[rownames(rowsum(Z, evt))])
big   <- names(evt_n)[evt_n >= params$min_evt_n]        # reliable centroids
cat(sprintf("plots: %d   EVTs: %d (%d with n >= %d used as clustering atoms)\n",
            nrow(emb), length(evt_n), length(big), params$min_evt_n))
```

## Ward-cluster the EVT centroids

```{r cluster, fig.width=11, fig.height=6}
cent_big <- cent[big, , drop = FALSE]
hc <- hclust(dist(cent_big), method = "ward.D2")
K  <- min(params$K, nrow(cent_big))
cl_big <- cutree(hc, k = K)                              # named by big EVT

# cluster centroids (rows ordered 1..K)
clcent <- t(sapply(sort(unique(cl_big)),
                   function(k) colMeans(cent_big[cl_big == k, , drop = FALSE])))

# EVT -> group map (the EVT-trained grouping): big EVTs from the Ward cut, rare
# EVTs attached to the nearest EVT-centroid cluster.
nearest_c <- function(v) which.min(colSums((t(clcent) - v)^2))
evt_group <- vapply(rownames(cent), function(e)
  if (e %in% big) cl_big[[e]] else nearest_c(cent[e, ]), integer(1))
names(evt_group) <- rownames(cent)

# Segment z_type: group centroids computed from the plots (seeded by the EVT
# grouping), then assign EVERY PLOT to the nearest group centroid by its own
# z_type location. An EVT may split across segments.
grp0 <- evt_group[evt]
glev <- sort(unique(grp0))
gc   <- t(sapply(glev, function(g) colMeans(Z[grp0 == g, , drop = FALSE])))
d2   <- sapply(seq_len(nrow(gc)), function(k) rowSums(sweep(Z, 2, gc[k, ], "-")^2))
emb$type_cluster <- factor(glev[max.col(-d2, ties.method = "first")])

# dendrogram of the big EVTs, boxed by cluster
op <- par(mar = c(6, 4, 3, 1))
plot(hc, labels = big, xlab = "EVT", sub = "", cex = 0.6,
     main = sprintf("EVT centroids in z_type -> %d clusters (Ward)", K))
rect.hclust(hc, k = K, border = 2:(K + 1))
par(op)
```

## EVT cohesion under location-based assignment

How much do EVTs stay together now that plots are assigned by embedding location?
(High = `z_type` respects EVT; an EVT splits only when its plots are
embedding-diverse.)

```{r verify}
tab <- table(evt = evt, cluster = emb$type_cluster)
cat(sprintf("EVT cohesion (plot-weighted: plots in their EVT's modal segment): %.3f\n",
            sum(apply(tab, 1, max)) / sum(tab)))
cat(sprintf("EVTs entirely within one segment: %d / %d (%.1f%%)\n",
            sum(rowSums(tab > 0) == 1), nrow(tab), 100 * mean(rowSums(tab > 0) == 1)))
```

## Cluster composition

Each cluster aggregates several EVTs (expected). `n_evts` = distinct EVTs pooled
into the cluster; `top_evts` = the largest few by plot count.

```{r composition}
emb %>%
  group_by(type_cluster) %>%
  summarize(n_plots = n(),
            n_evts  = dplyr::n_distinct(evt),
            top_evts = paste(names(sort(table(evt), decreasing = TRUE))[1:3], collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_plots))
```

## Does z_type respect EVT? (segmentation compactness)

Compare within-segment `z_type` dispersion of the EVT-trained segmentation to a
**free** k-means at the same `K` (which minimizes it). Ratio near 1 means the
EVT-trained segments are almost as tight as unconstrained clusters — i.e. `z_type`
already separates EVTs cleanly; a large ratio means the EVT supervision pulls the
segments away from the embedding's natural structure.

```{r compactness}
tot_within <- function(cl) {
  ss <- sum(vapply(split(seq_len(nrow(Z)), cl), function(ix) {
    m <- Z[ix, , drop = FALSE]; sum(sweep(m, 2, colMeans(m))^2)
  }, numeric(1)))
  ss / nrow(Z)
}
set.seed(1)
km <- kmeans(Z, centers = K, nstart = 10)

w_evt <- tot_within(emb$type_cluster)
w_km  <- tot_within(km$cluster)
data.frame(
  clustering = c("EVT-trained segmentation (this)", "free k-means (same K)"),
  within_var = round(c(w_evt, w_km), 3),
  ratio_to_kmeans = round(c(w_evt / w_km, 1), 3)
)

# homogeneity: how EVT-coherent the segments are (higher = each segment ~ few EVTs)
homogeneity <- 1 - sum(apply(tab, 2, function(col) {
  p <- col[col > 0] / sum(col); -sum(p * log(p))
}) * colSums(tab)) / sum(tab) /
  ( -sum((rowSums(tab)/sum(tab)) * log(rowSums(tab)/sum(tab))) )
cat(sprintf("homogeneity (segment -> one EVT): %.3f (higher = segments are EVT-coherent)\n",
            homogeneity))
```

## EVT -> segment map

Each EVT's **modal** segment and its **cohesion** (fraction of the EVT's plots in
that segment). Cohesion < 1 means the EVT split across segments under
location-based assignment.

```{r map}
# LANDFIRE EVT lookup: VALUE (the EVT code) -> EVT_NAME
evt_lookup <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  select(VALUE, EVT_NAME)

evt_map <- emb %>%
  group_by(VALUE = evt) %>%
  summarize(n_plots  = n(),
            cluster  = as.integer(names(sort(table(type_cluster), decreasing = TRUE))[1]),
            cohesion = round(max(table(type_cluster)) / n(), 3),
            .groups  = "drop") %>%
  mutate(VALUE = as.integer(VALUE)) %>%
  left_join(evt_lookup, by = "VALUE") %>%
  select(VALUE, EVT_NAME, n_plots, type_cluster = cluster, cohesion) %>%
  arrange(type_cluster, desc(n_plots))
evt_map
```

## Stratification: BALIVE, Volume, and Removals vs TCC

Build the fine bins as **EVT type clusters, then phase-subdivided**: take the
`r params$K` EVT-trained type segments, and split the top
`r round(100*params$phase_subdivide_frac)`% (by within-cluster `z_phase` spread)
into `r params$k_phase` phase sub-groups each (Ward on `z_phase`) — the same
type->phase refinement as the geometry hierarchical script, but with EVT-based
type bins. Then aggregate the fine bins into `K_strata` strata by Ward on the
cluster-mean response (supervised), and compare within-stratum SD against TCC and
a flat type+phase probe, by survey unit. Responses: **BALIVE**, **standing live
volume**, and **removals**. Stocks (BALIVE/volume) run on the **full plot
population** (non-forest plots contribute 0 — no forest-code filter); removals
additionally require remeasurement/GRM data, which naturally scopes them to
treed plots while still retaining land-use diversions.

```{r strat-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

cond <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, CONDPROP_UNADJ, BALIVE, FORTYPCD FROM COND WHERE STATECD='%s'",
  params$statecd)) %>% mutate(PLT_CN = as.character(PLT_CN))
balive <- cond %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")
fortyp <- cond %>%
  group_by(PLT_CN) %>% slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup() %>% select(PLT_CN, FORTYPCD)

dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# standing LIVE net cu ft volume per acre (STATUSCD=1)
vol <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
     FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

# annual per-acre removed GS volume; COALESCE inside SUM so no-removal plots are 0
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN))
```

```{r strat-tcc}
locs <- fia_locs %>% filter(CN %in% emb$CN) %>% distinct(CN, .keep_all = TRUE)
tccr <- rast(params$tcc)
pts  <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
tv   <- extract(tccr, project(pts, crs(tccr)), ID = FALSE)[[1]]
tv   <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
tcc_by_cn <- tibble(CN = locs$CN, tcc = tv)
```

```{r strat-assemble}
analysis <- emb %>%
  select(CN, type_cluster, all_of(emb_type_vars), all_of(emb_phase_vars)) %>%
  inner_join(balive, by = c("CN" = "PLT_CN")) %>%
  left_join(vol,     by = c("CN" = "PLT_CN")) %>%
  inner_join(fortyp, by = c("CN" = "PLT_CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD), by = "CN") %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  mutate(volume = coalesce(volume, 0)) %>%
  filter(!is.na(UNITCD), !is.na(tcc), tcc >= 0, tcc <= 100)   # full population; no forest-code filter
cat(sprintf("plots with covariates: %d across %d units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
```

```{r strat-helpers}
pooled_within <- function(df, groups, y) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
# Dalenius-Hodges cumulative-sqrt(f) boundaries (the K-1 interior breaks)
dh_breaks <- function(x, K, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(numeric(0))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
}
dh_strata <- function(x, K, nbins = 100) {
  bnd <- dh_breaks(x, K, nbins)
  if (length(bnd) == 0) return(factor(rep(1L, length(x))))
  factor(findInterval(as.numeric(x), c(-Inf, bnd, Inf)))
}
ward <- function(X, k) {
  d  <- dist(X)
  hc <- if (requireNamespace("fastcluster", quietly = TRUE))
          fastcluster::hclust(d, method = "ward.D2") else hclust(d, method = "ward.D2")
  cutree(hc, k = k)
}
# fine bins -> strata: Ward on the fine-bin-mean response (supervised aggregation)
bin_map <- function(bin, y, K) {
  d  <- tibble(fine_bin = bin, .y = y) %>% group_by(fine_bin) %>%
        summarize(n_plots = n(), mean_y = mean(.y), .groups = "drop")
  Kk <- min(K, nrow(d))
  d$stratum <- if (nrow(d) <= 1) 1L else
    cutree(hclust(dist(scale(d$mean_y)), method = "ward.D2"), Kk)
  d
}
agg_by <- function(bin, y, K) {
  d <- bin_map(bin, y, K)
  factor(d$stratum[match(bin, d$fine_bin)])
}
# fine-bin -> stratum table for display
show_map <- function(df, y) {
  bin_map(df$fine_bin, df[[y]], params$K_strata) %>%
    transmute(fine_bin, n_plots, mean_response = round(mean_y, 2), stratum) %>%
    arrange(stratum, desc(mean_response))
}
# flat OOF supervised probe on `vars` -> DH bins (the direct-supervision ceiling)
oof_dh <- function(df, y, vars, K) {
  set.seed(1); folds <- sample(rep(1:5, length.out = nrow(df)))
  d <- df %>% mutate(y_resp = .data[[y]])
  form <- reformulate(vars, "y_resp"); s <- numeric(nrow(df))
  for (f in 1:5) { fit <- lm(form, d[folds != f, ]); s[folds == f] <- predict(fit, d[folds == f, ]) }
  dh_strata(s, K)
}
# response-by-stratum boxplots for a named list of stratifications (one facet each)
strata_boxplot <- function(df, y, schemes, ylab = y, logy = FALSE) {
  long <- dplyr::bind_rows(lapply(names(schemes), function(nm)
    tibble(scheme = nm, stratum = as.character(schemes[[nm]]), val = df[[y]])))
  long$scheme <- factor(long$scheme, levels = names(schemes))
  long$yval   <- if (logy) log1p(long$val) else long$val
  ggplot(long, aes(x = stratum, y = yval)) +
    geom_boxplot(outlier.size = 0.4, fill = "grey85", linewidth = 0.3) +
    facet_wrap(~ scheme, scales = "free_x", nrow = 1) +
    labs(x = "stratum", y = if (logy) paste0("log1p(", ylab, ")") else ylab) +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
# per-stratum n / mean / sd for a named list of stratifications
strata_summary <- function(df, y, schemes) {
  dplyr::bind_rows(lapply(names(schemes), function(nm)
    tibble(scheme = nm, stratum = as.character(schemes[[nm]]), val = df[[y]]) %>%
      group_by(scheme, stratum) %>%
      summarize(n = dplyr::n(), mean = round(mean(val), 3), sd = round(sd(val), 3),
                .groups = "drop"))) %>%
    mutate(scheme = factor(scheme, levels = names(schemes))) %>%
    arrange(scheme, stratum)
}
# the four stratifications compared for the ORIGINAL (location-based) method
schemes_o <- function(df, y) list(
  "TCC FIA fixed"  = cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                         labels = c("0-10", "11-48", "49-85", "86-100")),
  "TCC DH"         = dh_strata(df$tcc, params$K_strata),
  "EVT type/phase" = agg_by(df$fine_bin, df[[y]], params$K_strata),
  "flat probe"     = oof_dh(df, y, emb_vars, params$K_strata))
```

### Fine bins: EVT type clusters -> phase subdivision

```{r fine-bins}
# within-cluster z_phase spread; subdivide the top fraction into k_phase groups
Zp   <- scale(as.matrix(analysis[, emb_phase_vars]))
pvar <- tapply(seq_len(nrow(analysis)), analysis$type_cluster,
               function(ix) sum(apply(Zp[ix, , drop = FALSE], 2, var)))
n_cl   <- length(pvar)
n_sub  <- max(1, round(params$phase_subdivide_frac * n_cl))
sub_cl <- names(sort(pvar, decreasing = TRUE))[seq_len(n_sub)]

analysis$phase_sub <- 1L
for (b in sub_cl) {
  ix <- which(as.character(analysis$type_cluster) == b)
  if (length(ix) >= 10 * params$k_phase)
    analysis$phase_sub[ix] <- ward(Zp[ix, , drop = FALSE], params$k_phase)
}
analysis$fine_bin <- interaction(analysis$type_cluster, analysis$phase_sub, drop = TRUE)
cat(sprintf("fine bins: %d (from %d type clusters, %d phase-subdivided)\n",
            nlevels(analysis$fine_bin), n_cl, length(sub_cl)))

# type clusters ranked by within-cluster z_phase variance, flagged if subdivided
tibble(type_cluster = names(pvar),
       n            = as.integer(table(analysis$type_cluster)[names(pvar)]),
       phase_var    = round(as.numeric(pvar), 3),
       subdivided   = names(pvar) %in% sub_cl) %>%
  arrange(desc(phase_var))
```

### Removals population (remeasured + GRM data)

```{r removals-pop}
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # GRM/remeasured; zeros kept
cat(sprintf("remeasured plots with GRM data: %d (%.1f%% with any removal)\n",
            nrow(ar), 100 * mean(ar$remv_cfgs > 0)))
```

### Response distributions by top EVT classes

Histograms of each response within the 10 most common EVT forest types (by plot
count over the full population). Shows how each target is distributed within a
forest type — and how much of the total spread is *within* a type, which caps how
much a type-based stratification can reduce variance.

```{r evt-hist-setup}
evt_names <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  transmute(evt = as.character(VALUE), EVT_NAME)
emb_evt <- emb %>% transmute(CN, evt = as.character(evt))

top10 <- analysis %>% inner_join(emb_evt, by = "CN") %>%
  count(evt, sort = TRUE) %>% slice_head(n = 10) %>% pull(evt)

lab_of <- function(e) {
  nm <- evt_names$EVT_NAME[match(e, evt_names$evt)]
  ifelse(is.na(nm), e, sprintf("%s (%s)", substr(nm, 1, 26), e))
}
hist_by_evt <- function(df, y, ylab, logx = FALSE) {
  d <- df %>% inner_join(emb_evt, by = "CN") %>% filter(evt %in% top10) %>%
    mutate(evt_lab = factor(lab_of(evt), levels = lab_of(top10)),
           val = if (logx) log1p(.data[[y]]) else .data[[y]])
  ggplot(d, aes(val)) +
    geom_histogram(bins = 30, fill = "grey40") +
    facet_wrap(~ evt_lab, scales = "free_y", ncol = 2) +
    labs(x = if (logx) paste0("log1p(", ylab, ")") else ylab, y = "plots") +
    theme_bw(base_size = 9)
}
```

```{r evt-hist-balive, fig.width=11, fig.height=9}
hist_by_evt(analysis, "BALIVE", "BALIVE")
```

```{r evt-hist-volume, fig.width=11, fig.height=9}
hist_by_evt(analysis, "volume", "volume")
```

```{r evt-hist-removals, fig.width=11, fig.height=9}
hist_by_evt(ar, "remv_cfgs", "removals", logx = TRUE)
```

### Response statistics by EVT class

Per-EVT mean, SD, 75th percentile, and fraction of zero-valued plots for each
response. BALIVE and volume are over the full population (`n`); removals are over
the remeasured/GRM subset (`rem_n`). `*_fz` is the zero fraction.

```{r evt-stats}
av <- analysis %>% inner_join(emb_evt, by = "CN") %>% group_by(evt) %>%
  summarize(n = n(),
            bal_mean = round(mean(BALIVE), 1), bal_sd = round(sd(BALIVE), 1),
            bal_q75 = round(quantile(BALIVE, .75), 1), bal_fz = round(mean(BALIVE == 0), 3),
            vol_mean = round(mean(volume), 0), vol_sd = round(sd(volume), 0),
            vol_q75 = round(quantile(volume, .75), 0), vol_fz = round(mean(volume == 0), 3),
            .groups = "drop")
rm_stats <- ar %>% inner_join(emb_evt, by = "CN") %>% group_by(evt) %>%
  summarize(rem_n = n(), rem_mean = round(mean(remv_cfgs), 1), rem_sd = round(sd(remv_cfgs), 1),
            rem_q75 = round(quantile(remv_cfgs, .75), 1), rem_fz = round(mean(remv_cfgs == 0), 3),
            .groups = "drop")

av %>% left_join(rm_stats, by = "evt") %>%
  mutate(VALUE = as.integer(evt)) %>%
  left_join(transmute(evt_names, VALUE = as.integer(evt), EVT_NAME), by = "VALUE") %>%
  select(VALUE, EVT_NAME, n,
         bal_mean, bal_sd, bal_q75, bal_fz,
         vol_mean, vol_sd, vol_q75, vol_fz,
         rem_n, rem_mean, rem_sd, rem_q75, rem_fz) %>%
  arrange(desc(n))
```

### Evaluation

```{r strat-eval}
# TCC break points actually used by the two TCC schemes
cat("FIA fixed TCC breaks:", "10, 48, 85", "(0-10 / 11-48 / 49-85 / 86-100)\n")
cat(sprintf("TCC DH (%d) breaks: %s\n", params$K_strata,
            paste(round(dh_breaks(analysis$tcc, params$K_strata), 1), collapse = ", ")))

resp_table <- function(df, y, vars = emb_vars) {
  s_fia  <- cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                labels = c("0-10", "11-48", "49-85", "86-100"))     # FIA operational
  s_tcc  <- dh_strata(df$tcc, params$K_strata)
  s_evt  <- agg_by(df$fine_bin, df[[y]], params$K_strata)
  s_flat <- oof_dh(df, y, vars, params$K_strata)
  vv <- c(pooled_within(df, "UNITCD", y),
          pooled_within(mutate(df, .s = s_fia),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_tcc),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_evt),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_flat), c("UNITCD", ".s"), y))
  data.frame(
    stratification  = c("UNITCD only (baseline)",
                        "UNITCD x TCC FIA fixed (0-10/11-48/49-85/86-100)",
                        sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                        sprintf("UNITCD x EVT type/phase hierarchical (%d)", params$K_strata),
                        sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var      = round(vv, 2),
    within_sd       = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1))
}
```

#### BALIVE

```{r strat-balive}
resp_table(analysis, "BALIVE")
```

Fine bin -> stratum mapping (BALIVE):

```{r strat-balive-map}
show_map(analysis, "BALIVE")
```

Strata sizes, mean, and SD (BALIVE):

```{r strat-balive-strata}
sch_bal <- schemes_o(analysis, "BALIVE")
strata_summary(analysis, "BALIVE", sch_bal)
```

Distribution of BALIVE within each stratum:

```{r strat-balive-box, fig.width=11, fig.height=4}
strata_boxplot(analysis, "BALIVE", sch_bal, ylab = "BALIVE")
```

#### Standing live volume

```{r strat-volume}
resp_table(analysis, "volume")
```

Fine bin -> stratum mapping (volume):

```{r strat-volume-map}
show_map(analysis, "volume")
```

Strata sizes, mean, and SD (volume):

```{r strat-volume-strata}
sch_vol <- schemes_o(analysis, "volume")
strata_summary(analysis, "volume", sch_vol)
```

Distribution of volume within each stratum:

```{r strat-volume-box, fig.width=11, fig.height=4}
strata_boxplot(analysis, "volume", sch_vol, ylab = "volume")
```

#### Removals

```{r strat-removals}
resp_table(ar, "remv_cfgs")
```

Fine bin -> stratum mapping (removals):

```{r strat-removals-map}
show_map(ar, "remv_cfgs")
```

Strata sizes, mean, and SD (removals):

```{r strat-removals-strata}
sch_rmv <- schemes_o(ar, "remv_cfgs")
strata_summary(ar, "remv_cfgs", sch_rmv)
```

Removals by stratum for each scheme (`log1p` axis — removals are zero-inflated,
so a scheme separates well when some strata lift off the zero floor):

```{r strat-removals-box, fig.width=11, fig.height=4}
strata_boxplot(ar, "remv_cfgs", sch_rmv, ylab = "removals", logy = TRUE)
```

```{r strat-cleanup}
dbDisconnect(con)
```

# Constrained-Ward Tree / non-Tree typology

*(A second, self-contained method — the original method above is untouched.)*

A domain-directed variant that refuses to mix LANDFIRE macro life-forms. Using
`EVT_LF`, every EVT is either **Tree** or **non-Tree**. We segment `z_type` with
Ward clustering under a **cannot-link constraint** (a Tree centroid may never
merge with a non-Tree centroid), so each of the `r params$K` segments is
macro-pure by construction. As before, the final plot assignment is by **nearest
segment centroid in `z_type` alone** — the macro side is never used to place a
plot, so we can measure *leakage*: plots that land embedding-nearest to a segment
whose macro differs from the plot's own EVT macro. Downstream, only Tree segments
are phase-subdivided, and the strata keep the split: `r params$K_strata_tree`
Tree strata + `r params$K_strata_nontree` non-Tree strata.

## Cannot-link Ward on EVT centroids

```{r c-load-lf}
evt_lf <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  transmute(VALUE = as.character(VALUE), EVT_LF)
lf_by_evt <- setNames(evt_lf$EVT_LF, evt_lf$VALUE)

is_tree_evt <- setNames(lf_by_evt[names(evt_n)] == "Tree", names(evt_n))
is_tree_evt[is.na(is_tree_evt)] <- FALSE                 # unknown LF -> non-Tree
cat(sprintf("EVTs: %d Tree / %d non-Tree (by EVT_LF); big EVTs: %d Tree / %d non-Tree\n",
            sum(is_tree_evt), sum(!is_tree_evt),
            sum(is_tree_evt[big]), sum(!is_tree_evt[big])))
```

```{r c-cluster, fig.width=11, fig.height=6}
cent_big <- cent[big, , drop = FALSE]
tree_big <- is_tree_evt[big]                             # named logical, big EVTs

# cannot-link: inflate every cross-macro distance so Ward defers those merges
D <- as.matrix(dist(cent_big))
D[outer(tree_big, tree_big, `!=`)] <- max(D) * 1000
hc_c   <- hclust(as.dist(D), method = "ward.D2")
K      <- min(params$K, nrow(cent_big))
cl_big_c <- cutree(hc_c, k = K)                          # named by big EVT

seg_ids   <- sort(unique(cl_big_c))
clcent_c  <- t(sapply(seg_ids, function(k)
  colMeans(cent_big[cl_big_c == k, , drop = FALSE])))
# segment macro = majority macro of its big EVTs (pure if the constraint held)
seg_macro0 <- vapply(seg_ids, function(k) mean(tree_big[cl_big_c == k]) >= 0.5, logical(1))

# dendrogram, boxed by segment
op <- par(mar = c(6, 4, 3, 1))
plot(hc_c, labels = ifelse(tree_big, paste0(big, "*"), big), xlab = "EVT (* = Tree)",
     sub = "", cex = 0.6,
     main = sprintf("Cannot-link Ward: EVT centroids -> %d macro-pure segments", K))
rect.hclust(hc_c, k = K, border = 2:(K + 1))
par(op)
```

```{r c-assign}
# rare EVTs seed the grouping via nearest SAME-MACRO segment centroid
nearest_c_macro <- function(v, want_tree) {
  cand <- which(seg_macro0 == want_tree)
  cand[which.min(colSums((t(clcent_c[cand, , drop = FALSE]) - v)^2))]
}
evt_group_c <- vapply(rownames(cent), function(e) {
  if (e %in% big) return(match(cl_big_c[[e]], seg_ids))
  nearest_c_macro(cent[e, ], isTRUE(is_tree_evt[e]))
}, integer(1))
names(evt_group_c) <- rownames(cent)

# segment centroids from the plots; assign EVERY plot by nearest centroid in
# z_type ONLY (macro side never used to place a plot)
grp0_c <- evt_group_c[evt]
glev_c <- sort(unique(grp0_c))
gc_c   <- t(sapply(glev_c, function(g) colMeans(Z[grp0_c == g, , drop = FALSE])))
d2_c   <- sapply(seq_len(nrow(gc_c)), function(k) rowSums(sweep(Z, 2, gc_c[k, ], "-")^2))
seg_pos <- glev_c[max.col(-d2_c, ties.method = "first")]
emb$type_cluster_c <- factor(seg_pos)
emb$macro_c <- ifelse(seg_macro0[seg_pos], "Tree", "nonTree")
```

## Leakage under embedding-only assignment

Segments are macro-pure by construction; leakage is what the *location-based*
assignment costs — plots whose nearest `z_type` segment belongs to the other
macro than their own EVT. Low leakage means the Tree / non-Tree boundary is
clean in the embedding.

```{r c-leakage}
plot_evt_macro <- ifelse(is_tree_evt[evt], "Tree", "nonTree")
cat(sprintf("segments: %d Tree / %d non-Tree\n", sum(seg_macro0), sum(!seg_macro0)))
cat(sprintf("plot macro leakage (assigned macro != EVT macro): %.2f%%\n",
            100 * mean(emb$macro_c != plot_evt_macro)))
cat(sprintf("  of Tree-EVT plots landing in a non-Tree segment: %.2f%%\n",
            100 * mean(emb$macro_c[plot_evt_macro == "Tree"] == "nonTree")))
cat(sprintf("  of non-Tree-EVT plots landing in a Tree segment: %.2f%%\n",
            100 * mean(emb$macro_c[plot_evt_macro == "nonTree"] == "Tree")))

# per-segment purity of the seeding EVTs (cannot-link check)
tibble(segment = seg_ids,
       macro   = ifelse(seg_macro0, "Tree", "nonTree"),
       n_big_evts = as.integer(table(cl_big_c)[as.character(seg_ids)]),
       n_tree_evts    = as.integer(tapply(tree_big, cl_big_c, sum)[as.character(seg_ids)]),
       n_nontree_evts = as.integer(tapply(!tree_big, cl_big_c, sum)[as.character(seg_ids)])) %>%
  mutate(segment = match(segment, glev_c),
         n_plots = as.integer(table(factor(seg_pos, levels = glev_c))[as.character(segment)])) %>%
  arrange(macro, desc(n_plots))
```

## EVT -> segment map (constrained)

Each EVT's **modal** segment under embedding-only assignment, its macro side,
cohesion (fraction of the EVT's plots in that segment), and a `leaks` flag —
`TRUE` when the EVT's plots land, on balance, in a segment of the *other* macro
than its own `EVT_LF`.

```{r c-map}
evt_lookup2 <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  select(VALUE, EVT_NAME, EVT_LF) %>% mutate(VALUE = as.integer(VALUE))

evt_map_c <- emb %>%
  group_by(VALUE = evt) %>%
  summarize(n_plots  = n(),
            segment  = as.integer(names(sort(table(type_cluster_c), decreasing = TRUE))[1]),
            macro    = names(sort(table(macro_c), decreasing = TRUE))[1],
            cohesion = round(max(table(type_cluster_c)) / n(), 3),
            .groups  = "drop") %>%
  mutate(VALUE = as.integer(VALUE)) %>%
  left_join(evt_lookup2, by = "VALUE") %>%
  mutate(evt_lf_macro = ifelse(!is.na(EVT_LF) & EVT_LF == "Tree", "Tree", "nonTree"),
         leaks = macro != evt_lf_macro) %>%
  select(VALUE, EVT_NAME, EVT_LF, n_plots, segment, macro, cohesion, leaks) %>%
  arrange(macro, segment, desc(n_plots))
evt_map_c
```

## Stratification (constrained typology)

Tree segments are phase-subdivided (top `r round(100*params$tree_phase_frac)`% by
`z_phase` spread into `r params$k_phase` groups each); non-Tree segments are not.
The strata are then `r params$K_strata_tree` Tree strata (Ward on the Tree fine
bins' mean response) plus `r params$K_strata_nontree` non-Tree strata (Ward on the
non-Tree segments' mean response).

Each response is stratified two ways: **self-trained** (the fine-bin -> stratum
aggregation trained on that response) and **BALIVE-trained** (the aggregation
trained once on BALIVE and applied *unchanged*). Comparing them shows whether a
binning learned on one attribute transfers to attributes it never saw — the
type/phase fine bins are already response-agnostic (EVT + `z_phase` only); only
this last aggregation step is supervised. On the BALIVE panel the two rows
coincide by construction (a useful sanity check).

```{r c-analysis}
analysis_c <- analysis %>%
  inner_join(select(emb, CN, type_cluster_c, macro_c), by = "CN")
cat(sprintf("plots: %d (%d Tree / %d non-Tree)\n",
            nrow(analysis_c), sum(analysis_c$macro_c == "Tree"),
            sum(analysis_c$macro_c == "nonTree")))
```

```{r c-fine-bins}
# phase-subdivide the top fraction of TREE segments only
Zp_c    <- scale(as.matrix(analysis_c[, emb_phase_vars]))
tree_cl <- unique(as.character(analysis_c$type_cluster_c[analysis_c$macro_c == "Tree"]))
pvar_c  <- vapply(tree_cl, function(b) {
  ix <- which(as.character(analysis_c$type_cluster_c) == b)
  sum(apply(Zp_c[ix, , drop = FALSE], 2, var))
}, numeric(1))
n_sub_c  <- max(1, round(params$tree_phase_frac * length(tree_cl)))
sub_cl_c <- names(sort(pvar_c, decreasing = TRUE))[seq_len(n_sub_c)]

analysis_c$phase_sub_c <- 1L
for (b in sub_cl_c) {
  ix <- which(as.character(analysis_c$type_cluster_c) == b)
  if (length(ix) >= 10 * params$k_phase)
    analysis_c$phase_sub_c[ix] <- ward(Zp_c[ix, , drop = FALSE], params$k_phase)
}
analysis_c$fine_bin_c <- interaction(analysis_c$type_cluster_c, analysis_c$phase_sub_c, drop = TRUE)
cat(sprintf("Tree segments: %d (%d subdivided into %d phase groups); fine bins: %d\n",
            length(tree_cl), length(sub_cl_c), params$k_phase, nlevels(analysis_c$fine_bin_c)))

# Tree segments ranked by z_phase variance, flagged if subdivided
tibble(tree_segment = tree_cl,
       n         = as.integer(table(analysis_c$type_cluster_c)[tree_cl]),
       phase_var = round(as.numeric(pvar_c), 3),
       subdivided = tree_cl %in% sub_cl_c) %>%
  arrange(desc(phase_var))
```

```{r c-strata-helpers}
# Train the constrained aggregation on (train_df, train_y): Tree fine bins ->
# K_strata_tree, non-Tree segments -> K_strata_nontree, both by Ward on the
# group-mean response. Returns a mapping (fine_bin/segment -> stratum label)
# that can be applied to ANY df, so a binning trained on one response can be
# scored on another.
fit_agg_macro <- function(train_df, train_y) {
  dt  <- train_df[train_df$macro_c == "Tree", ]
  bmt <- bin_map(droplevels(factor(dt$fine_bin_c)), dt[[train_y]], params$K_strata_tree)
  dn  <- train_df[train_df$macro_c == "nonTree", ]
  bmn <- bin_map(droplevels(factor(dn$type_cluster_c)), dn[[train_y]], params$K_strata_nontree)
  list(tree    = setNames(paste0("T", bmt$stratum), as.character(bmt$fine_bin)),
       nontree = setNames(paste0("N", bmn$stratum), as.character(bmn$fine_bin)))
}
apply_agg_macro <- function(df, mapping) {
  s <- character(nrow(df)); is_tree <- df$macro_c == "Tree"
  s[is_tree]  <- mapping$tree[as.character(df$fine_bin_c[is_tree])]
  s[!is_tree] <- mapping$nontree[as.character(df$type_cluster_c[!is_tree])]
  factor(s)
}
# self-trained: aggregation trained and applied on the same response
agg_macro <- function(df, y) apply_agg_macro(df, fit_agg_macro(df, y))
# a single BALIVE-trained mapping, applied UNCHANGED to volume/removals to test
# whether a BALIVE binning transfers to attributes it was not trained on
map_balive_c <- fit_agg_macro(analysis_c, "BALIVE")

show_map_c <- function(df, y) {
  dt <- df[df$macro_c == "Tree", ]
  bmt <- bin_map(droplevels(factor(dt$fine_bin_c)), dt[[y]], params$K_strata_tree) %>%
    transmute(bin = as.character(fine_bin), macro = "Tree", n_plots,
              mean_response = round(mean_y, 2), stratum = paste0("T", stratum))
  dn <- df[df$macro_c == "nonTree", ]
  bmn <- bin_map(droplevels(factor(dn$type_cluster_c)), dn[[y]], params$K_strata_nontree) %>%
    transmute(bin = paste0("seg ", as.character(fine_bin)), macro = "nonTree", n_plots,
              mean_response = round(mean_y, 2), stratum = paste0("N", stratum))
  bind_rows(bmt, bmn) %>% arrange(stratum, desc(mean_response))
}

# stratifications compared for the CONSTRAINED method. Two EVT variants:
#  - self-trained: strata aggregated on the response being scored
#  - BALIVE-trained: the fixed BALIVE mapping applied unchanged (transfer test)
schemes_c <- function(df, y) list(
  "TCC FIA fixed"      = cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                             labels = c("0-10", "11-48", "49-85", "86-100")),
  "TCC DH"             = dh_strata(df$tcc, params$K_strata),
  "EVT self-trained"   = agg_macro(df, y),
  "EVT BALIVE-trained" = apply_agg_macro(df, map_balive_c),
  "flat probe"         = oof_dh(df, y, emb_vars, params$K_strata))

resp_table_c <- function(df, y, vars = emb_vars) {
  sch <- schemes_c(df, y)
  vv  <- c(pooled_within(df, "UNITCD", y),
           sapply(sch, function(s) pooled_within(mutate(df, .s = s), c("UNITCD", ".s"), y)))
  kk <- sprintf("%d+%d", params$K_strata_tree, params$K_strata_nontree)
  data.frame(
    stratification = c("UNITCD only (baseline)",
                       "UNITCD x TCC FIA fixed (0-10/11-48/49-85/86-100)",
                       sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                       sprintf("UNITCD x EVT constrained, self-trained (%s)", kk),
                       sprintf("UNITCD x EVT constrained, BALIVE-trained (%s)", kk),
                       sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var = round(vv, 2),
    within_sd  = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1))
}
```

```{r c-removals-pop}
ar_c <- analysis_c %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # GRM/remeasured; zeros kept
cat(sprintf("remeasured plots with GRM data: %d (%d Tree / %d non-Tree; %.1f%% with any removal)\n",
            nrow(ar_c), sum(ar_c$macro_c == "Tree"), sum(ar_c$macro_c == "nonTree"),
            100 * mean(ar_c$remv_cfgs > 0)))
```

### BALIVE (constrained)

```{r c-strat-balive}
resp_table_c(analysis_c, "BALIVE")
```

```{r c-strat-balive-map}
show_map_c(analysis_c, "BALIVE")
```

Strata sizes, mean, and SD (BALIVE):

```{r c-strat-balive-strata}
sch_bal_c <- schemes_c(analysis_c, "BALIVE")
strata_summary(analysis_c, "BALIVE", sch_bal_c)
```

Distribution of BALIVE within each stratum:

```{r c-strat-balive-box, fig.width=11, fig.height=4}
strata_boxplot(analysis_c, "BALIVE", sch_bal_c, ylab = "BALIVE")
```

### Standing live volume (constrained)

```{r c-strat-volume}
resp_table_c(analysis_c, "volume")
```

```{r c-strat-volume-map}
show_map_c(analysis_c, "volume")
```

Strata sizes, mean, and SD (volume):

```{r c-strat-volume-strata}
sch_vol_c <- schemes_c(analysis_c, "volume")
strata_summary(analysis_c, "volume", sch_vol_c)
```

Distribution of volume within each stratum:

```{r c-strat-volume-box, fig.width=11, fig.height=4}
strata_boxplot(analysis_c, "volume", sch_vol_c, ylab = "volume")
```

### Removals (constrained)

```{r c-strat-removals}
resp_table_c(ar_c, "remv_cfgs")
```

```{r c-strat-removals-map}
show_map_c(ar_c, "remv_cfgs")
```

Strata sizes, mean, and SD (removals):

```{r c-strat-removals-strata}
sch_rmv_c <- schemes_c(ar_c, "remv_cfgs")
strata_summary(ar_c, "remv_cfgs", sch_rmv_c)
```

Removals by stratum for each scheme (`log1p` axis; the constrained method's
non-Tree plots form `r params$K_strata_nontree` strata `N1..N`):

```{r c-strat-removals-box, fig.width=11, fig.height=4}
strata_boxplot(ar_c, "remv_cfgs", sch_rmv_c, ylab = "removals", logy = TRUE)
```

# Full-embedding clustering (type + phase, one step)

*(A third stratification method — the two above are untouched.)*

The EVT methods above cluster only `z_type` (then refine with a phase step). This
one clusters the **full embedding** — all `r length(emb_type_vars)` `z_type` +
`r length(emb_phase_vars)` `z_phase` dims — in a **single k-means step**, so type
*and* temporal-phase structure are captured jointly, with no separate
phase-subdivision stage (k-means rather than EVT-centroid Ward, which would be
one-per-EVT at this K). The `r params$K_full` clusters are aggregated to
`r params$K_strata` strata by Ward on the cluster-mean response. It uses **no EVT
and no FIA target** to form the clusters; only the final aggregation is supervised.

Each response is compared across: the **original EVT type/phase hierarchical**
binning (`z_type`), this **full one-step clustering**, **TCC** (fixed + DH), and
the **flat supervised probe**. For volume and removals we also report a
**transfer** table: each clustering's fine-bin -> stratum aggregation trained once
on **BALIVE** and applied unchanged, versus trained on the response itself —
showing how well a BALIVE-tuned stratification carries to other attributes.

```{r full-cluster}
Zfull <- scale(as.matrix(emb[, emb_vars]))          # standardized z_type + z_phase
set.seed(1)
kmf <- kmeans(Zfull, centers = params$K_full, nstart = 10, iter.max = 50)
emb$full_cluster <- factor(kmf$cluster)
cat(sprintf("full-embedding k-means (one step): %d clusters over %d dims; sizes %d..%d\n",
            length(unique(kmf$cluster)), length(emb_vars),
            min(table(kmf$cluster)), max(table(kmf$cluster))))
```

```{r full-assemble}
analysis_f <- analysis %>% inner_join(select(emb, CN, full_cluster), by = "CN")
ar_f       <- analysis_f %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # removals pop
cat(sprintf("plots: %d   remeasured with GRM: %d\n", nrow(analysis_f), nrow(ar_f)))
```

```{r full-helpers}
# fit/apply a fine-bin -> stratum aggregation, so a binning trained on one
# response can be scored on another (the BALIVE-trained transfer test)
fit_agg <- function(train_df, train_y, binvar) {
  bm <- bin_map(train_df[[binvar]], train_df[[train_y]], params$K_strata)
  setNames(as.character(bm$stratum), as.character(bm$fine_bin))
}
apply_agg <- function(df, binvar, mapping) factor(unname(mapping[as.character(df[[binvar]])]))
map_bal_evt  <- fit_agg(analysis_f, "BALIVE", "fine_bin")      # EVT type/phase, BALIVE-trained
map_bal_full <- fit_agg(analysis_f, "BALIVE", "full_cluster")  # full one-step,  BALIVE-trained

# main comparison: TCC, original EVT hierarchical, full one-step clustering, flat probe
schemes_f <- function(df, y) list(
  "TCC FIA fixed"   = cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                          labels = c("0-10", "11-48", "49-85", "86-100")),
  "TCC DH"          = dh_strata(df$tcc, params$K_strata),
  "EVT type/phase"  = agg_by(df$fine_bin, df[[y]], params$K_strata),
  "full type+phase" = agg_by(df$full_cluster, df[[y]], params$K_strata),
  "flat probe"      = oof_dh(df, y, emb_vars, params$K_strata))

resp_table_f <- function(df, y) {
  sch <- schemes_f(df, y)
  vv  <- c(pooled_within(df, "UNITCD", y),
           sapply(sch, function(s) pooled_within(mutate(df, .s = s), c("UNITCD", ".s"), y)))
  data.frame(
    stratification = c("UNITCD only (baseline)",
                       "UNITCD x TCC FIA fixed (0-10/11-48/49-85/86-100)",
                       sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                       sprintf("UNITCD x EVT type/phase hierarchical, z_type (%d)", params$K_strata),
                       sprintf("UNITCD x full-embedding type+phase, one step (%d -> %d)",
                               params$K_full, params$K_strata),
                       sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var = round(vv, 2), within_sd = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1))
}
# transfer: self-trained vs BALIVE-trained aggregation, for both clusterings
transfer_table_f <- function(df, y) {
  base <- pooled_within(df, "UNITCD", y)
  sch <- list(
    "EVT type/phase (self-trained)"    = apply_agg(df, "fine_bin",     fit_agg(df, y, "fine_bin")),
    "EVT type/phase (BALIVE-trained)"  = apply_agg(df, "fine_bin",     map_bal_evt),
    "full type+phase (self-trained)"   = apply_agg(df, "full_cluster", fit_agg(df, y, "full_cluster")),
    "full type+phase (BALIVE-trained)" = apply_agg(df, "full_cluster", map_bal_full))
  vv <- sapply(sch, function(s) pooled_within(mutate(df, .s = s), c("UNITCD", ".s"), y))
  data.frame(stratification = names(sch), within_var = round(vv, 2),
             within_sd = round(sqrt(vv), 3), pct_of_baseline = round(100 * vv / base, 1),
             row.names = NULL)
}
# full-cluster -> stratum mapping for display
show_map_f <- function(df, y) {
  bin_map(df$full_cluster, df[[y]], params$K_strata) %>%
    transmute(cluster = fine_bin, n_plots, mean_response = round(mean_y, 2), stratum) %>%
    arrange(stratum, desc(mean_response))
}
```

### BALIVE (full-embedding)

```{r full-balive}
resp_table_f(analysis_f, "BALIVE")
```

Strata sizes, mean, and SD (BALIVE):

```{r full-balive-strata}
sch_bal_f <- schemes_f(analysis_f, "BALIVE")
strata_summary(analysis_f, "BALIVE", sch_bal_f)
```

```{r full-balive-box, fig.width=12, fig.height=4}
strata_boxplot(analysis_f, "BALIVE", sch_bal_f, ylab = "BALIVE")
```

### Standing live volume (full-embedding)

```{r full-volume}
resp_table_f(analysis_f, "volume")
```

Transfer — clusterings trained on BALIVE, applied to volume:

```{r full-volume-transfer}
transfer_table_f(analysis_f, "volume")
```

```{r full-volume-strata}
sch_vol_f <- schemes_f(analysis_f, "volume")
strata_summary(analysis_f, "volume", sch_vol_f)
```

```{r full-volume-box, fig.width=12, fig.height=4}
strata_boxplot(analysis_f, "volume", sch_vol_f, ylab = "volume")
```

### Removals (full-embedding)

```{r full-removals}
resp_table_f(ar_f, "remv_cfgs")
```

Transfer — clusterings trained on BALIVE, applied to removals:

```{r full-removals-transfer}
transfer_table_f(ar_f, "remv_cfgs")
```

```{r full-removals-strata}
sch_rmv_f <- schemes_f(ar_f, "remv_cfgs")
strata_summary(ar_f, "remv_cfgs", sch_rmv_f)
```

```{r full-removals-box, fig.width=12, fig.height=4}
strata_boxplot(ar_f, "remv_cfgs", sch_rmv_f, ylab = "removals", logy = TRUE)
```

# Forest / non-forest detection (OOF)

Does the embedding detect non-forest as well as TCC? The stratification tables
suggest TCC's full-population edge is really forest/non-forest separation, so
here we test that directly, as a **classification** task rather than through
BALIVE variance. Label: a plot is **non-forest** when its dominant FIA condition
has no forest type (`is.na(FORTYPCD)`) — the agency's own forest/non-forest
determination, independent of the embedding and of BALIVE.

Four detectors, all scored out-of-fold (5-fold) where a model is fit:

1. **TCC 0-10 (baseline)** — the operational rule: `tcc <= 10` -> non-forest.
2. **logit z_type** — logistic regression of non-forest on the 48 `z_type` dims.
3. **logit z_type + z_phase** — same, adding the 8 `z_phase` dims.
4. **k-20 type-cluster aggregation** — the unsupervised `r params$K`-segment
   EVT-trained `type_cluster`, aggregated by each cluster's training non-forest
   rate (OOF). This is the *clustering* pathway that the hierarchical
   stratification actually uses, so it isolates whether the clustering — not the
   representation — is what loses the signal.

`AUC` is threshold-free (how separable the classes are under each score). The
operating-point metrics (`accuracy`/`sensitivity`/`specificity`/`precision`) are
all evaluated at a **matched budget**: each score flags the same number of plots
as the `tcc <= 10` rule, so they are directly comparable to the baseline.
`sensitivity` is recall of the non-forest class.

```{r detect-setup}
det <- analysis %>% mutate(nonforest = as.integer(is.na(FORTYPCD)))
lab <- det$nonforest
cat(sprintf("plots: %d   non-forest (is.na FORTYPCD): %d (%.1f%%)\n",
            nrow(det), sum(lab), 100 * mean(lab)))

set.seed(1); folds <- sample(rep(1:5, length.out = nrow(det)))

# rank-based (Mann-Whitney) AUC; higher score = more likely non-forest
auc <- function(score, label) {
  r <- rank(score); n1 <- sum(label == 1); n0 <- sum(label == 0)
  (sum(r[label == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
op_metrics <- function(pred_pos, label) {
  tp <- sum(pred_pos & label == 1); fp <- sum(pred_pos & label == 0)
  fn <- sum(!pred_pos & label == 1); tn <- sum(!pred_pos & label == 0)
  c(accuracy = (tp + tn) / length(label), sensitivity = tp / (tp + fn),
    specificity = tn / (tn + fp), precision = tp / (tp + fp))
}
# OOF logistic P(non-forest)
oof_logit <- function(df, vars) {
  form <- reformulate(vars, "nonforest"); s <- numeric(nrow(df))
  for (f in 1:5) {
    fit <- glm(form, family = binomial, data = df[folds != f, ])
    s[folds == f] <- predict(fit, df[folds == f, ], type = "response")
  }
  s
}
# OOF cluster aggregation: assign each plot its cluster's training non-forest rate
oof_cluster <- function(df, cl) {
  cl <- as.character(cl); s <- numeric(nrow(df))
  for (f in 1:5) {
    rate <- tapply(df$nonforest[folds != f], cl[folds != f], mean)
    gm   <- mean(df$nonforest[folds != f])
    v <- rate[cl[folds == f]]; v[is.na(v)] <- gm; s[folds == f] <- v
  }
  s
}
```

```{r detect-eval}
p_type <- oof_logit(det, emb_type_vars)
p_tp   <- oof_logit(det, emb_vars)
p_cl   <- oof_cluster(det, det$type_cluster)

k    <- sum(det$tcc <= 10)                       # matched flagging budget
topk <- function(sc) sc >= sort(sc, decreasing = TRUE)[k]
mk <- function(nm, sc, pred) {
  m <- op_metrics(pred, lab)
  data.frame(scheme = nm, AUC = round(auc(sc, lab), 3),
             accuracy = round(m["accuracy"], 3),
             sens_nonforest = round(m["sensitivity"], 3),
             specificity = round(m["specificity"], 3),
             precision = round(m["precision"], 3), row.names = NULL)
}
cat(sprintf("matched budget: %d plots flagged non-forest (= plots with tcc <= 10)\n", k))
bind_rows(
  mk("TCC 0-10 (baseline)",           100 - det$tcc, det$tcc <= 10),
  mk("logit z_type (OOF)",            p_type,        topk(p_type)),
  mk("logit z_type + z_phase (OOF)",  p_tp,          topk(p_tp)),
  mk("k-20 type-cluster agg (OOF)",   p_cl,          topk(p_cl)))
```

Read it as: if **logit z_type** matches or beats the TCC baseline on AUC, the
embedding detects non-forest at least as well as canopy cover — so the
hierarchical method's weakness is not the representation. If the **k-20
cluster aggregation** trails the logits, the *clustering* is where the signal is
lost (20 type segments cannot carve out a clean non-forest region the way a
supervised direction does), which is exactly the failure mode behind the BALIVE
stratification gap.

## Notes

- **EVT-trained, location-assigned.** EVT groups define the segment centroids;
  each plot is then assigned by nearest centroid in `z_type`, so an EVT may split
  across segments. The `EVT cohesion` diagnostic reports how much they stay
  together (high if `z_type` respects EVT).
- **The compactness ratio is the headline diagnostic.** Near 1.0 = `z_type`
  already encodes EVT structure (the model learned vegetation type), so the
  EVT-trained segments are almost as tight as free clusters. Far above 1 = EVTs
  are smeared across the embedding and the supervision is expensive.
- **Rare EVTs** (`n < min_evt_n`) only seed the *grouping* via the nearest
  EVT-centroid cluster; their plots, like all plots, are then assigned by
  location. Raise `min_evt_n` to keep noisier EVT centroids out of the grouping.
- **Choosing K.** Read it off the dendrogram; `K` is a param. Unweighted centroids
  give rare and common EVTs equal say in the grouping -- weight `cent_big` rows by
  `sqrt(n)` if you want common EVTs to dominate the merges.
- **Using it downstream.** `emb$type_cluster` (or `evt_map`) is a drop-in
  replacement for the k-means/Ward `type_bin` step in the hierarchical
  stratification -- an EVT-supervised type binning that never sees the FIA target.
- **Fine bins = EVT type + phase.** The EVT clusters are phase-subdivided (top
  `phase_subdivide_frac` by `z_phase` spread, `k_phase` groups each via Ward) to
  match the geometry hierarchical script; set `phase_subdivide_frac = 0` to test
  EVT-type-only fine bins.
- **Responses.** BALIVE and standing live volume (`TREE.VOLCFNET`, `STATUSCD=1`)
  are stocks on the **full plot population** (no forest-code filter; non-forest
  plots contribute 0). Removals (`TREE_GRM_COMPONENT`, annual per-acre removed GS
  volume) additionally require remeasurement/GRM data, so the inner join to `remv`
  scopes them to treed plots (retaining land-use diversions) — this replaces the
  old `FORTYPCD`-not-NA proxy, which dropped diverted plots and is no longer used. The `EVT type/phase hierarchical` row aggregates the fine bins
  to `K_strata` by Ward on the fine-bin-mean response; `flat embedding supervised`
  is the direct OOF (type+phase) probe ceiling for the same information.
- **Constrained Tree / non-Tree method** (second section, independent of the
  original). Cannot-link Ward on `EVT_LF` macro life-form makes every segment
  macro-pure at the clustering step; plots are still assigned by nearest `z_type`
  centroid alone, so `leakage` measures how clean the Tree/non-Tree boundary is
  in the embedding. Only Tree segments are phase-subdivided (top
  `tree_phase_frac`, `k_phase` groups); strata are `K_strata_tree` Tree strata
  (Ward on Tree fine-bin means) + `K_strata_nontree` non-Tree strata (Ward on
  non-Tree segment means). Tunable via `tree_phase_frac`, `K_strata_tree`,
  `K_strata_nontree` — the original method's `phase_subdivide_frac` / `K_strata`
  are left unchanged.
- **Full-embedding clustering (third method).** A single k-means step on the full
  `z_type + z_phase` space into `K_full` clusters (default 50; k-means rather than
  EVT-centroid Ward, which would be one-per-EVT at this K), aggregated to
  `K_strata` by Ward on the cluster-mean response — type and phase captured in one
  step, no phase-subdivision stage. The section compares it against the original
  EVT type/phase hierarchical binning, TCC, and the flat probe, and includes a
  BALIVE-trained transfer table (`fit_agg`/`apply_agg`) showing how each
  clustering, aggregated on BALIVE, carries to volume and removals. Tunable via
  `K_full`.
- **Forest / non-forest detection (OOF).** A direct classification check
  (`is.na(FORTYPCD)` label) of whether the embedding separates non-forest as well
  as TCC, comparing the `tcc <= 10` rule, OOF logistic probes on `z_type` and
  `z_type + z_phase`, and an OOF aggregation of the `K`-segment `type_cluster`.
  AUC is threshold-free; the operating-point metrics use a budget matched to the
  `tcc <= 10` rule. A logit that matches TCC while the cluster aggregation trails
  it localizes the stratification gap to the clustering, not the representation.
- **Self-trained vs BALIVE-trained (transfer test).** Each response is scored
  with two EVT stratifications: one whose fine-bin -> stratum aggregation is
  trained on that response (`agg_macro`/`fit_agg_macro`), and one that reuses a
  single BALIVE-trained mapping (`map_balive_c`) unchanged. The gap between them
  is the cost of transferring a BALIVE binning to volume/removals; on BALIVE they
  are identical by construction.
