---
title: "HDBSCAN post-stratification"
subtitle: "Density-based embedding strata vs TCC and a flat probe"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  K_type: 20         # target HDBSCAN clusters, z_type space
  K_tp: 50           # target HDBSCAN clusters, z_type + z_phase space
  K_strata: 5        # final strata after aggregation (and flat-probe / TCC-DH bins)
  minpts_grid: "10,20,40,70,120,200,320,500"   # minPts searched to hit target k
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 5, dpi = 110)
```

## Overview

Density-based post-stratification. HDBSCAN clusters the embedding without assuming
a globally Euclidean geometry — it uses only *local* density/reachability, so it
can follow curved, variable-density structure that k-means/Ward (which fit
compact, convex, globally-Euclidean clusters) cannot. Two clusterings are built:
one on the **type** space (`z_type`), one on the **type + phase** space
(`z_type + z_phase`). Each is aggregated to strata by the training target, and
compared against the operational TCC bins and a flat supervised probe.

Stratifications compared, per target:

1. **TCC FIA fixed** — canopy cover at the operational breaks (0-10/11-48/49-85/86-100).
2. **HDBSCAN type** (k ≈ `r params$K_type` on `z_type`) → aggregated to
   `r params$K_strata` strata by Ward on the cluster-mean target.
3. **HDBSCAN type+phase** (k ≈ `r params$K_tp` on `z_type + z_phase`) → aggregated
   to `r params$K_strata` strata by Ward on the cluster-mean target.
4. **flat probe** — OOF linear probe of the full embedding → target, DH-binned.

Targets: **A** BALIVE, **B** standing live volume, **C** removal volume. For
**B** and **C** we also report a **transfer** table: each HDBSCAN clustering
aggregated once on **BALIVE** and applied unchanged, versus aggregated on the
target itself.

**HDBSCAN has no k.** Cluster count is governed by `minPts`; we search a `minPts`
grid and take the run whose cluster count is closest to the target. HDBSCAN also
labels low-density points as **noise** — those are assigned to their nearest
clustered neighbour (1-NN) so every plot lands in a bin. Achieved cluster counts
and noise fractions are reported; high-dimensional density estimation is hard, so
if the counts land far from the targets or noise is large, that is itself evidence
about how cleanly the embedding clusters.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
library(dbscan)
```

## Load embeddings

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(if_all(all_of(emb_vars), ~ !is.na(.)))          # complete embeddings only
cat(sprintf("plots with complete embeddings: %d\n", nrow(emb)))
```

## HDBSCAN clustering

```{r hdbscan}
# assign noise (cluster 0) to nearest clustered neighbour (1-NN in the same space)
assign_noise <- function(X, cl) {
  noise <- cl == 0
  if (!any(noise) || all(noise)) return(cl)
  keep <- which(!noise)
  if (requireNamespace("FNN", quietly = TRUE)) {
    nn <- FNN::get.knnx(X[keep, , drop = FALSE], X[noise, , drop = FALSE], k = 1)$nn.index[, 1]
    cl[noise] <- cl[keep][nn]
  } else {                                                # fallback: nearest centroid
    ids   <- sort(unique(cl[keep]))
    cents <- t(sapply(ids, function(k) colMeans(X[cl == k, , drop = FALSE])))
    d <- sapply(seq_len(nrow(cents)), function(j) rowSums(sweep(X[noise, , drop = FALSE], 2, cents[j, ], "-")^2))
    cl[noise] <- ids[max.col(-d, ties.method = "first")]
  }
  cl
}
# search minPts for the run closest to k_target; return clustering + diagnostics
hdbscan_k <- function(X, k_target, grid) {
  runs <- lapply(grid, function(mp) {
    h <- dbscan::hdbscan(X, minPts = mp)
    list(minPts = mp, cl = h$cluster,
         k = length(setdiff(unique(h$cluster), 0)), noise = mean(h$cluster == 0))
  })
  ks    <- vapply(runs, function(r) r$k, integer(1))
  valid <- which(ks >= 1); if (!length(valid)) valid <- seq_along(ks)
  best  <- runs[[ valid[which.min(abs(ks[valid] - k_target))] ]]
  list(cluster = factor(assign_noise(X, best$cl)),
       minPts = best$minPts, k = best$k, noise = best$noise,
       grid = data.frame(minPts = grid, k = ks,
                         noise = round(vapply(runs, function(r) r$noise, numeric(1)), 3)))
}

grid  <- as.integer(strsplit(params$minpts_grid, ",")[[1]])
Ztype <- scale(as.matrix(emb[, emb_type_vars]))          # standardized z_type
Zall  <- scale(as.matrix(emb[, emb_vars]))               # standardized z_type + z_phase

ht <- hdbscan_k(Ztype, params$K_type, grid)
hf <- hdbscan_k(Zall,  params$K_tp,   grid)
emb$hdb_type <- ht$cluster
emb$hdb_tp   <- hf$cluster
ktype <- ht$k; ktp <- hf$k

cat(sprintf("type       : k=%d clusters (minPts=%d, target %d, noise %.1f%% reassigned)\n",
            ht$k, ht$minPts, params$K_type, 100 * ht$noise))
cat(sprintf("type+phase : k=%d clusters (minPts=%d, target %d, noise %.1f%% reassigned)\n",
            hf$k, hf$minPts, params$K_tp, 100 * hf$noise))
```

minPts search paths (cluster count `k` and noise fraction at each `minPts`):

```{r hdbscan-grid}
list(type = ht$grid, type_phase = hf$grid)
```

## FIA targets and TCC

```{r fia-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

cond <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, CONDPROP_UNADJ, BALIVE FROM COND WHERE STATECD='%s'",
  params$statecd)) %>% mutate(PLT_CN = as.character(PLT_CN))
balive <- cond %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")

dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# A/B: standing LIVE net cu ft volume per acre (STATUSCD=1)
vol <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
     FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>% mutate(PLT_CN = as.character(PLT_CN))

# C: annual per-acre removed GS volume; COALESCE inside SUM so no-removal plots are 0
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>% mutate(PLT_CN = as.character(PLT_CN))

fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN))
dbDisconnect(con)
```

```{r tcc}
locs <- fia_locs %>% filter(CN %in% emb$CN) %>% distinct(CN, .keep_all = TRUE)
tccr <- rast(params$tcc)
pts  <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
tv   <- extract(tccr, project(pts, crs(tccr)), ID = FALSE)[[1]]
tv   <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
tcc_by_cn <- tibble(CN = locs$CN, tcc = tv)
```

```{r assemble}
analysis <- emb %>%
  select(CN, hdb_type, hdb_tp, all_of(emb_vars)) %>%
  inner_join(balive, by = c("CN" = "PLT_CN")) %>%
  left_join(vol,     by = c("CN" = "PLT_CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD), by = "CN") %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  mutate(volume = coalesce(volume, 0)) %>%
  filter(!is.na(UNITCD), !is.na(tcc), tcc >= 0, tcc <= 100)     # full population
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))     # removals: remeasured/GRM
cat(sprintf("plots: %d across %d units;  remeasured with GRM: %d\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD), nrow(ar)))
```

## Helpers

```{r helpers}
pooled_within <- function(df, groups, y) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
# Dalenius-Hodges cumulative-sqrt(f) boundaries
dh_breaks <- function(x, K, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(numeric(0))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
}
dh_strata <- function(x, K, nbins = 100) {
  bnd <- dh_breaks(x, K, nbins)
  if (length(bnd) == 0) return(factor(rep(1L, length(x))))
  factor(findInterval(as.numeric(x), c(-Inf, bnd, Inf)))
}
# fine bins -> strata: Ward on the fine-bin-mean response (supervised aggregation)
bin_map <- function(bin, y, K) {
  d  <- tibble(fine_bin = bin, .y = y) %>% group_by(fine_bin) %>%
        summarize(n_plots = n(), mean_y = mean(.y), .groups = "drop")
  Kk <- min(K, nrow(d))
  d$stratum <- if (nrow(d) <= 1) 1L else
    cutree(hclust(dist(scale(d$mean_y)), method = "ward.D2"), Kk)
  d
}
agg_by <- function(bin, y, K) {
  d <- bin_map(bin, y, K); factor(d$stratum[match(bin, d$fine_bin)])
}
# fit/apply a fine-bin -> stratum map so a binning trained on one target can be
# scored on another (BALIVE-trained transfer)
fit_agg <- function(train_df, train_y, binvar) {
  bm <- bin_map(train_df[[binvar]], train_df[[train_y]], params$K_strata)
  setNames(as.character(bm$stratum), as.character(bm$fine_bin))
}
apply_agg <- function(df, binvar, mapping) factor(unname(mapping[as.character(df[[binvar]])]))
# flat OOF supervised probe on the full embedding -> DH bins
oof_dh <- function(df, y, vars, K) {
  set.seed(1); folds <- sample(rep(1:5, length.out = nrow(df)))
  d <- df %>% mutate(y_resp = .data[[y]])
  form <- reformulate(vars, "y_resp"); s <- numeric(nrow(df))
  for (f in 1:5) { fit <- lm(form, d[folds != f, ]); s[folds == f] <- predict(fit, d[folds == f, ]) }
  dh_strata(s, K)
}
# per-stratum n / mean / sd and response-by-stratum boxplots
strata_summary <- function(df, y, schemes) {
  dplyr::bind_rows(lapply(names(schemes), function(nm)
    tibble(scheme = nm, stratum = as.character(schemes[[nm]]), val = df[[y]]) %>%
      group_by(scheme, stratum) %>%
      summarize(n = dplyr::n(), mean = round(mean(val), 3), sd = round(sd(val), 3),
                .groups = "drop"))) %>%
    mutate(scheme = factor(scheme, levels = names(schemes))) %>% arrange(scheme, stratum)
}
strata_boxplot <- function(df, y, schemes, ylab = y, logy = FALSE) {
  long <- dplyr::bind_rows(lapply(names(schemes), function(nm)
    tibble(scheme = nm, stratum = as.character(schemes[[nm]]), val = df[[y]])))
  long$scheme <- factor(long$scheme, levels = names(schemes))
  long$yval   <- if (logy) log1p(long$val) else long$val
  ggplot(long, aes(x = stratum, y = yval)) +
    geom_boxplot(outlier.size = 0.4, fill = "grey85", linewidth = 0.3) +
    facet_wrap(~ scheme, scales = "free_x", nrow = 1) +
    labs(x = "stratum", y = if (logy) paste0("log1p(", ylab, ")") else ylab) +
    theme_bw(base_size = 10) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# BALIVE-trained maps (aggregation trained once on BALIVE, for the transfer test)
map_bal_type <- fit_agg(analysis, "BALIVE", "hdb_type")
map_bal_tp   <- fit_agg(analysis, "BALIVE", "hdb_tp")

# main comparison (self-trained: aggregation on the target being scored)
schemes <- function(df, y) list(
  "TCC FIA fixed"      = cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                             labels = c("0-10", "11-48", "49-85", "86-100")),
  "TCC DH"             = dh_strata(df$tcc, params$K_strata),
  "HDBSCAN type"       = agg_by(df$hdb_type, df[[y]], params$K_strata),
  "HDBSCAN type+phase" = agg_by(df$hdb_tp,   df[[y]], params$K_strata),
  "flat probe"         = oof_dh(df, y, emb_vars, params$K_strata))

resp_table <- function(df, y) {
  sch <- schemes(df, y)
  vv  <- c(pooled_within(df, "UNITCD", y),
           sapply(sch, function(s) pooled_within(mutate(df, .s = s), c("UNITCD", ".s"), y)))
  data.frame(
    stratification = c("UNITCD only (baseline)",
                       "UNITCD x TCC FIA fixed (0-10/11-48/49-85/86-100)",
                       sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                       sprintf("UNITCD x HDBSCAN type (k=%d -> %d)", ktype, params$K_strata),
                       sprintf("UNITCD x HDBSCAN type+phase (k=%d -> %d)", ktp, params$K_strata),
                       sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var = round(vv, 2), within_sd = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1),
    # vv[2] = TCC FIA fixed; RE = V_TCC/V_method (>1 beats TCC); ess_srs = N*S2/V_method
    RE_vs_tcc = round(vv[2] / vv, 2),
    ess_srs   = round(nrow(df) * var(df[[y]]) / vv))
}
# transfer: self-trained vs BALIVE-trained aggregation, for both HDBSCAN clusterings
transfer_table <- function(df, y) {
  base <- pooled_within(df, "UNITCD", y)
  sch <- list(
    "HDBSCAN type (self-trained)"         = apply_agg(df, "hdb_type", fit_agg(df, y, "hdb_type")),
    "HDBSCAN type (BALIVE-trained)"       = apply_agg(df, "hdb_type", map_bal_type),
    "HDBSCAN type+phase (self-trained)"   = apply_agg(df, "hdb_tp",   fit_agg(df, y, "hdb_tp")),
    "HDBSCAN type+phase (BALIVE-trained)" = apply_agg(df, "hdb_tp",   map_bal_tp))
  vv <- sapply(sch, function(s) pooled_within(mutate(df, .s = s), c("UNITCD", ".s"), y))
  data.frame(stratification = names(sch), within_var = round(vv, 2),
             within_sd = round(sqrt(vv), 3), pct_of_baseline = round(100 * vv / base, 1),
             row.names = NULL)
}
```

## A. BALIVE

```{r a-balive}
resp_table(analysis, "BALIVE")
```

```{r a-balive-strata}
sch_a <- schemes(analysis, "BALIVE")
strata_summary(analysis, "BALIVE", sch_a)
```

```{r a-balive-box, fig.width=11, fig.height=4}
strata_boxplot(analysis, "BALIVE", sch_a, ylab = "BALIVE")
```

## B. Standing live volume

```{r b-volume}
resp_table(analysis, "volume")
```

Transfer — HDBSCAN strata trained on BALIVE, applied to volume:

```{r b-volume-transfer}
transfer_table(analysis, "volume")
```

```{r b-volume-strata}
sch_b <- schemes(analysis, "volume")
strata_summary(analysis, "volume", sch_b)
```

```{r b-volume-box, fig.width=11, fig.height=4}
strata_boxplot(analysis, "volume", sch_b, ylab = "volume")
```

## C. Removal volume

```{r c-removals}
resp_table(ar, "remv_cfgs")
```

Transfer — HDBSCAN strata trained on BALIVE, applied to removals:

```{r c-removals-transfer}
transfer_table(ar, "remv_cfgs")
```

```{r c-removals-strata}
sch_c <- schemes(ar, "remv_cfgs")
strata_summary(ar, "remv_cfgs", sch_c)
```

```{r c-removals-box, fig.width=11, fig.height=4}
strata_boxplot(ar, "remv_cfgs", sch_c, ylab = "removals", logy = TRUE)
```

## Notes

- **Why HDBSCAN.** k-means/Ward fit compact convex clusters under a single global
  Euclidean metric; HDBSCAN builds clusters from *local* mutual-reachability
  density, so it can follow curved / variable-density manifolds where the space is
  only locally Euclidean. It picks the number of clusters from density structure
  rather than a fixed `k`.
- **Targeting k.** Since HDBSCAN has no `k`, we search `minpts_grid` and take the
  run whose cluster count is closest to `K_type` / `K_tp`. The `hdbscan-grid`
  table shows the `minPts` -> (k, noise) curve; widen or refine the grid if the
  achieved counts miss the targets.
- **Noise.** Low-density points (HDBSCAN cluster 0) are reassigned to their nearest
  clustered neighbour (1-NN, local — consistent with the local-Euclidean premise),
  so every plot is stratified. A large reassigned fraction means the embedding does
  not cluster cleanly at that scale.
- **Aggregation is the only supervised step.** The `K_type` / `K_tp` clusters are
  formed with no FIA target; only the final Ward aggregation of clusters to
  `K_strata` uses the response — self-trained (on the scored target) or, in the
  transfer tables, BALIVE-trained and applied unchanged to volume/removals.
- **Population.** BALIVE and volume run on the full plot population (no forest-code
  filter; non-forest contributes 0); removals additionally require GRM/remeasured
  data (`ar`), keeping forested zero-removal plots as 0.
- **Standardization.** Both spaces are `scale()`d before HDBSCAN so every embedding
  dimension contributes comparably to the reachability distances.
