---
title: "FIA Removals Post-Stratification: TCC vs Embedding"
subtitle: "Within-stratum variance of removed live basal area by survey unit, Virginia FIA plots"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  yr_start: 2017     # T2 measurement window (inclusive) -- the "current" remeasured plots
  yr_end:   2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
```

## Overview

A **removals** post-stratification analysis, parallel to the BALIVE one. The key
simplification (thanks to how FIA records trees): we do **not** difference two
measurements. Each tree at its measurement already carries `STATUSCD`, where
**`3 = cut and removed`** *since the previous visit*. So we just:

1. Take the **remeasured plots** whose measurement falls in `[yr_start, yr_end]`
   (removals are only defined where there was a prior visit — `PREV_PLT_CN` not
   null).
2. Pull their trees and aggregate the **removed** ones (`STATUSCD = 3`) into a
   per-plot **removed live basal area** (sq ft/acre), annualized by the plot's
   remeasurement period `REMPER`.
3. Stratify by **TCC** and by the **embedding (at that measurement)**, and compare
   the within-stratum variance of removals, within each survey unit (`UNITCD`).

> **This is true GRM-style removals** (harvest / cut), not net change — a tree is
> counted only if FIA flagged it cut. Contrast the earlier "BALIVE difference"
> idea, which nets in growth and mortality and needs a plot-to-plot link; this
> needs neither.

Why the embedding should shine here: removals = harvest = a **fast disturbance
event**, exactly what the phase pathway (`z_phase`, conditioned on
years-since-fast-change) is built to detect. A single-date TCC snapshot barely
encodes *whether a cut happened*.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Data assembly

### Remeasured plots in the window

```{r plots}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

plots <- dbGetQuery(
  con,
  sprintf("SELECT CN, PREV_PLT_CN, UNITCD, INVYR, MEASYEAR, REMPER
             FROM PLOT
            WHERE STATECD='%s' AND INVYR BETWEEN %d AND %d",
          params$statecd, params$yr_start, params$yr_end)
) %>%
  mutate(CN = as.character(CN), PREV_PLT_CN = as.character(PREV_PLT_CN)) %>%
  filter(!is.na(PREV_PLT_CN), !is.na(REMPER), REMPER > 0)   # remeasured only

cat(sprintf("remeasured plots in [%d, %d]: %d\n",
            params$yr_start, params$yr_end, nrow(plots)))
```

### Removed basal area per plot (from the TREE table, no linking)

Basal area of one tree (sq ft) = `0.005454 * DIA^2` (DIA in inches); scaled to a
per-acre contribution by `TPA_UNADJ`. We aggregate in SQL and also count removed
trees with vs without a `DIA`, so we can see whether `DIA` is populated for cut
trees (if not, switch the size source to `TREE_GRM_MIDPT` — see Notes).

```{r removals}
remv <- dbGetQuery(
  con,
  sprintf(
    "SELECT PLT_CN,
            SUM(CASE WHEN STATUSCD=3 THEN 0.005454*DIA*DIA*TPA_UNADJ ELSE 0 END) AS remv_ba,
            SUM(CASE WHEN STATUSCD=3 THEN 1 ELSE 0 END)                          AS remv_n,
            SUM(CASE WHEN STATUSCD=3 AND DIA IS NOT NULL THEN 1 ELSE 0 END)      AS remv_n_wdia
       FROM TREE
      WHERE STATECD='%s'
      GROUP BY PLT_CN",
    params$statecd
  )
) %>%
  mutate(PLT_CN = as.character(PLT_CN))

# DIA-coverage diagnostic for removed trees
tot <- remv %>% summarize(n = sum(remv_n), n_wdia = sum(remv_n_wdia))
cat(sprintf("removed trees: %d total, %.1f%% with a non-null DIA\n",
            tot$n, 100 * tot$n_wdia / max(tot$n, 1)))

plots <- plots %>%
  left_join(remv, by = c("CN" = "PLT_CN")) %>%
  mutate(remv_ba     = coalesce(remv_ba, 0),          # no removed trees -> 0
         remv_ba_ann = remv_ba / REMPER)              # sq ft/acre/yr
```

### TCC at the plot location

Science-TCC is categorical, so `extract` returns a factor — coerce to numeric and
NA the nodata codes (>100).

```{r tcc}
fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID))

locs <- fia_locs %>% filter(CN %in% plots$CN) %>% distinct(CN, .keep_all = TRUE)

tcc     <- rast(params$tcc)
pts     <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
pts_aea <- project(pts, crs(tcc))
vals <- extract(tcc, pts_aea, ID = FALSE); names(vals) <- "tcc"
vals$tcc <- as.numeric(as.character(vals$tcc))
vals$tcc[vals$tcc > 100] <- NA_real_

tcc_by_cn <- tibble(CN = locs$CN, tcc = vals$tcc)
```

### Embedding at the measurement

```{r embeddings}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb_data <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  distinct(CN, .keep_all = TRUE)
```

### One analysis frame

```{r assemble}
analysis <- plots %>%
  select(CN, UNITCD, REMPER, remv_ba, remv_ba_ann) %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  inner_join(select(emb_data, CN, all_of(emb_vars)), by = "CN") %>%
  filter(!is.na(UNITCD), !is.na(remv_ba_ann), !is.na(tcc), tcc >= 0, tcc <= 100)

cat(sprintf("plots in analysis frame: %d across %d survey units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
cat(sprintf("plots with any removal: %.1f%%\n", 100 * mean(analysis$remv_ba_ann > 0)))
summary(analysis$remv_ba_ann)
```

## Define the stratifications

Response for the supervised embedding score is `remv_ba_ann`. Everything else
mirrors the BALIVE post-stratification report.

```{r strata}
resp <- "remv_ba_ann"

# --- TCC strata: fixed canopy bins + equal-count quartiles ------------------
analysis <- analysis %>%
  mutate(tcc_stratum = cut(tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                           labels = c("0-10", "11-48", "49-85", "86-100")))
analysis$tcc_q <- factor(dplyr::ntile(analysis$tcc, 4), labels = paste0("T", 1:4))

# --- Embedding strata, UNSUPERVISED: 4 k-means regions of z_type ------------
set.seed(1)
Z  <- scale(as.matrix(analysis[, emb_type_vars]))
km <- kmeans(Z, centers = 4, nstart = 25)
analysis$emb_km <- factor(km$cluster, labels = paste0("K", 1:4))

# --- Embedding strata, SUPERVISED: OOF read-out of removals, quartile-binned -
set.seed(1)
folds <- sample(rep(1:5, length.out = nrow(analysis)))
oof_score <- function(vars) {
  form <- reformulate(vars, response = resp)
  s <- numeric(nrow(analysis))
  for (f in 1:5) {
    fit <- lm(form, data = analysis[folds != f, ])
    s[folds == f] <- predict(fit, newdata = analysis[folds == f, ])
  }
  s
}
analysis$emb_sup_type <- factor(dplyr::ntile(oof_score(emb_type_vars), 4), labels = paste0("St", 1:4))
analysis$emb_sup      <- factor(dplyr::ntile(oof_score(emb_vars),      4), labels = paste0("S",  1:4))
analysis$emb_tcc_sup  <- factor(dplyr::ntile(oof_score(c(emb_vars, "tcc")), 4), labels = paste0("C", 1:4))

cat("TCC fixed-bin counts:\n");                    print(table(analysis$tcc_stratum))
cat("\nEmbedding supervised type+phase counts:\n"); print(table(analysis$emb_sup))
```

## Within-stratum variance of removals

```{r within-var-fn}
pooled_within <- function(df, groups, y = resp) {
  g <- df %>%
    group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
```

### Headline: overall pooled within-stratum variance

Lower `within_var` / `pct_of_baseline < 100` = more homogeneous strata = a more
efficient post-stratified estimator of removals.

```{r headline}
v_unit  <- pooled_within(analysis, "UNITCD")
v_tcc   <- pooled_within(analysis, c("UNITCD", "tcc_stratum"))
v_tccq  <- pooled_within(analysis, c("UNITCD", "tcc_q"))
v_km    <- pooled_within(analysis, c("UNITCD", "emb_km"))
v_supt  <- pooled_within(analysis, c("UNITCD", "emb_sup_type"))
v_sup   <- pooled_within(analysis, c("UNITCD", "emb_sup"))
v_ct    <- pooled_within(analysis, c("UNITCD", "emb_tcc_sup"))

vv <- c(v_unit, v_tcc, v_tccq, v_km, v_supt, v_sup, v_ct)
data.frame(
  stratification  = c("UNITCD only (baseline)",
                      "UNITCD x TCC fixed bins (4)",
                      "UNITCD x TCC quartiles (4)",
                      "UNITCD x embedding k-means (4, unsup)",
                      "UNITCD x embedding supervised type-only (4, OOF)",
                      "UNITCD x embedding supervised type+phase (4, OOF)",
                      "UNITCD x TCC + embedding supervised (4, OOF)"),
  within_var      = round(vv, 2),
  within_sd       = round(sqrt(vv), 3),
  pct_of_baseline = round(100 * vv / v_unit, 1)
)
```

### Per survey unit

```{r per-unit}
per_unit <- analysis %>%
  group_by(UNITCD) %>%
  group_modify(~ tibble(
    n         = nrow(.x),
    pct_cut   = round(100 * mean(.x[[resp]] > 0), 1),
    var_total = var(.x[[resp]]),
    var_tcc   = pooled_within(.x, "tcc_stratum"),
    var_emb   = pooled_within(.x, "emb_sup")
  )) %>%
  ungroup() %>%
  mutate(tcc_ratio = round(var_tcc / var_total, 3),
         emb_ratio = round(var_emb / var_total, 3),
         winner    = if_else(var_emb < var_tcc, "embedding", "tcc"),
         across(c(var_total, var_tcc, var_emb), ~ round(.x, 2)))
per_unit
```

### Does each stratifier concentrate the harvested plots?

For a rare, zero-inflated event, a good stratifier puts the cut plots together.
Mean removals by stratum shows how peaked each scheme is.

```{r plot-conc, fig.width=11, fig.height=5}
op <- par(mfrow = c(1, 2), mar = c(4.4, 4.6, 3, 1))
barplot(tapply(analysis[[resp]], analysis$tcc_stratum, mean), col = "firebrick",
        las = 2, ylab = "mean removed BA (sq ft/acre/yr)",
        main = "Mean removals by TCC stratum")
barplot(tapply(analysis[[resp]], analysis$emb_sup, mean), col = "steelblue",
        las = 2, ylab = "mean removed BA (sq ft/acre/yr)",
        main = "Mean removals by embedding stratum")
par(op)
```

## Cleanup

```{r cleanup}
dbDisconnect(con)
```

## Notes

- **Removed-tree size source.** We compute removed BA from `TREE.DIA` at the
  measurement; the diagnostic above reports what fraction of `STATUSCD=3` trees
  have a non-null `DIA`. If that fraction is low, switch the size source to the
  GRM midpoint: join `TREE_GRM_COMPONENT` (`SUBP_TPAREMV_UNADJ_*`, component
  `LIKE 'CUT%%'`/`'DIVERSION%%'`) x `TREE_GRM_MIDPT` (`DIA`/`VOLCFNET` at period
  midpoint). That is FIA's own removals size and also needs no manual linking.
- **Land-use diversions.** `STATUSCD=3` is cut/removed; land-use *diversions*
  (forest -> non-forest) are handled in the GRM `COMPONENT` codes, not `STATUSCD`.
  If you want full removals (cut + diversion), use the GRM route above.
- **Unadjusted per-acre.** `remv_ba` uses `TPA_UNADJ` without the `POP_STRATUM`
  `ADJ_FACTOR_*`; fine for a *strata efficiency* comparison (adjustment is ~constant
  within design), not for an official population estimate.
- **Heavy zero-inflation.** Removals are ~0 on most plots; variance lives in the
  few cut plots. Also consider a **harvest-incidence** response
  (`as.integer(remv_ba_ann > 0)`) to stratify *occurrence* separately from
  magnitude — often the more stratifiable signal, and a natural fit for `z_phase`.
- **Expected result.** Because removals is a disturbance signal, expect the
  embedding (esp. `type+phase`) to beat TCC by a wider margin than it did for
  BALIVE, and the `type-only` vs `type+phase` gap to be large — the phase pathway
  is doing the work.
