---
title: "FIA Removals Post-Stratification: TCC vs Embedding"
subtitle: "Within-stratum variance of removed volume by survey unit, Virginia FIA plots"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  yr_start: 2017     # measurement window (inclusive) for the remeasured plots
  yr_end:   2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
```

## Overview

A **removals** post-stratification analysis, parallel to the BALIVE one, using
FIA's own GRM removal volumes — no plot-to-plot differencing.

1. Take the **remeasured plots** whose measurement falls in `[yr_start, yr_end]`
   (`PREV_PLT_CN` not null).
2. Per plot, sum FIA's per-tree removal volume × annual removal expansion from
   **`TREE_GRM_COMPONENT`**:

   $$\texttt{remv\_cfgs} = \sum_{\text{trees}} \texttt{REMVCFGS\_FOREST} \times
     \texttt{SUBP\_TPAREMV\_UNADJ\_GS\_FOREST}$$

   giving **annual per-acre removed merchantable growing-stock volume**
   (cu ft/acre/yr). (`REMVCFAL_FOREST`/`REMVBFSL_FOREST` give all-live sound cu ft
   and sawtimber board ft — see Notes to switch.)
3. Stratify by **TCC** and by the **embedding (at that measurement)** and compare
   the within-stratum variance of removals, within each survey unit (`UNITCD`).

> **This is true GRM removals** — cut **and** land-use diversion — at the plot,
> already annualized (`SUBP_TPAREMV_UNADJ_*` is per-year, so we do **not** divide
> by `REMPER`). Trees that were not removed contribute 0 (their `TPAREMV` is 0).

Why the embedding should shine here: removals = harvest = a **fast disturbance
event**, exactly what the phase pathway (`z_phase`, conditioned on
years-since-fast-change) is built to detect; a single-date TCC snapshot barely
encodes *whether a cut happened*.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Data assembly

### Remeasured plots in the window

```{r plots}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

plots <- dbGetQuery(
  con,
  sprintf("SELECT CN, PREV_PLT_CN, UNITCD, INVYR
             FROM PLOT
            WHERE STATECD='%s' AND INVYR BETWEEN %d AND %d",
          params$statecd, params$yr_start, params$yr_end)
) %>%
  mutate(CN = as.character(CN), PREV_PLT_CN = as.character(PREV_PLT_CN)) %>%
  filter(!is.na(PREV_PLT_CN))                    # remeasured only (removals defined)

cat(sprintf("remeasured plots in [%d, %d]: %d\n",
            params$yr_start, params$yr_end, nrow(plots)))
```

### Removed volume per plot (TREE_GRM_COMPONENT)

Sum FIA's per-tree removal volume × annual removal TPA. We restrict the (national)
GRM table to our plots via a temporary join table, and compute all three removal
bases (GS cu ft, AL sound cu ft, SL board ft) so the response is easy to swap.
`SUBP_TPAREMV_UNADJ_*` is already annual, so **no `REMPER` division**.

```{r removals}
# push the plot CNs into a temp table so the big GRM table is filtered by join
dbWriteTable(con, "tmp_plots", tibble(PLT_CN = plots$CN),
             temporary = TRUE, overwrite = TRUE)

remv <- dbGetQuery(
  con,
  "SELECT g.PLT_CN,
          SUM(g.REMVCFGS_FOREST * g.SUBP_TPAREMV_UNADJ_GS_FOREST) AS remv_cfgs,
          SUM(g.REMVCFAL_FOREST * g.SUBP_TPAREMV_UNADJ_AL_FOREST) AS remv_cfal,
          SUM(g.REMVBFSL_FOREST * g.SUBP_TPAREMV_UNADJ_SL_FOREST) AS remv_bfsl
     FROM TREE_GRM_COMPONENT g
     JOIN tmp_plots t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN"
) %>%
  mutate(PLT_CN = as.character(PLT_CN))
# NOTE: if TREE_GRM_COMPONENT lacks PLT_CN in your build, join TREE on
# g.TRE_CN = TREE.CN to recover TREE.PLT_CN.

plots <- plots %>%
  left_join(remv, by = c("CN" = "PLT_CN")) %>%
  mutate(across(c(remv_cfgs, remv_cfal, remv_bfsl), ~ coalesce(.x, 0)))  # no cut trees -> 0
```

### TCC at the plot location

Science-TCC is categorical, so `extract` returns a factor — coerce to numeric and
NA the nodata codes (>100).

```{r tcc}
fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID))

locs <- fia_locs %>% filter(CN %in% plots$CN) %>% distinct(CN, .keep_all = TRUE)

tcc     <- rast(params$tcc)
pts     <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
pts_aea <- project(pts, crs(tcc))
vals <- extract(tcc, pts_aea, ID = FALSE); names(vals) <- "tcc"
vals$tcc <- as.numeric(as.character(vals$tcc))
vals$tcc[vals$tcc > 100] <- NA_real_

tcc_by_cn <- tibble(CN = locs$CN, tcc = vals$tcc)
```

### Embedding at the measurement

```{r embeddings}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb_data <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  distinct(CN, .keep_all = TRUE)
```

### One analysis frame

```{r assemble}
analysis <- plots %>%
  select(CN, UNITCD, remv_cfgs, remv_cfal, remv_bfsl) %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  inner_join(select(emb_data, CN, all_of(emb_vars)), by = "CN") %>%
  filter(!is.na(UNITCD), !is.na(remv_cfgs), !is.na(tcc), tcc >= 0, tcc <= 100)

cat(sprintf("plots in analysis frame: %d across %d survey units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
cat(sprintf("plots with any GS removal: %.1f%%\n", 100 * mean(analysis$remv_cfgs > 0)))
summary(analysis$remv_cfgs)
```

## Define the stratifications

Response for the supervised embedding score is `remv_cfgs` (annual per-acre GS
removed volume). Everything else mirrors the BALIVE post-stratification report.

```{r strata}
resp <- "remv_cfgs"

# --- TCC strata: fixed canopy bins + equal-count quartiles ------------------
analysis <- analysis %>%
  mutate(tcc_stratum = cut(tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                           labels = c("0-10", "11-48", "49-85", "86-100")))
analysis$tcc_q <- factor(dplyr::ntile(analysis$tcc, 4), labels = paste0("T", 1:4))

# --- Embedding strata, UNSUPERVISED: 4 k-means regions of z_type ------------
set.seed(1)
Z  <- scale(as.matrix(analysis[, emb_type_vars]))
km <- kmeans(Z, centers = 4, nstart = 25)
analysis$emb_km <- factor(km$cluster, labels = paste0("K", 1:4))

# --- Embedding strata, SUPERVISED: OOF read-out of removals, quartile-binned -
set.seed(1)
folds <- sample(rep(1:5, length.out = nrow(analysis)))
oof_score <- function(vars) {
  form <- reformulate(vars, response = resp)
  s <- numeric(nrow(analysis))
  for (f in 1:5) {
    fit <- lm(form, data = analysis[folds != f, ])
    s[folds == f] <- predict(fit, newdata = analysis[folds == f, ])
  }
  s
}
analysis$emb_sup_type <- factor(dplyr::ntile(oof_score(emb_type_vars), 4), labels = paste0("St", 1:4))
analysis$emb_sup      <- factor(dplyr::ntile(oof_score(emb_vars),      4), labels = paste0("S",  1:4))
analysis$emb_tcc_sup  <- factor(dplyr::ntile(oof_score(c(emb_vars, "tcc")), 4), labels = paste0("C", 1:4))

cat("TCC fixed-bin counts:\n");                    print(table(analysis$tcc_stratum))
cat("\nEmbedding supervised type+phase counts:\n"); print(table(analysis$emb_sup))
```

## Within-stratum variance of removals

```{r within-var-fn}
pooled_within <- function(df, groups, y = resp) {
  g <- df %>%
    group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
```

### Headline: overall pooled within-stratum variance

Lower `within_var` / `pct_of_baseline < 100` = more homogeneous strata = a more
efficient post-stratified estimator of removals.

```{r headline}
v_unit  <- pooled_within(analysis, "UNITCD")
v_tcc   <- pooled_within(analysis, c("UNITCD", "tcc_stratum"))
v_tccq  <- pooled_within(analysis, c("UNITCD", "tcc_q"))
v_km    <- pooled_within(analysis, c("UNITCD", "emb_km"))
v_supt  <- pooled_within(analysis, c("UNITCD", "emb_sup_type"))
v_sup   <- pooled_within(analysis, c("UNITCD", "emb_sup"))
v_ct    <- pooled_within(analysis, c("UNITCD", "emb_tcc_sup"))

vv <- c(v_unit, v_tcc, v_tccq, v_km, v_supt, v_sup, v_ct)
data.frame(
  stratification  = c("UNITCD only (baseline)",
                      "UNITCD x TCC fixed bins (4)",
                      "UNITCD x TCC quartiles (4)",
                      "UNITCD x embedding k-means (4, unsup)",
                      "UNITCD x embedding supervised type-only (4, OOF)",
                      "UNITCD x embedding supervised type+phase (4, OOF)",
                      "UNITCD x TCC + embedding supervised (4, OOF)"),
  within_var      = round(vv, 2),
  within_sd       = round(sqrt(vv), 3),
  pct_of_baseline = round(100 * vv / v_unit, 1)
)
```

### Per survey unit

```{r per-unit}
per_unit <- analysis %>%
  group_by(UNITCD) %>%
  group_modify(~ tibble(
    n         = nrow(.x),
    pct_cut   = round(100 * mean(.x[[resp]] > 0), 1),
    var_total = var(.x[[resp]]),
    var_tcc   = pooled_within(.x, "tcc_stratum"),
    var_emb   = pooled_within(.x, "emb_sup")
  )) %>%
  ungroup() %>%
  mutate(tcc_ratio = round(var_tcc / var_total, 3),
         emb_ratio = round(var_emb / var_total, 3),
         winner    = if_else(var_emb < var_tcc, "embedding", "tcc"),
         across(c(var_total, var_tcc, var_emb), ~ round(.x, 2)))
per_unit
```

### Does each stratifier concentrate the harvested plots?

For a rare, zero-inflated event, a good stratifier puts the cut plots together.
Mean removals by stratum shows how peaked each scheme is.

```{r plot-conc, fig.width=11, fig.height=5}
op <- par(mfrow = c(1, 2), mar = c(4.4, 4.6, 3, 1))
barplot(tapply(analysis[[resp]], analysis$tcc_stratum, mean), col = "firebrick",
        las = 2, ylab = "mean removed GS volume (cu ft/acre/yr)",
        main = "Mean removals by TCC stratum")
barplot(tapply(analysis[[resp]], analysis$emb_sup, mean), col = "steelblue",
        las = 2, ylab = "mean removed GS volume (cu ft/acre/yr)",
        main = "Mean removals by embedding stratum")
par(op)
```

## Cleanup

```{r cleanup}
dbDisconnect(con)
```

## Notes

- **Which removal basis.** `remv_cfgs` = merchantable growing-stock cu ft
  (`REMVCFGS_FOREST` x `SUBP_TPAREMV_UNADJ_GS_FOREST`). Swap `resp` to `remv_cfal`
  (all-live sound cu ft) or `remv_bfsl` (sawtimber board ft) — both are already
  computed on the analysis frame.
- **Already annual.** `SUBP_TPAREMV_UNADJ_*` is a per-year expansion, so the
  product is annual per-acre removals; do **not** divide by `REMPER`.
- **Full removals.** GRM `REMV` includes cut **and** land-use diversion, so this
  is complete removals, not cut-only.
- **PLT_CN join.** If `TREE_GRM_COMPONENT` in your build has no `PLT_CN` column,
  join `TREE` on `g.TRE_CN = TREE.CN` to recover `TREE.PLT_CN`.
- **Unadjusted per-acre.** Uses `SUBP_TPAREMV_UNADJ_*` without the `POP_STRATUM`
  `ADJ_FACTOR_*`; fine for a *strata efficiency* comparison, not an official
  population estimate. (Also SUBP only — if you need micro/macroplot removals, add
  the `MICR_`/`MACR_` TPAREMV terms.)
- **Heavy zero-inflation.** Removals are ~0 on most plots; variance lives in the
  few cut plots. Consider a **harvest-incidence** response
  (`as.integer(remv_cfgs > 0)`) to stratify *occurrence* separately from
  magnitude — often the more stratifiable signal, and a natural fit for `z_phase`.
- **Expected result.** Because removals is a disturbance signal, expect the
  embedding (esp. `type+phase`) to beat TCC by a wider margin than for BALIVE, and
  the `type-only` vs `type+phase` gap to be large — the phase pathway is doing the
  work.
