---
title: "EVT-trained Type Segmentation"
subtitle: "z_type segments trained on EVT, plots assigned by embedding location"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  evt_table: "~/Dropbox/git/vq-vae/data/LF2024_EVT.csv"
  fia_locs: "/data/FIADB/spatial/va_merge.csv"
  tcc: "/data/usfs_science_tcc_v2023-5_wgs84/science_tcc_conus_wgs84_v2023-5_20230101_20231231.tif"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  K: 20            # number of EVT-cohesive type clusters
  min_evt_n: 20    # min plots for an EVT to be a clustering atom; smaller -> nearest cluster
  phase_subdivide_frac: 0.25  # fraction of type clusters (highest z_phase spread) to subdivide
  k_phase: 3       # phase sub-groups per subdivided cluster
  K_strata: 5      # final strata after supervised aggregation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5, dpi = 110)
```

## Overview

**Segment** the type embedding `z_type` using EVT as supervision, then assign each
plot to a segment by **its own location in `z_type`** — not by its EVT label.

1. Represent each EVT by its **centroid** in standardized `z_type` (reliable EVTs,
   `n >= min_evt_n`; rare EVTs attach to the nearest).
2. **Ward-cluster the EVT centroids** into `K` groups — EVT trains *which regions
   of the embedding form*.
3. Compute each group's centroid from its plots and **assign every plot to the
   nearest group centroid** — a Voronoi segmentation of `z_type` space.

So the segment boundaries are learned from EVT, but plots land in bins by *where
they sit* in the embedding; an EVT may split across bins if its plots are
embedding-diverse. EVT is external to the FIA responses, so this supervision
cannot leak into BALIVE/volume/removals. Diagnostics report how cohesive EVTs
remain and how compact the segmentation is vs a free k-means.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(terra)
```

## Load and per-EVT centroids

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(!is.na(evt), evt != -9999)                     # drop nodata EVT

Z    <- scale(as.matrix(emb[, emb_type_vars]))          # standardized z_type
evt  <- as.character(emb$evt)
evt_n <- table(evt)

# per-EVT centroid in standardized z_type
cent  <- rowsum(Z, evt) / as.numeric(evt_n[rownames(rowsum(Z, evt))])
big   <- names(evt_n)[evt_n >= params$min_evt_n]        # reliable centroids
cat(sprintf("plots: %d   EVTs: %d (%d with n >= %d used as clustering atoms)\n",
            nrow(emb), length(evt_n), length(big), params$min_evt_n))
```

## Ward-cluster the EVT centroids

```{r cluster, fig.width=11, fig.height=6}
cent_big <- cent[big, , drop = FALSE]
hc <- hclust(dist(cent_big), method = "ward.D2")
K  <- min(params$K, nrow(cent_big))
cl_big <- cutree(hc, k = K)                              # named by big EVT

# cluster centroids (rows ordered 1..K)
clcent <- t(sapply(sort(unique(cl_big)),
                   function(k) colMeans(cent_big[cl_big == k, , drop = FALSE])))

# EVT -> group map (the EVT-trained grouping): big EVTs from the Ward cut, rare
# EVTs attached to the nearest EVT-centroid cluster.
nearest_c <- function(v) which.min(colSums((t(clcent) - v)^2))
evt_group <- vapply(rownames(cent), function(e)
  if (e %in% big) cl_big[[e]] else nearest_c(cent[e, ]), integer(1))
names(evt_group) <- rownames(cent)

# Segment z_type: group centroids computed from the plots (seeded by the EVT
# grouping), then assign EVERY PLOT to the nearest group centroid by its own
# z_type location. An EVT may split across segments.
grp0 <- evt_group[evt]
glev <- sort(unique(grp0))
gc   <- t(sapply(glev, function(g) colMeans(Z[grp0 == g, , drop = FALSE])))
d2   <- sapply(seq_len(nrow(gc)), function(k) rowSums(sweep(Z, 2, gc[k, ], "-")^2))
emb$type_cluster <- factor(glev[max.col(-d2, ties.method = "first")])

# dendrogram of the big EVTs, boxed by cluster
op <- par(mar = c(6, 4, 3, 1))
plot(hc, labels = big, xlab = "EVT", sub = "", cex = 0.6,
     main = sprintf("EVT centroids in z_type -> %d clusters (Ward)", K))
rect.hclust(hc, k = K, border = 2:(K + 1))
par(op)
```

## EVT cohesion under location-based assignment

How much do EVTs stay together now that plots are assigned by embedding location?
(High = `z_type` respects EVT; an EVT splits only when its plots are
embedding-diverse.)

```{r verify}
tab <- table(evt = evt, cluster = emb$type_cluster)
cat(sprintf("EVT cohesion (plot-weighted: plots in their EVT's modal segment): %.3f\n",
            sum(apply(tab, 1, max)) / sum(tab)))
cat(sprintf("EVTs entirely within one segment: %d / %d (%.1f%%)\n",
            sum(rowSums(tab > 0) == 1), nrow(tab), 100 * mean(rowSums(tab > 0) == 1)))
```

## Cluster composition

Each cluster aggregates several EVTs (expected). `n_evts` = distinct EVTs pooled
into the cluster; `top_evts` = the largest few by plot count.

```{r composition}
emb %>%
  group_by(type_cluster) %>%
  summarize(n_plots = n(),
            n_evts  = dplyr::n_distinct(evt),
            top_evts = paste(names(sort(table(evt), decreasing = TRUE))[1:3], collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_plots))
```

## Does z_type respect EVT? (segmentation compactness)

Compare within-segment `z_type` dispersion of the EVT-trained segmentation to a
**free** k-means at the same `K` (which minimizes it). Ratio near 1 means the
EVT-trained segments are almost as tight as unconstrained clusters — i.e. `z_type`
already separates EVTs cleanly; a large ratio means the EVT supervision pulls the
segments away from the embedding's natural structure.

```{r compactness}
tot_within <- function(cl) {
  ss <- sum(vapply(split(seq_len(nrow(Z)), cl), function(ix) {
    m <- Z[ix, , drop = FALSE]; sum(sweep(m, 2, colMeans(m))^2)
  }, numeric(1)))
  ss / nrow(Z)
}
set.seed(1)
km <- kmeans(Z, centers = K, nstart = 10)

w_evt <- tot_within(emb$type_cluster)
w_km  <- tot_within(km$cluster)
data.frame(
  clustering = c("EVT-trained segmentation (this)", "free k-means (same K)"),
  within_var = round(c(w_evt, w_km), 3),
  ratio_to_kmeans = round(c(w_evt / w_km, 1), 3)
)

# homogeneity: how EVT-coherent the segments are (higher = each segment ~ few EVTs)
homogeneity <- 1 - sum(apply(tab, 2, function(col) {
  p <- col[col > 0] / sum(col); -sum(p * log(p))
}) * colSums(tab)) / sum(tab) /
  ( -sum((rowSums(tab)/sum(tab)) * log(rowSums(tab)/sum(tab))) )
cat(sprintf("homogeneity (segment -> one EVT): %.3f (higher = segments are EVT-coherent)\n",
            homogeneity))
```

## EVT -> segment map

Each EVT's **modal** segment and its **cohesion** (fraction of the EVT's plots in
that segment). Cohesion < 1 means the EVT split across segments under
location-based assignment.

```{r map}
# LANDFIRE EVT lookup: VALUE (the EVT code) -> EVT_NAME
evt_lookup <- read_csv(params$evt_table, show_col_types = FALSE) %>%
  select(VALUE, EVT_NAME)

evt_map <- emb %>%
  group_by(VALUE = evt) %>%
  summarize(n_plots  = n(),
            cluster  = as.integer(names(sort(table(type_cluster), decreasing = TRUE))[1]),
            cohesion = round(max(table(type_cluster)) / n(), 3),
            .groups  = "drop") %>%
  mutate(VALUE = as.integer(VALUE)) %>%
  left_join(evt_lookup, by = "VALUE") %>%
  select(VALUE, EVT_NAME, n_plots, type_cluster = cluster, cohesion) %>%
  arrange(type_cluster, desc(n_plots))
evt_map
```

## Stratification: BALIVE, Volume, and Removals vs TCC

Build the fine bins as **EVT type clusters, then phase-subdivided**: take the
`r params$K` EVT-trained type segments, and split the top
`r round(100*params$phase_subdivide_frac)`% (by within-cluster `z_phase` spread)
into `r params$k_phase` phase sub-groups each (Ward on `z_phase`) — the same
type->phase refinement as the geometry hierarchical script, but with EVT-based
type bins. Then aggregate the fine bins into `K_strata` strata by Ward on the
cluster-mean response (supervised), and compare within-stratum SD against TCC and
a flat type+phase probe, by survey unit. Responses: **BALIVE**, **standing live
volume**, and **removals**. Stocks (BALIVE/volume) run on all forested plots;
removals additionally require remeasurement (with forested zero-removal plots
kept as 0).

```{r strat-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)

cond <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, CONDPROP_UNADJ, BALIVE, FORTYPCD FROM COND WHERE STATECD='%s'",
  params$statecd)) %>% mutate(PLT_CN = as.character(PLT_CN))
balive <- cond %>%
  mutate(BALIVE = if_else(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")
fortyp <- cond %>%
  group_by(PLT_CN) %>% slice_max(CONDPROP_UNADJ, n = 1, with_ties = FALSE) %>%
  ungroup() %>% select(PLT_CN, FORTYPCD)

dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# standing LIVE net cu ft volume per acre (STATUSCD=1)
vol <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
     FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

# annual per-acre removed GS volume; COALESCE inside SUM so no-removal plots are 0
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN
    GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

fia_locs <- read_csv(params$fia_locs, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN))
```

```{r strat-tcc}
locs <- fia_locs %>% filter(CN %in% emb$CN) %>% distinct(CN, .keep_all = TRUE)
tccr <- rast(params$tcc)
pts  <- vect(locs, geom = c("LON_ACTUAL", "LAT_ACTUAL"), crs = "EPSG:4269")
tv   <- extract(tccr, project(pts, crs(tccr)), ID = FALSE)[[1]]
tv   <- as.numeric(as.character(tv)); tv[tv > 100] <- NA_real_
tcc_by_cn <- tibble(CN = locs$CN, tcc = tv)
```

```{r strat-assemble}
analysis <- emb %>%
  select(CN, type_cluster, all_of(emb_type_vars), all_of(emb_phase_vars)) %>%
  inner_join(balive, by = c("CN" = "PLT_CN")) %>%
  left_join(vol,     by = c("CN" = "PLT_CN")) %>%
  inner_join(fortyp, by = c("CN" = "PLT_CN")) %>%
  inner_join(select(fia_locs, CN, UNITCD), by = "CN") %>%
  inner_join(tcc_by_cn, by = "CN") %>%
  mutate(volume = coalesce(volume, 0)) %>%
  filter(!is.na(FORTYPCD), !is.na(UNITCD), !is.na(tcc), tcc >= 0, tcc <= 100)
cat(sprintf("forested plots with covariates: %d across %d units\n",
            nrow(analysis), dplyr::n_distinct(analysis$UNITCD)))
```

```{r strat-helpers}
pooled_within <- function(df, groups, y) {
  g <- df %>% group_by(across(all_of(groups))) %>%
    summarize(n = n(), ss = sum((.data[[y]] - mean(.data[[y]]))^2), .groups = "drop")
  sum(g$ss) / (sum(g$n) - nrow(g))
}
# Dalenius-Hodges cumulative-sqrt(f) boundaries (the K-1 interior breaks)
dh_breaks <- function(x, K, nbins = 100) {
  x <- as.numeric(x); rng <- range(x, finite = TRUE)
  if (!is.finite(diff(rng)) || diff(rng) == 0) return(numeric(0))
  h <- hist(x, breaks = seq(rng[1], rng[2], length.out = nbins + 1), plot = FALSE)
  csf <- cumsum(sqrt(h$counts)); tgt <- seq(0, max(csf), length.out = K + 1)[-c(1, K + 1)]
  sort(unique(vapply(tgt, function(t) h$breaks[which(csf >= t)[1] + 1L], numeric(1))))
}
dh_strata <- function(x, K, nbins = 100) {
  bnd <- dh_breaks(x, K, nbins)
  if (length(bnd) == 0) return(factor(rep(1L, length(x))))
  factor(findInterval(as.numeric(x), c(-Inf, bnd, Inf)))
}
ward <- function(X, k) {
  d  <- dist(X)
  hc <- if (requireNamespace("fastcluster", quietly = TRUE))
          fastcluster::hclust(d, method = "ward.D2") else hclust(d, method = "ward.D2")
  cutree(hc, k = k)
}
# fine bins -> strata: Ward on the fine-bin-mean response (supervised aggregation)
bin_map <- function(bin, y, K) {
  d  <- tibble(fine_bin = bin, .y = y) %>% group_by(fine_bin) %>%
        summarize(n_plots = n(), mean_y = mean(.y), .groups = "drop")
  Kk <- min(K, nrow(d))
  d$stratum <- cutree(hclust(dist(scale(d$mean_y)), method = "ward.D2"), Kk)
  d
}
agg_by <- function(bin, y, K) {
  d <- bin_map(bin, y, K)
  factor(d$stratum[match(bin, d$fine_bin)])
}
# fine-bin -> stratum table for display
show_map <- function(df, y) {
  bin_map(df$fine_bin, df[[y]], params$K_strata) %>%
    transmute(fine_bin, n_plots, mean_response = round(mean_y, 2), stratum) %>%
    arrange(stratum, desc(mean_response))
}
# flat OOF supervised probe on `vars` -> DH bins (the direct-supervision ceiling)
oof_dh <- function(df, y, vars, K) {
  set.seed(1); folds <- sample(rep(1:5, length.out = nrow(df)))
  d <- df %>% mutate(y_resp = .data[[y]])
  form <- reformulate(vars, "y_resp"); s <- numeric(nrow(df))
  for (f in 1:5) { fit <- lm(form, d[folds != f, ]); s[folds == f] <- predict(fit, d[folds == f, ]) }
  dh_strata(s, K)
}
```

### Fine bins: EVT type clusters -> phase subdivision

```{r fine-bins}
# within-cluster z_phase spread; subdivide the top fraction into k_phase groups
Zp   <- scale(as.matrix(analysis[, emb_phase_vars]))
pvar <- tapply(seq_len(nrow(analysis)), analysis$type_cluster,
               function(ix) sum(apply(Zp[ix, , drop = FALSE], 2, var)))
n_cl   <- length(pvar)
n_sub  <- max(1, round(params$phase_subdivide_frac * n_cl))
sub_cl <- names(sort(pvar, decreasing = TRUE))[seq_len(n_sub)]

analysis$phase_sub <- 1L
for (b in sub_cl) {
  ix <- which(as.character(analysis$type_cluster) == b)
  if (length(ix) >= 10 * params$k_phase)
    analysis$phase_sub[ix] <- ward(Zp[ix, , drop = FALSE], params$k_phase)
}
analysis$fine_bin <- interaction(analysis$type_cluster, analysis$phase_sub, drop = TRUE)
cat(sprintf("fine bins: %d (from %d type clusters, %d phase-subdivided)\n",
            nlevels(analysis$fine_bin), n_cl, length(sub_cl)))

# type clusters ranked by within-cluster z_phase variance, flagged if subdivided
tibble(type_cluster = names(pvar),
       n            = as.integer(table(analysis$type_cluster)[names(pvar)]),
       phase_var    = round(as.numeric(pvar), 3),
       subdivided   = names(pvar) %in% sub_cl) %>%
  arrange(desc(phase_var))
```

### Removals population (forested + remeasured)

```{r removals-pop}
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # remeasured; zeros kept
cat(sprintf("forested remeasured plots: %d (%.1f%% with any removal)\n",
            nrow(ar), 100 * mean(ar$remv_cfgs > 0)))
```

### Evaluation

```{r strat-eval}
# TCC break points actually used by the two TCC schemes
cat("FIA fixed TCC breaks:", "10, 48, 85", "(0-10 / 11-48 / 49-85 / 86-100)\n")
cat(sprintf("TCC DH (%d) breaks: %s\n", params$K_strata,
            paste(round(dh_breaks(analysis$tcc, params$K_strata), 1), collapse = ", ")))

resp_table <- function(df, y, vars = emb_vars) {
  s_fia  <- cut(df$tcc, breaks = c(-Inf, 10, 48, 85, Inf),
                labels = c("0-10", "11-48", "49-85", "86-100"))     # FIA operational
  s_tcc  <- dh_strata(df$tcc, params$K_strata)
  s_evt  <- agg_by(df$fine_bin, df[[y]], params$K_strata)
  s_flat <- oof_dh(df, y, vars, params$K_strata)
  vv <- c(pooled_within(df, "UNITCD", y),
          pooled_within(mutate(df, .s = s_fia),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_tcc),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_evt),  c("UNITCD", ".s"), y),
          pooled_within(mutate(df, .s = s_flat), c("UNITCD", ".s"), y))
  data.frame(
    stratification  = c("UNITCD only (baseline)",
                        "UNITCD x TCC FIA fixed (0-10/11-48/49-85/86-100)",
                        sprintf("UNITCD x TCC DH (%d)", params$K_strata),
                        sprintf("UNITCD x EVT type/phase hierarchical (%d)", params$K_strata),
                        sprintf("UNITCD x flat embedding supervised (%d)", params$K_strata)),
    within_var      = round(vv, 2),
    within_sd       = round(sqrt(vv), 3),
    pct_of_baseline = round(100 * vv / vv[1], 1))
}
```

#### BALIVE

```{r strat-balive}
resp_table(analysis, "BALIVE")
```

Fine bin -> stratum mapping (BALIVE):

```{r strat-balive-map}
show_map(analysis, "BALIVE")
```

#### Standing live volume

```{r strat-volume}
resp_table(analysis, "volume")
```

Fine bin -> stratum mapping (volume):

```{r strat-volume-map}
show_map(analysis, "volume")
```

#### Removals

```{r strat-removals}
resp_table(ar, "remv_cfgs")
```

Fine bin -> stratum mapping (removals):

```{r strat-removals-map}
show_map(ar, "remv_cfgs")
```

```{r strat-cleanup}
dbDisconnect(con)
```

## Notes

- **EVT-trained, location-assigned.** EVT groups define the segment centroids;
  each plot is then assigned by nearest centroid in `z_type`, so an EVT may split
  across segments. The `EVT cohesion` diagnostic reports how much they stay
  together (high if `z_type` respects EVT).
- **The compactness ratio is the headline diagnostic.** Near 1.0 = `z_type`
  already encodes EVT structure (the model learned vegetation type), so the
  EVT-trained segments are almost as tight as free clusters. Far above 1 = EVTs
  are smeared across the embedding and the supervision is expensive.
- **Rare EVTs** (`n < min_evt_n`) only seed the *grouping* via the nearest
  EVT-centroid cluster; their plots, like all plots, are then assigned by
  location. Raise `min_evt_n` to keep noisier EVT centroids out of the grouping.
- **Choosing K.** Read it off the dendrogram; `K` is a param. Unweighted centroids
  give rare and common EVTs equal say in the grouping -- weight `cent_big` rows by
  `sqrt(n)` if you want common EVTs to dominate the merges.
- **Using it downstream.** `emb$type_cluster` (or `evt_map`) is a drop-in
  replacement for the k-means/Ward `type_bin` step in the hierarchical
  stratification -- an EVT-supervised type binning that never sees the FIA target.
- **Fine bins = EVT type + phase.** The EVT clusters are phase-subdivided (top
  `phase_subdivide_frac` by `z_phase` spread, `k_phase` groups each via Ward) to
  match the geometry hierarchical script; set `phase_subdivide_frac = 0` to test
  EVT-type-only fine bins.
- **Responses.** BALIVE and standing live volume (`TREE.VOLCFNET`, `STATUSCD=1`)
  are stocks on all forested plots; removals (`TREE_GRM_COMPONENT`, annual per-acre
  removed GS volume) additionally require remeasurement, with forested no-removal
  plots kept as 0. The `EVT type/phase hierarchical` row aggregates the fine bins
  to `K_strata` by Ward on the fine-bin-mean response; `flat embedding supervised`
  is the direct OOF (type+phase) probe ceiling for the same information.
