---
title: "kNN embedding models for FIA attributes"
subtitle: "TPA, hardwood/softwood TPA, basal area, live volume, removals"
author: "FoR-EST project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  emb_csv: "~/Dropbox/git/vq-vae/frl/runs/frl_v0_exp033/va_merge_embeddings.csv"
  fiadb: "/data/FIADB/SQLite_FIADB_ENTIRE.db"
  statecd: "51"
  split_col: "split"          # train/val/test column in the embeddings CSV
  k_grid: "5,10,20,40"        # neighbour counts tuned on the validation split
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 110)
```

## Overview

kNN regression from the **entire embedding** (`z_type` + `z_phase`, all
dimensions) to six FIA attributes:

| target | source |
|--------|--------|
| **TPA** (all live) | `TREE.TPA_UNADJ`, `STATUSCD=1` |
| **TPA hardwood** | as above, `SPCD >= 300` |
| **TPA softwood** | as above, `SPCD < 300` |
| **basal area** | `COND.BALIVE` |
| **standing live volume** | `TREE.VOLCFNET * TPA_UNADJ`, `STATUSCD=1` |
| **removals** | `TREE_GRM_COMPONENT` annual removed GS volume |

Uses the existing **train / val / test** split: the model (train neighbours) is
fit on train, `k` is tuned on val, and test is scored once. The embedding is
standardized with **train-only** statistics (no leakage). Distances are Euclidean
over the full embedding — note kNN in this many dimensions is subject to distance
concentration, so treat these as a baseline "is the signal linearly *and* locally
present" check against the linear probes elsewhere.

Populations: TPA / basal area / volume are over the full plot population
(non-forest plots contribute 0); removals additionally require GRM/remeasured
data.

```{r libraries}
library(tidyverse)
library(DBI)
library(RSQLite)
library(FNN)
```

## Load embeddings and split

```{r load}
emb_type_vars  <- paste0("z_type_",  0:47)
emb_phase_vars <- paste0("z_phase_", 0:7)
emb_vars       <- c(emb_type_vars, emb_phase_vars)

emb <- read_csv(params$emb_csv, col_types = cols(CN = "c", ID = "c")) %>%
  mutate(CN = sub("\\.0$", "", CN), ID = sub("\\.0$", "", ID)) %>%
  filter(if_all(all_of(emb_vars), ~ !is.na(.)))

# train/val/test come from the `split` column in the embeddings CSV, the same
# source fia_embedding_evaluation.Rmd uses (reg_data$split)
if (!params$split_col %in% names(emb))
  stop(sprintf("split column '%s' not in embeddings CSV; columns include: %s",
               params$split_col, paste(head(names(emb), 15), collapse = ", ")))
split_df <- emb %>% transmute(CN, split = as.character(.data[[params$split_col]])) %>%
  filter(split %in% c("train", "val", "test")) %>%
  distinct(CN, .keep_all = TRUE)
cat("split counts:\n"); print(table(split_df$split))
```

## FIA targets

```{r fia-data}
con <- dbConnect(RSQLite::SQLite(), params$fiadb)
dbWriteTable(con, "tmp_cn", tibble(PLT_CN = unique(emb$CN)), temporary = TRUE, overwrite = TRUE)

# basal area (live) per acre, from COND
balive <- dbGetQuery(con, sprintf(
  "SELECT PLT_CN, CONDPROP_UNADJ, BALIVE FROM COND WHERE STATECD='%s'", params$statecd)) %>%
  mutate(PLT_CN = as.character(PLT_CN), BALIVE = ifelse(is.na(BALIVE), 0, BALIVE)) %>%
  group_by(PLT_CN) %>%
  summarize(BALIVE = sum(BALIVE * CONDPROP_UNADJ) / sum(CONDPROP_UNADJ), .groups = "drop")

# TPA (total / hardwood SPCD>=300 / softwood SPCD<300) and live volume, from TREE
tree_stats <- dbGetQuery(con,
  "SELECT g.PLT_CN,
     SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_total,
     SUM(CASE WHEN g.STATUSCD=1 AND g.SPCD>=300 THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_hard,
     SUM(CASE WHEN g.STATUSCD=1 AND g.SPCD<300  THEN COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS tpa_soft,
     SUM(CASE WHEN g.STATUSCD=1 THEN COALESCE(g.VOLCFNET,0)*COALESCE(g.TPA_UNADJ,0) ELSE 0 END) AS volume
   FROM TREE g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))

# removals: annual per-acre removed GS volume (GRM; 0 for no-removal plots)
remv <- dbGetQuery(con,
  "SELECT g.PLT_CN,
          SUM(COALESCE(g.REMVCFGS_FOREST,0) * COALESCE(g.SUBP_TPAREMV_UNADJ_GS_FOREST,0)) AS remv_cfgs
     FROM TREE_GRM_COMPONENT g JOIN tmp_cn t ON g.PLT_CN = t.PLT_CN GROUP BY g.PLT_CN") %>%
  mutate(PLT_CN = as.character(PLT_CN))
dbDisconnect(con)
```

```{r assemble}
analysis <- emb %>% select(CN, all_of(emb_vars)) %>%
  inner_join(split_df, by = "CN") %>%
  left_join(tree_stats, by = c("CN" = "PLT_CN")) %>%
  left_join(balive,     by = c("CN" = "PLT_CN")) %>%
  mutate(across(c(tpa_total, tpa_hard, tpa_soft, volume, BALIVE), ~ coalesce(., 0)))
ar <- analysis %>% inner_join(remv, by = c("CN" = "PLT_CN"))   # removals: GRM/remeasured
cat(sprintf("plots (stocks): %d   plots (removals): %d\n", nrow(analysis), nrow(ar)))

# train-only standardization of the embedding (fit scaler on train rows)
tr_all <- analysis$split == "train"
mu     <- colMeans(as.matrix(analysis[tr_all, emb_vars]))
sdv    <- apply(as.matrix(analysis[tr_all, emb_vars]), 2, sd); sdv[sdv == 0] <- 1
```

## kNN evaluation

```{r knn-helpers}
kgrid <- as.integer(strsplit(params$k_grid, ",")[[1]])
r2   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)

run_knn <- function(df, yname, label) {
  X <- scale(as.matrix(df[, emb_vars]), center = mu, scale = sdv)
  y <- df[[yname]]
  tr <- which(df$split == "train"); va <- which(df$split == "val"); te <- which(df$split == "test")
  # tune k on validation
  val_r2 <- sapply(kgrid, function(k)
    r2(y[va], FNN::knn.reg(X[tr, , drop = FALSE], X[va, , drop = FALSE], y[tr], k = k)$pred))
  kbest <- kgrid[which.max(val_r2)]
  pte <- FNN::knn.reg(X[tr, , drop = FALSE], X[te, , drop = FALSE], y[tr], k = kbest)$pred
  list(
    metric = data.frame(target = label, k_best = kbest,
                        n_train = length(tr), n_val = length(va), n_test = length(te),
                        val_R2 = round(max(val_r2), 3), test_R2 = round(r2(y[te], pte), 3),
                        test_RMSE = round(sqrt(mean((y[te] - pte)^2)), 2),
                        test_MAE = round(mean(abs(y[te] - pte)), 2)),
    pred = tibble(target = label, obs = y[te], pred = pte))
}

targets <- list(
  list(analysis, "tpa_total", "TPA (all live)"),
  list(analysis, "tpa_hard",  "TPA hardwood"),
  list(analysis, "tpa_soft",  "TPA softwood"),
  list(analysis, "BALIVE",    "basal area"),
  list(analysis, "volume",    "standing live volume"),
  list(ar,       "remv_cfgs", "removals"))
res <- lapply(targets, function(t) run_knn(t[[1]], t[[2]], t[[3]]))
```

### Test-set performance

`k_best` is chosen on validation; `test_R2/RMSE/MAE` are the held-out scores.

```{r knn-metrics}
bind_rows(lapply(res, `[[`, "metric"))
```

### Predicted vs observed (test)

```{r knn-scatter, fig.width=11, fig.height=7}
preds <- bind_rows(lapply(res, `[[`, "pred")) %>%
  mutate(target = factor(target, levels = sapply(targets, `[[`, 3)))
ggplot(preds, aes(obs, pred)) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.4) +
  geom_point(alpha = 0.15, size = 0.5) +
  facet_wrap(~ target, scales = "free", ncol = 3) +
  labs(x = "observed", y = "kNN predicted") +
  theme_bw(base_size = 10)
```

## Notes

- **Split source.** The `train`/`val`/`test` label is the `split` column in the
  embeddings CSV — the same source `fia_embedding_evaluation.Rmd` uses
  (`reg_data$split`). Override with `split_col` if needed; the chunk errors and
  lists available columns if the column is missing.
- **No leakage.** The embedding is standardized with train-only mean/sd, `k` is
  tuned only on validation, and test is scored once with the selected `k`.
- **Hardwood/softwood split** uses the FIA convention `SPCD < 300` = softwood
  (conifer), `SPCD >= 300` = hardwood.
- **Populations.** TPA/basal area/volume use all plots (non-forest = 0); removals
  use the GRM/remeasured subset (`ar`), keeping forested zero-removal plots as 0.
- **Dimensionality caveat.** kNN uses Euclidean distance over the full
  `r length(emb_vars)`-dim embedding; distance concentration can blunt it. Compare
  `test_R2` here against the linear-probe R² in the other scripts — a large gap in
  either direction is informative about whether the signal is linear vs local.
- **Tuning.** `k` is tuned over `k_grid`; widen it (e.g. add 80, 160) if the best
  `k` sits at the edge of the grid.
